---
title: "Replicating Jones Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, tidy = TRUE)
```

1. Transects were filtered to only include broadcast burning, wildfire, or untreated points.

Note: Incomplete transects (those with thinning-only points filtered out) were allowed in the dataset. EVT groups without broadcast/wildfire treatments were removed from controls. The EVT group "other" was also removed for consistent context.

2. Six models were constructed for different combinations of starting severity and transect type:

Low x Entering, Moderate x Entering, High x Entering
Low x Exiting, Moderate x Exiting, High x Exiting

3. The model formula is:

rdnbr ~ type + treated + (1|Incid_Name/Transect_ID)

Where type = control or entering/exiting & treated = untreated or broadcast/wildfire


```{r, include=FALSE}
library(tidyverse)
library(sf)
library(lme4)
library(car)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)
library(here)
library(performance)
library(sjPlot) #Table of model outputs

transects <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "transectsDS") 

transects <- filter(transects,!Incid_Name %in% c("TRAIL SPRINGS", "JUNKINS"))
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Rearrangement", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))
transects$Event_ID <- factor(transects$Event_ID)
transects$Incid_Name <- factor(transects$Incid_Name)
transects$TransectID <- factor(transects$Transect_ID)

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)

#Sort start of transects into L/M/H based on average severity of first 5 points
transects <- transects %>%
  group_by(Transect_ID) %>% 
mutate(
    start_value = round(mean(rdnbr[Point_ID %in% 1:5], na.rm = TRUE)),
    START_SEV = case_when(
      start_value <= 338              ~ "Low",
      start_value >= 339 & start_value <= 726 ~ "Moderate",
      start_value >= 727              ~ "High",
    )
  ) %>%
  ungroup() %>%
  select(-start_value)

#Drop geometry
transects <- st_drop_geometry(transects)
```

# Dataset Summary

```{r, results = 'asis'}
jones_df <- transects %>%
  filter(!TRT_TYPE %in% c("Rearrangement", "Removal")) %>%
  filter(!EVT_Group %in% c("other", "pinyon-juniper", "spruce-fir")) #removing due to lack of treatments in pj or spruce fir, removing other for lack of context
```

## Transects by Type in each Fire and Starting Severity

```{r, results = 'asis'}
# Build abbreviation lookup once, outside the pipe
fire_abbrevs <- jones_df %>% 
  st_drop_geometry() %>% 
  distinct(Incid_Name) %>% 
  arrange(Incid_Name) %>% 
  mutate(abbrev = abbreviate(Incid_Name, minlength = 3)) %>% 
  tibble::deframe()  

jones_df %>% 
  st_drop_geometry() %>% 
  count(START_SEV, Incid_Name, type) %>% 
  dplyr::mutate(
    n = n / 10,
    START_SEV = factor(START_SEV, levels = c("Low", "Moderate", "High")),
    type = factor(type, levels = c("control", "entering", "exiting")),
    Incid_Name = fire_abbrevs[Incid_Name]  # swap to abbreviation before pivoting
  ) %>% 
  tidyr::pivot_wider(
    names_from = Incid_Name,
    values_from = n,
    values_fill = 0
  ) %>% 
  dplyr::arrange(START_SEV, type) %>% 
  dplyr::mutate(
    Total = rowSums(dplyr::across(where(is.numeric)))
  ) %>% 
  dplyr::group_split(START_SEV) %>% 
  purrr::iwalk(~ {
    cat("\n\n###", as.character(unique(.x$START_SEV)), "\n\n")
    print(
      knitr::kable(
        dplyr::select(.x, -START_SEV),
        format = "pipe"
      )
    )
  })


# Print the key once, after all three tables
cat("\n\n**Key:**", paste(names(fire_abbrevs), fire_abbrevs, sep = " = ", collapse = "; "), "\n\n")
```

## Points by Treatment Type in each EVT Group and Starting Severity

```{r, results = 'asis'}
#Table of Transects by Type in each EVT Group
jones_df %>% 
  st_drop_geometry() %>% 
  count(START_SEV, EVT_Group, treated) %>% 
  dplyr::mutate(
    n = n,
    START_SEV = factor(START_SEV, levels = c("Low", "Moderate", "High"))
  ) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  ) %>% 
  dplyr::arrange(START_SEV, treated) %>% 
  dplyr::group_split(START_SEV) %>% 
  purrr::iwalk(~ {
    cat("\n\n###", as.character(unique(.x$START_SEV)), "\n\n")
    print(
      knitr::kable(
        dplyr::select(.x, -START_SEV),
        format = "pipe"
      )
    )
  })
```


```{r}
# Filter by TRT_TYPE to only include broadcast or wildfire
black_df <- jones_df %>%
  filter(type != "exiting")

black_df$TRT_TYPE <- factor(black_df$TRT_TYPE, levels = c("Untreated", "Broadcast Burn", "Wildfire"))
black_df$type <- factor(black_df$type, levels = c("control", "entering"))

black_low_df <- black_df %>%
  filter(START_SEV == "Low") #593 obs
black_mod_df <- black_df %>%
  filter(START_SEV == "Moderate") #352 obs
black_high_df <- black_df %>%
  filter(START_SEV == "High") #1385 obs
```


```{r}
# Filter by TRT_TYPE to only include broadcast or wildfire
black2_df <- jones_df %>%
  filter(type != "entering")

black2_df$TRT_TYPE <- factor(black2_df$TRT_TYPE, levels = c("Untreated", "Broadcast Burn", "Wildfire"))
black2_df$type <- factor(black2_df$type, levels = c("control", "exiting"))

black2_low_df <- black2_df %>%
  filter(START_SEV == "Low") #593 obs
black2_mod_df <- black2_df %>%
  filter(START_SEV == "Moderate") #352 obs
black2_high_df <- black2_df %>%
  filter(START_SEV == "High") #1385 obs
```

# Entering Models

### Low Entering Model

```{r}
black_low_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black_low_df, REML = TRUE)

tab_model(black_low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_low_fit), resid(black_low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Moderate Entering Model

```{r}
black_mod_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black_mod_df, REML = TRUE)

tab_model(black_mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_mod_fit), resid(black_mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### High Entering Model

```{r}
black_high_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black_high_df, REML = TRUE)

tab_model(black_high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_high_fit), resid(black_high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Entering Effect

```{r}
# Extract population-level predictions (averaged over random effects)
pred_low  <- ggpredict(black_low_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Low") %>%
  filter(!is.na(predicted))

pred_mod  <- ggpredict(black_mod_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Moderate") %>%
  filter(!is.na(predicted))

pred_high <- ggpredict(black_high_fit, terms = c("type", "treated")) %>%
  mutate(severity = "High") %>%
  filter(!is.na(predicted))

# Combine, drop the non-existent control+treated combination,
# and relabel x-axis categories
pred_all <- bind_rows(pred_low, pred_mod, pred_high) %>%
  filter(!(x == "control" & group == "Yes"))

pred_all$severity <- factor(pred_all$severity, levels = c("High", "Moderate", "Low"))

# Plot
ggplot(pred_all, aes(x = x, y = predicted, group = group, shape = group)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.15, position = position_dodge(width = 0.3)) +
  facet_wrap(~severity, nrow = 1) +
  labs(
    x = NULL,
    y = "Predicted RdNBR"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```

# Exiting Models

### Low Exiting Model

```{r}
black2_low_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black2_low_df, REML = TRUE)

tab_model(black2_low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_low_fit), resid(black2_low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 


```

### Moderate Exiting Model

```{r}
black2_mod_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black2_mod_df, REML = TRUE)

tab_model(black2_mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_mod_fit), resid(black2_mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### High Exiting Model

```{r}
black2_high_fit <- lmer(rdnbr ~ type + treated + (1|Incid_Name/Transect_ID), data = black2_high_df, REML = TRUE)

tab_model(black2_high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_high_fit), resid(black2_high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Exiting Effect

```{r}
# Extract population-level predictions (averaged over random effects)
pred_low2  <- ggpredict(black2_low_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Low") %>%
  filter(!is.na(predicted))

pred_mod2  <- ggpredict(black2_mod_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Moderate") %>%
  filter(!is.na(predicted))

pred_high2 <- ggpredict(black2_high_fit, terms = c("type", "treated")) %>%
  mutate(severity = "High") %>%
  filter(!is.na(predicted))

# Combine, drop the non-existent control+treated combination,
# and relabel x-axis categories
pred_all2 <- bind_rows(pred_low2, pred_mod2, pred_high2) %>%
  filter(!(x == "control" & group == "Yes"))

pred_all2$severity <- factor(pred_all2$severity, levels = c("High", "Moderate", "Low"))

pred_all2$group <- factor(pred_all2$group, levels = c("Yes", "No"))

# Plot
ggplot(pred_all2, aes(x = x, y = predicted, group = group, shape = group)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.15, position = position_dodge(width = 0.3)) +
  facet_wrap(~severity, nrow = 1) +
  labs(
    x = NULL,
    y = "Predicted RdNBR"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )
```