---
title: "Weekly Check In 8.17.26"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, tidy = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(car)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)
library(here)

transects <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "transectsDS") 

transects <- filter(transects,!Incid_Name %in% c("TRAIL SPRINGS", "JUNKINS"))
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Rearrangement", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)

#Sort start of transects into L/M/H based on average severity of first 5 points
transects <- transects %>%
  group_by(Transect_ID) %>% 
mutate(
    start_value = mean(rdnbr[Point_ID %in% 1:5], na.rm = TRUE),
    START_SEV = case_when(
      start_value < 300              ~ "Low",
      start_value >= 300 & start_value <= 500 ~ "Moderate",
      start_value > 500              ~ "High",
    )
  ) %>%
  ungroup() %>%
  select(-start_value)
```

## Response Variable Distribution

```{r}
hist(transects$cbi, breaks = 30)
hist(transects$rdnbr, breaks = 30)

ggplot(transects, aes(x = rdnbr, groups = Incid_Name)) +
  facet_wrap(~Incid_Name) +
  geom_histogram()

ggplot(transects, aes(x = cbi, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()

# View distribution of response faceted by starting severity
ggplot(transects, aes(x = rdnbr, fill = START_SEV)) +
  geom_density(alpha = 0.5) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.5) + #add after_stat to scale each group based on sample size (area under curve = 1 across all groups, not for each individual group)
  theme_bw()

#Faceted by EVT for easier viewing
ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  facet_wrap(~EVT_Group, ncol = 2) +
  theme_bw() +
  theme(legend.position = "none")

ggplot(transects, aes(x = rdnbr, fill = TRT_TYPE)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = type)) +
  geom_density(aes(y = after_stat(count/sum(count))), alpha = 0.5) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = Incid_Name)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  facet_wrap(~Incid_Name, ncol = 4) +
  theme_bw() +
  theme(legend.position = "none")

ggplot(transects, aes(x = rdnbr, fill = extreme_area)) +
  geom_density(alpha = 0.7) +
  theme_bw() #not scaled for sample size, getting error
```

# L/M/H Models

```{r}
transects %>% 
  st_drop_geometry() %>% 
  group_by(Incid_Name) %>%
  count(START_SEV) %>% 
  tidyr::pivot_wider(
    names_from = Incid_Name,
    values_from = n,
    values_fill = 0
  ) %>%
  print(n = Inf, width = Inf)

transects %>% 
  st_drop_geometry() %>% 
  group_by(EVT_Group) %>%
  count(TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  ) %>%
  print(n = Inf, width = Inf)

low_df <- transects %>%
  filter(START_SEV == "Low") #940 obs
mod_df <- transects %>%
  filter(START_SEV == "Moderate") #550 obs
high_df <- transects %>%
  filter(START_SEV == "High") #1860 obs

summary(low_df$type)
summary(mod_df$type)
summary(high_df$type)

low_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
mod_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
high_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
```

# Low Starting Severity Model

```{r}
low_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = low_df, REML = TRUE)

summary(low_fit)
vif(low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_fit), resid(low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Moderate Starting Severity Model

```{r}
mod_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mod_df, REML = TRUE)

summary(mod_fit)
vif(mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mod_fit), resid(mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# High Starting Severity Model

```{r}
high_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = high_df, REML = TRUE)

summary(high_fit)
vif(high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(high_fit), resid(high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + L/M/H Models

```{r}
low_enter_df <- low_df %>%
  filter(type == "entering") #940 obs -> 410 obs
mod_enter_df <- mod_df %>%
  filter(type == "entering") #550 obs -> 240 obs
high_enter_df <- high_df %>%
  filter(type == "entering") #1860 obs -> 780 obs
```

# Entering + Low Starting Severity Model

```{r}
low_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = low_enter_df, REML = TRUE)

summary(low_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_ent_fit), resid(low_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + Moderate Starting Severity Model

```{r}
mod_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = mod_enter_df, REML = TRUE)

summary(mod_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mod_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mod_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mod_ent_fit), resid(mod_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + High Starting Severity Model

```{r}
high_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = high_enter_df, REML = TRUE)

summary(high_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(high_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(high_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(high_ent_fit), resid(high_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Forest Type Models

```{r}
pondo_df <- transects %>%
  filter(EVT_Group == "ponderosa") %>% #857 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

mixed_df <- transects %>%
  filter(EVT_Group == "mixed conifer") %>% #775 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

lodge_df <- transects %>%
  filter(EVT_Group == "lodgepole") %>% #749 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #200 observations
```

# Ponderosa Model

```{r}
pondo_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = pondo_df, REML = TRUE)

summary(pondo_fit)
vif(pondo_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(pondo_fit), main = "Q-Q Plot of Residuals")
qqline(resid(pondo_fit), col = "red")

# Residuals vs Fitted
plot(fitted(pondo_fit), resid(pondo_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Mixed Conifer Model

```{r}
mixed_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mixed_df, REML = TRUE)

summary(mixed_fit)
vif(mixed_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mixed_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mixed_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mixed_fit), resid(mixed_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Lodgepole Model

```{r}
lodge_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = lodge_df, REML = TRUE) #Remove Event_ID in random effects because the only fire is Cameron Peak

summary(lodge_fit)
vif(lodge_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(lodge_fit), main = "Q-Q Plot of Residuals")
qqline(resid(lodge_fit), col = "red")

# Residuals vs Fitted
plot(fitted(lodge_fit), resid(lodge_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Dispersion L/M/H

Uneven sample size and heterogeneity within fires means that the variance within fires varies widely

```{r}
table(transects$Incid_Name)

transects %>%
  st_drop_geometry %>%
  group_by(Incid_Name) %>%
  summarise(n = n(), var_rdnbr = var(rdnbr), sd_rdnbr = sd(rdnbr)) %>%
  arrange(desc(var_rdnbr))
```

# Low Starting Severity + Dispersion Model

```{r}
low_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = low_df, dispformula = ~Event_ID, REML = TRUE)

summary(low_disp_fit)

sim_low_disp <- simulateResiduals(low_disp_fit)
plot(sim_low_disp)
```

# Moderate Starting Severity + Dispersion Model

Warning message:
In finalizeTMB(TMBStruc, obj, fit, h, data.tmb.old) :
  Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting')

```{r}
mod_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mod_df, dispformula = ~ Event_ID, REML = TRUE)

summary(mod_disp_fit)

sim_mod_disp <- simulateResiduals(mod_disp_fit)
plot(sim_mod_disp)
```

# High Starting Severity + Dispersion Model

Warning message:
In finalizeTMB(TMBStruc, obj, fit, h, data.tmb.old) :
  Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting')

```{r}
high_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = high_df, dispformula = ~ Event_ID, REML = TRUE)

summary(high_disp_fit)

sim_high_disp <- simulateResiduals(high_disp_fit)
plot(sim_high_disp)
```