---
title: "Intersect Treatments and Subsequent Wildfire"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 2
  pdf_document:
    toc: true
    toc_depth: '2'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(root.dir = 'C:/Users/Elizabeth Buhr/Documents/fuels-analysis/')
```

```{r}
library(sf)
library(tidyverse)
```

# 1. Data Organization

```{r}
target_crs <- "EPSG:26913"
```

## Load Forest Tracker 

This file is missing an assigned CRS and includes complex geometry types which are difficult to intersect, so this section also addresses those items. 

```{r}
#read file 
treats <- st_read("data/raw/CO_Forest_Tracker_2024/CO_Forest_Tracker_2024.gpkg")

#Missing crs - found from information tab in qgis (EPSG:26913 - NAD83 / UTM zone 13N)
st_crs(treats) <- 26913
treats <- st_transform(treats, target_crs) #reproject

#Simplify and clean
treats <- treats %>%
  st_zm(drop=TRUE, what = "ZM") %>% #drop third and forth dimension
  st_cast("MULTIPOLYGON") %>% #convert to simpler geometry 
  filter(MGT_TYPE != "REFOREST") %>%
  st_make_valid(treats)
```

## Load MTBS Fire Perimeters

I am filtering for fires from 2015 on (trying 2010 11/18/25). Perimeter data was pre-processed by HART lab.

```{r}
# Using perimeter data for 1984-2024
fires <- st_read("data/raw/mtbs_perimeter_data/mtbs_perims_DD.shp") %>% 
  st_transform(target_crs) %>% 
  mutate(Ig_Date = as.Date(Ig_Date), year=year(Ig_Date)) %>%   # convert to date
  filter(year>=2010) %>%  # filter for fires that occurred in or after 2015
  filter(Incid_Type=="Wildfire") %>%
  filter(grepl("CO", Event_ID)) %>% #filter out repeated fire names in other states, maybe should use state boundary instead
  st_make_valid(fires)
```

# 2. Find Fires of Interest 

We will find which wildfires intersect with previous treatments, then filter both the fires and treats datasets to only the relevant items. We will then save the filtered datasets for mapping.

## Intersect treatments and subsequent wildfires
```{r}
# Intersect treatments and subsequent wildfires
intersect <- st_intersection(treats, fires) %>%
  filter(YEAR_COMP < year) %>% #treat year before fire year
  dplyr::select("Event_ID", "Incid_Name", "year", "OBJECTID", "MGT_TYPE", "RXFIRE_MGT", "CANOPY_MGT", "SURF_MGT", "YEAR_COMP", "ACRES_MGT", "ACRES_GIS", "FOR_TYPE")
```

## Subset fires to only those that intersect 
```{r}
# Subset Fires to only those that intersect with prior fuel treatments
fires <- fires %>% 
  filter(Event_ID %in% unique(intersect%>% pull(Event_ID))) %>%
  dplyr::select("Event_ID", "Incid_Name", "Ig_Date", "year")
```

## Subset treatments to only those that intersect

This dataset has more observations than "intersect" because "OBJECTID" is not unique (multiple polygons can have the same OBJECTID because they are part of the same project). I think that's alright, we can see the whole project, but when doing treated area calculations later on we will want to be sure it is only the intersecting area.

```{r}
# Subset treatments to only those that intersect subsequent wildfire 
treats <- treats %>% 
  filter(OBJECTID %in% unique(intersect%>% pull(OBJECTID))) %>%
  dplyr::select("OBJECTID", "MGT_TYPE", "RXFIRE_MGT", "CANOPY_MGT", "SURF_MGT", "YEAR_COMP", "ACRES_MGT", "ACRES_GIS", "FOR_TYPE")
```


# 3. Save Geopackage

```{r}
# Save the intersection result as the first layer

st_write(
  obj = intersect,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "intersection",
  driver = "GPKG",
  delete_dsn = TRUE
)

# Append the subsetted fires as a new layer
st_write(
  obj = fires,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "fires_with_prior_treatment",
  driver = "GPKG",
  delete_layer= TRUE
)

# Append the subsetted treatments as a new layer
st_write(
  obj = treats,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "treatments_preceding_fire",
  driver = "GPKG",
  delete_layer= TRUE
)
```
