---
title: "1d--Fires-Selection"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(root.dir = 'C:/Users/Elizabeth Buhr/Documents/fuels-analysis/')
```

```{r}
#load packages
library(sf) 
library(tidyverse)
```

#Load list of fires with relevant info

```{r}
#read csv
fires <- read.csv("data/processed/my_fires.csv")
colnames(fires)
```

# Filter fires for proportion treated and minimum elevation

So far, the only filters for site/fire selection have been selecting for fires withing the past ~15 years (2010-2024) which intersect with previous fuels reduction. Now, I will impose 2 more filters:
--The proportion of the wildfire treated must be equal to or greater than 0.1, when rounded to two decimal points. This is to improve site sampling within each fire.
--The minimum elevation of the fire must be greater than 1500 m. This is to exclude lower-elevation vegetation types.

```{r}
myfires <- fires %>%
  filter(Proportion_Treated >= 0.1) %>%
  filter(Proportion_Treated < 1) %>%
  filter(Fire_Min_Elev > 1500)

write.csv(myfires, "data/processed/my_fires.csv")

myfiresIDs <- myfires$Event_ID

```

# Update all geopackage layers to reflect fire selection

```{r}
#path to file 
gpkg_path <- "data/processed/fuels_analysis.gpkg"

#check layer name
st_layers(gpkg_path)

#read the layers
fires <- st_read(gpkg_path, layer = "fires")
intersection <- st_read(gpkg_path, layer = "intersection")
treatments <- st_read(gpkg_path, layer = "treatments")

#subset fires
fires_sub <- fires %>% 
  filter(Event_ID %in% myfiresIDs)

#subset the intersection
inter_sub <- intersection %>% 
  filter(Event_ID %in% myfiresIDs)

#subset the treatments
treat_sub <- treatments %>%
  filter(OBJECTID %in% inter_sub$OBJECTID)

```

# Save new geopackage

```{r}
# copy over my fires layer
st_write(
  obj = fires_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "fires",
  driver = "GPKG",
  delete_layer = TRUE
)


# write new interactions layer that includes wildfire
st_write(
  obj = inter_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "intersection",
  driver = "GPKG",
  delete_layer = TRUE
)

# write new treatments layer that includes wildfire
st_write(
  obj = treat_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "treatments",
  driver = "GPKG",
  delete_layer= TRUE
)
```

# Prepare fires for use in Parks GEE methods

```{r}
## Prepares the selected fires perimeter dataset for use in Sean Parks' GEE code for calculating fire severity
st_read("data/processed/fuels_analysis.gpkg", layer = "fires") %>% # import selected perimeters
  mutate(date = as.Date(Ig_Date), "%Y%m%d") %>% # convert ignition date to date object
  mutate(Start_Day=min(lubridate::yday(date))-14) %>% # set first possible date 
  mutate(End_Day=max(lubridate::yday(date))+14) %>% # set last possible date 
  mutate(Fire_Year=as.character(year), Fire_ID=Event_ID) %>% # create variables with column names used in Parks code
  dplyr::select(Fire_ID, Fire_Year, Start_Day, End_Day) %>% # include only Fire ID, Fire Year, Start Day and End Day columns in dataset
  st_write("data/processed/GEE/fires_GEE_0713.shp", delete_dsn = TRUE)


#Do it for treatment fires too
## Prepares the selected fires perimeter dataset for use in Sean Parks' GEE code for calculating fire severity
st_read("data/processed/GEE/treatment_wildfires.shp") %>% # import selected perimeters
  mutate(date = as.Date(Ig_Date), "%Y%m%d") %>% # convert ignition date to date object
  mutate(Start_Day=min(lubridate::yday(date))-14) %>% # set first possible date 
  mutate(End_Day=max(lubridate::yday(date))+14) %>% # set last possible date 
  mutate(Fire_Year= as.character(YEAR_COMP), Fire_ID=Event_ID) %>% # create variables with column names used in Parks code
  dplyr::select(Fire_ID, Fire_Year, Start_Day, End_Day) %>% # include only Fire ID, Fire Year, Start Day and End Day columns in dataset
  st_write("data/processed/GEE/fires_GEE_Treatments_0713.shp", delete_dsn = TRUE)
```

# Make a map of fires across Colorado

```{r}
library(sf)
library(tidyverse)
library(terra)
library(shadowtext)
library(tidyterra)

fires_sub <- fires_sub %>%
  filter(Incid_Name != "ALKALI")

# Create the numbering lookup table
fire_numbers <- tibble(
  Incid_Name = c("EIGHT FOUR TWO", "TRAIL SPRINGS", "IRON", "CALWOOD",
                 "EAST FORK", "BEAVER", "LITTLE MESA", "DOE CANYON",
                 "PLATEAU", "CAMERON PEAK", "LAKE CHRISTINE",
                 "WESTON PASS", "DUCKETT", "JUNKINS", "LITTLE SAND"),
  fire_num = 1:15
)

# Join numbers to fires_sub
fires_sub <- fires_sub %>%
  left_join(fire_numbers, by = "Incid_Name")

fires_sub <- fires_sub %>%
  mutate(nudge_y_val = ifelse(fire_num == 5, 6500, 0))  

## Cities
cities <- st_read("D:/Spatial_Data/Cities/Colorado_City_Point_Locations.shp") %>%
  st_transform(crs = crs(fires_sub)) %>%
  filter(name %in% c("Denver"))
## State Boundary 
state <- st_read("D:/Spatial_Data/States/cb_2018_us_state_20m.shp") %>%
  st_transform(crs = crs(fires_sub)) %>%
  filter(NAME == "Colorado")

#Roads
roads <- st_read("D:/Spatial_Data/Highways/Highways%3A_Functional_Class.shp") %>%
  st_transform(crs=crs(fires_sub)) %>%
  filter(FUNCCLASS == "1  Interstate")

elev <- rast("D:/Spatial_Data/LANDFIRE/LF2020_Elev_CONUS/Tif/LF2020_Elev_CONUS.tif")

# Reproject state to match the raster's CRS first, then crop
state_elev_crs <- st_transform(state, crs(elev))

elev <- elev %>%
  crop(vect(state_elev_crs)) %>%        # crop in native CRS first (faster)
  mask(vect(state_elev_crs)) %>%        # mask in native CRS
  aggregate(fact = 3, fun = "mean") %>% # reduce resolution before reprojecting
  project(crs(fires_sub))              # reproject last

# Convert to data frame
elev_df <- as.data.frame(elev, xy = TRUE) %>%
  rename(elevation = 3)# Elevation raster
elev <- rast("D:/Spatial_Data/LANDFIRE/LF2020_Elev_CONUS/Tif/Lf2020_Elev_CONUS.tif") %>%
  project(crs(fires_sub)) %>%
  aggregate(elev, fact = 3, fun = "mean") %>%
  crop(vect(state)) %>%
  mask(vect(state))

# Make a map!
map <- ggplot() +
  
  # Elevation basemap
  geom_raster(
    data = elev_df,
    aes(x = x, y = y, fill = elevation)
  ) +
  scale_fill_hypso_c(palette = "dem_screen", name = "Elevation (m)") +
  
  # Fires 
  geom_sf(
    data = fires_sub,
    fill = "red",
    color = NA,
    alpha = 0.7
  ) +
  
   # Fire numbers
  geom_shadowtext(
    data = fires_sub,
    aes(label = fire_num, geometry = geom),
    nudge_y = fires_sub$nudge_y_val,
    stat = "sf_coordinates",
    size = 5,
    fontface = "bold",
    color = "black",
    bg.color = "white"
  ) +
  #Roads
  geom_sf(
    data = roads,
    fill = NA,
    color = "grey",
    linewidth = 0.5
  ) +
  
  # State outline
  geom_sf(
    data = state,
    fill = NA,
    color = "black",
    linewidth = 0.5
  ) +
  coord_sf(expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.caption = element_text(family = "Times New Roman", size = 12),
    legend.title = element_text(family = "Times New Roman", size = 12),
    legend.text = element_text(family = "Times New Roman", size = 12),
    legend.position = c(0.75, 0.5),  # x, y as proportion of plot area (adjust as needed)
    legend.justification = c(0, 0),
    legend.background = element_rect(fill = "white", color = NA)
  ) 

map
ggsave(
  filename = "documents/ProposalMap.png",
  plot = map,          # your ggplot object
  width = 6,
  height = 5,
  units = "in",
  dpi = 300               # 300 for print, 600 for very high quality
)
```

#Try adding table

```{r}
library(gridExtra)
library(patchwork)

fire_table <- tibble(
  `#` = 1:15,
  `Fire Name` = c("Eight Four Two", "Trail Springs", "Iron", "Calwood", "East Fork",
                  "Beaver", "Little Mesa", "Doe Canyon", "Plateau", "Cameron Peak",
                  "Lake Christine", "Weston Pass", "Duckett", "Junkins", "Little Sand"),
  `Year` = c(2017, 2023, 2023, 2020, 2020, 2010, 2023, 2019, 2018, 2020, 2018, 2018, 2011, 2016, 2012),
  `Area (ha)` = c(2410, 561, 3150, 3981, 706, 1010, 1638, 2991, 7991, 84541, 5061, 5329, 1904, 7698, 10619),
  `Prop. Treated` = c(0.84, 0.42, 0.27, 0.26, 0.23, 0.23, 0.20, 0.18, 0.16, 0.14, 0.14, 0.13, 0.11, 0.10, 0.10)
)

# Split into two halves
table_left <- fire_table[1:8, ]
table_right <- fire_table[9:15, ]

# # Convert each to grob
# grob_left <- tableGrob(table_left, rows = NULL, theme = ttheme_minimal(base_size = 8))
# grob_right <- tableGrob(table_right, rows = NULL, theme = ttheme_minimal(base_size = 8))

# # Place side by side
# table_combined <- arrangeGrob(grob_left, grob_right, ncol = 2, widths = c(1, 1))

# Pad right table with a blank row to match left
table_right_padded <- bind_rows(
  table_right,
  tibble(`#` = NA_integer_, `Fire Name` = " ", `Year` = NA_integer_, `Area (ha)` = NA_integer_, `Prop. Treated` = NA_real_)
) %>%
  mutate(across(where(is.numeric), ~ as.character(.))) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "", .)))

# Also convert left to character for consistency
table_left_chr <- table_left %>%
  mutate(across(everything(), as.character))


my_theme <- ttheme_minimal(
  base_size = 12,
  base_family = "Times New Roman",
  core = list(fg_params = list(fontfamily = "Times New Roman", fontsize = 12, hjust = 0.5, x = 0.5)),
  colhead = list(fg_params = list(fontfamily = "Times New Roman", fontsize = 12, fontface = "bold", hjust = 0.5, x = 0.5))
)


grob_left <- tableGrob(table_left_chr, rows = NULL, theme = my_theme)
grob_right <- tableGrob(table_right_padded, rows = NULL, theme = my_theme)

table_left_chr <- table_left_chr %>%
  mutate(` ` = "")  # blank spacer column

grob_left <- tableGrob(table_left_chr, rows = NULL, theme = my_theme)
table_combined <- cbind(grob_left, grob_right)


# Combine map and table
combined <- map / wrap_elements(full = table_combined, clip = FALSE) +
  plot_layout(heights = c(3, 1)) &
  theme(plot.margin = margin(1, 1, 1, 1),
        panel.spacing = unit(0, "pt"))

ggsave(
  filename = "documents/ProposalMap1.png",
  plot = combined,
  width = 8.5,
  height = 9.5,  # taller to accommodate table
  units = "in",
  dpi = 300
)
```

