---
title: "1d--Fires-Selection"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(root.dir = 'C:/Users/Elizabeth Buhr/Documents/fuels-analysis/')
```

```{r}
#load packages
library(sf) 
library(tidyverse)
```

#Load list of fires with relevant info

```{r}
#read csv
fires <- read.csv("data/processed/my_fires.csv")
colnames(fires)
```

# Filter fires for proportion treated and minimum elevation

So far, the only filters for site/fire selection have been selecting for fires withing the past ~15 years (2010-2024) which intersect with previous fuels reduction. Now, I will impose 2 more filters:
--The proportion of the wildfire treated must be equal to or greater than 0.1, when rounded to two decimal points. This is to improve site sampling within each fire.
--The minimum elevation of the fire must be greater than 1500 m. This is to exclude lower-elevation vegetation types.

```{r}
myfires <- fires %>%
  filter(Proportion_Treated >= 0.1) %>%
  filter(Proportion_Treated < 1) %>%
  filter(Fire_Min_Elev > 1500)

write.csv(myfires, "data/processed/my_fires.csv")

myfiresIDs <- myfires$Event_ID

```

# Update all geopackage layers to reflect fire selection

```{r}
#path to file 
gpkg_path <- "data/processed/fuels_analysis.gpkg"

#check layer name
st_layers(gpkg_path)

#read the layers
fires <- st_read(gpkg_path, layer = "fires")
intersection <- st_read(gpkg_path, layer = "intersection")
treatments <- st_read(gpkg_path, layer = "treatments")

#subset fires
fires_sub <- fires %>% 
  filter(Event_ID %in% myfiresIDs)

#subset the intersection
inter_sub <- intersection %>% 
  filter(Event_ID %in% myfiresIDs)

#subset the treatments
treat_sub <- treatments %>%
  filter(OBJECTID %in% inter_sub$OBJECTID)

```

# Save new geopackage

```{r}
# copy over my fires layer
st_write(
  obj = fires_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "fires",
  driver = "GPKG",
  delete_layer = TRUE
)


# write new interactions layer that includes wildfire
st_write(
  obj = inter_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "intersection",
  driver = "GPKG",
  delete_layer = TRUE
)

# write new treatments layer that includes wildfire
st_write(
  obj = treat_sub,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "treatments",
  driver = "GPKG",
  delete_layer= TRUE
)
```

# Prepare fires for use in Parks GEE methods

```{r}
## Prepares the selected fires perimeter dataset for use in Sean Parks' GEE code for calculating fire severity
st_read("data/processed/fuels_analysis.gpkg", layer = "fires") %>% # import selected perimeters
  mutate(date = as.Date(Ig_Date), "%Y%m%d") %>% # convert ignition date to date object
  mutate(Start_Day=min(lubridate::yday(date))-14) %>% # set first possible date 
  mutate(End_Day=max(lubridate::yday(date))+14) %>% # set last possible date 
  mutate(Fire_Year=as.character(year), Fire_ID=Event_ID) %>% # create variables with column names used in Parks code
  dplyr::select(Fire_ID, Fire_Year, Start_Day, End_Day) %>% # include only Fire ID, Fire Year, Start Day and End Day columns in dataset
  st_write("data/processed/GEE/fires_GEE.shp", delete_dsn = TRUE)


GEEFires <- st_read("data/processed/GEE/fires_GEE.shp")
library(terra)
crs(GEEFires)
```


