---
title: "3a--Site-Selection"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: word_document
---

This document identifies sites within selected fires (fires from 2010-2024, >=10% extent treated area, >1500 min elevation, in Colorado) for multi-scale landscape analysis (plots) and fine-scale progression analysis (transects) following methods from Jones et al 2025. Sites will be used to analyze the effect of treatment extent and distance from treatment edge, among other treatment characteristics and covariates.

Plot Selection:

  1. Generate points with random x-y coordinates in each fire, minimum 500m from the fire perimeter and 1000m away from each other. 
  2. Generate buffered circles surrounding points with varying radii (100, 250, and 500m). Adjusted distances from Jones 2025 due to fewer sampling opportunities in my fires.

Transect Selection:

  1.  Create radial lines from point of ignition to fire perimeter. Generate points 50m apart along these lines.
  2. Identify sequences of 10 points which have 5 points in a treatment and 5 outside.
  3. Thin treatment transects to 500m apart on all sides.
  4. Select control transects (sequence of 10 points outside of a treatment by at least 500m).
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

target_crs <- "EPSG:26913"
```

# Load packages
```{r}
library(tidyverse)
library(sf)
library(terra)
library(lwgeom) #spherical functions for transects
library(RcppRoll) #rolling operations for transects
library(slider) #2nd try sliding operations for transects
```


# MTBS Fire Perimeters (Already selected)
```{r}
fires <- st_read("data/processed/fuels_analysis.gpkg", layer = "fires") %>%
  mutate(
    Area_SqM = as.numeric(st_area(.))   # Get area in square meters
  ) %>%
  st_transform(target_crs)
```

# Point Selection with Perimeter Buffering and Minimum Distance
```{r}
# 1. Shrink the fires by 50m to satisfy the edge requirement
# st_buffer with a negative value creates an inner buffer
fire_interiors <- st_buffer(fires, dist = -50)

# 2. Function to sample points with a minimum distance constraint
sample_with_min_dist <- function(polygon, min_dist = 1000) {
  # Start with an initial random sample (oversample slightly)
  # We use "regular" or "random" then filter
  pts <- st_sample(polygon, size = 1000) #adjust size as desired 
  
  if (length(pts) == 0) return(NULL)
  
  # Thin the points to ensure they are 1000m apart
  # This uses a simple iterative 'greedy' approach
  kept_pts <- pts[1]
  if (length(pts) > 1) {
    for (i in 2:length(pts)) {
      distances <- st_distance(pts[i], kept_pts)
      if (all(as.numeric(distances) >= min_dist)) {
        kept_pts <- c(kept_pts, pts[i])
      }
    }
  }
  return(kept_pts)
}

# 3. Apply to each fire
# Using purrr::map or a loop to keep the points associated with fire IDs
sampled_points_list <- lapply(1:nrow(fire_interiors), function(i) {
  pts <- sample_with_min_dist(fire_interiors[i, ], min_dist = 1000)
  if (is.null(pts)) return(NULL)
  
  # Convert to sf object and keep the original fire's attributes
  st_sf(geometry = pts) %>%
    mutate(Incid_Name = fire_interiors$Incid_Name[i]) %>%
    mutate(Event_ID = fire_interiors$Event_ID[i])
})

# Combine all points into one object
final_points <- do.call(rbind, sampled_points_list)

SumStatsPlots <- final_points %>%
  group_by(Incid_Name) %>%
  arrange(Incid_Name) %>%
  summarize(n = n()) %>%
  st_drop_geometry()
SumStatsPlots
```

# Creating circle plots with varying radii

```{r}
# Define the radii you want
radii <- c(100, 250, 500)

# Create a list of sf objects, one for each radius
buffer_plots <- lapply(radii, function(r) {
  st_buffer(final_points, dist = r) %>%
    mutate(buffer_radius = r)
})

# All plots in one object
all_buffers <- do.call(rbind, buffer_plots) %>%
  mutate(unique_id = row_number())

# Separate Objects
plots_100m <- buffer_plots[[1]]
plots_250m <- buffer_plots[[2]]
plots_500m <- buffer_plots[[3]]


st_write(
  obj = plots_100m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_100m",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_250m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_250m",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_500m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_500m",
  driver = "GPKG",
  delete_layer= TRUE
)
```

# Summary Statistics

```{r}
#Collapse treatment dataset
treatments <- st_read("data/processed/fuels_analysis.gpkg", layer = "treatments") %>%
  st_make_valid() %>%
  st_transform(target_crs)
flat <- st_union(treatments)

# Calculate the total area of each circle plot
all_buffers <- all_buffers %>%
  mutate(total_plot_area = st_area(.))

# Intersect plots with treatments
intersections <- st_intersection(all_buffers, flat)

# Sum treated area per plot
treated_proportions <- intersections %>%
  mutate(treated_area = st_area(.)) %>%
  st_drop_geometry() %>%
  # Adding total_plot_area to the group_by preserves it
  group_by(unique_id, total_plot_area) %>% 
  summarise(total_treated_area = sum(treated_area), .groups = "drop") %>%
  # Now total_plot_area exists for this calculation
  mutate(prop_treated = as.numeric(total_treated_area / total_plot_area))

# Join back to the original buffers to include 0s
# (Plots with no treatment overlap are dropped in st_intersection)
final_analysis <- all_buffers %>%
  st_drop_geometry() %>%
  left_join(treated_proportions) %>%
  mutate(prop_treated = replace_na(prop_treated, 0))
```

```{r}
ggplot(final_analysis, aes(x = prop_treated)) +
  geom_histogram(bins = 20) +
  # Create a separate panel for each radius
  facet_wrap(~buffer_radius, labeller = label_both) + 
  theme_bw() +
  labs(
    title = "Distribution of Treatment Proportion by Buffer Size",
    x = "Proportion of Area Treated",
    y = "Number of Plots"
  ) +
  scale_x_continuous(labels = scales::percent)
```


# Transect Selection

This section selects transects entering and exiting treatment boundaries as well as control transects.

# Transect Generation

## Step 1: Prepare data by finding fire origins and creating a csv with 360 rows and an empty column for each fire. In ArcGIS, create radial lines from the fire origin. Generate points every 50m along each radial line. Create columns with a line ID and distance along line.

### Fire Origin and prep

Before generating lines in ArcGIS, we will prepare our data by filtering our fire occurrence data from InFORM to only include our fires of interest.

InForm (NIFC) was downloaded here: https://data-nifc.opendata.arcgis.com/datasets/nifc::inform-fire-occurrence-data-records/about 

Two fires (Junkins and Lake Christine) have ignition points outside of the fire perimeters. The Junkins point of origin was thus gathered from this report: https://wildfiretoday.com/tag/junkins-fire/. The Lake Christine point of origin was approximated as the southern most point (while the southern most point may be slightly west of the actual origin, due to a change in wind direction it should be more accurate for approximating fire spread).

```{r}
occurrence <- st_read("data/spatial/InFORM_FireOccurrence_Public_1766480872051252493.gpkg") %>%
  st_transform(target_crs) %>%
  mutate(FireDiscoveryDateTime = as.POSIXct(FireDiscoveryDateTime, tz = "UTC")) %>%
  filter(year(FireDiscoveryDateTime) >= 2010,
         year(FireDiscoveryDateTime) <= 2024) %>%
  mutate(occurrence_year = year(FireDiscoveryDateTime)) %>%
  mutate(IncidentName = toupper(IncidentName))%>%
  dplyr::select(occurrence_year, IncidentName)

##Adjust points which are outside of perimeter########################
#Add in Junkins point of origin by selecting western most point
junkcoords <- fires %>%
  filter(Incid_Name == "JUNKINS") %>%
  st_coordinates()
junkx <- min(junkcoords[ ,1]) #Minimum X coordinate = western most, which we can see is the origin from the report
junkorig <- junkcoords[junkcoords[, 1] == junkx, ]
junkorig <- st_sfc(
  st_point(c(junkorig[1], junkorig[2])),
  crs = st_crs(fires)
)
junkorigfin <- st_sf(
  occurrence_year = 2016,
  IncidentName   = "JUNKINS",
  SHAPE           = junkorig
)
#Add Lake Christine by selecting southern most point
lchristcoords <- fires %>%
  filter(Incid_Name == "LAKE CHRISTINE") %>%
  st_coordinates()
lchristy <- min(lchristcoords[ ,2]) #Minimum Y coordinate = southernmost
lchristorig <- lchristcoords[lchristcoords[, 2] == lchristy, ]
lchristorig <- st_sfc(
  st_point(c(lchristorig[1], lchristorig[2])),
  crs = st_crs(fires)
)
lchristorigfin <- st_sf(
  occurrence_year = 2018,
  IncidentName   = "LAKE CHRISTINE",
  SHAPE           = lchristorig
)
#Add our new origin points back to our occurrence dataset
occurrence <- rbind(occurrence, junkorigfin, lchristorigfin)
####################################################################


occurrence_in_fires <- st_join(
  occurrence,
  fires,
  join = st_intersects,
  left = FALSE
) %>%
  filter(occurrence_year == year,
         IncidentName == Incid_Name) %>%
  dplyr::select(c("Event_ID", "Incid_Name", "Ig_Date", "year", "Area_SqM"))

st_write(
  obj = occurrence_in_fires,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "ignition points",
  driver = "GPKG",
  delete_layer= TRUE
)

#We need a csv as well for the ArcGIS methods
# Extract the Projected Coordinates (X and Y)
# This creates new columns for X and Y based on your SHAPE field
coords <- st_coordinates(occurrence_in_fires)
occurrence_in_fires$X_Coord <- coords[,1]
occurrence_in_fires$Y_Coord <- coords[,2]

# Drop the spatial geometry temporarily to work with it as a table
occurrence_in_fires_df <- st_drop_geometry(occurrence_in_fires)

# Calculate a distance for each fire so that the radial lines cover the fire
occurrence_in_fires_df <- occurrence_in_fires_df %>%
  mutate(LineDist = sqrt(Area_SqM) * 3)

# Create a data frame of angles
angles <- data.frame(Bearing = 1:360)

# The "Cross Join": This multiplies your fires by the 360 angles
# Every fire will now have 360 rows, one for each degree.
radial_table <- occurrence_in_fires_df %>%
  crossing(angles)

# Save as a CSV to bring back into ArcGIS Pro
write.csv(radial_table, "data/processed/Radial_Lines_Input.csv", row.names = FALSE)
```

In ArcGIS I created radial lines from the ignition points to the fire perimeter using the "Bearing Distance to Line" tools. This generated 360 geodetic lines to the fire perimeter. The input for generating the lines was a csv made below. Preserve the attributes so that the Event_ID is retained.I then generated points along the lines, and did a spatial join with the fires layer using an attribute filter with Event_ID. This gives me points within the fire perimeter spaced every 50m (planar) from the ignition point. Export the results of the spatial join as AllPoints. You will have to take out the TargetFID and Date fields for export.



## Step 2: Identify which transects (set of 10 sequential points) intersect treatments with 5 points in/out

```{r}
#load points
points <- st_read("data/processed/AllPoints/AllPoints.shp") %>%
  st_transform(target_crs)

#label treated points
spiltmilk1 <- st_intersects(points, flat, sparse = FALSE)
points$is_treated <- rowSums(spiltmilk1) > 0

#identify the treatment progression transects
valid_origins <- points %>%
  group_by(Incid_Name, Bearing) %>% 
  arrange(ORIG_LEN, .by_group = TRUE) %>%
  mutate(
    # Count how many of the NEXT 5 points are treated
    sum_next_5 = roll_sum(as.numeric(is_treated), n = 5, align = "left", fill = NA),
    # Get the sum of the 5 points AFTER those (points 6-10)
    sum_after_5 = lead(sum_next_5, 5)
  ) %>%
  filter(
    (sum_next_5 == 5 & sum_after_5 == 0) | # 5 IN then 5 OUT
    (sum_next_5 == 0 & sum_after_5 == 5)   # 5 OUT then 5 IN
  ) %>%
  mutate(type = ifelse(sum_next_5 == 5, "exiting", "entering")) %>%
  ungroup()
```

## Add points back to identified origins to create transects

```{r}
# Order points and index within each radial
pts <- points %>%
  arrange(Event_ID, ORIG_FID, ORIG_SEQ) %>%
  group_by(Event_ID, ORIG_FID) %>%
  mutate(row = row_number()) %>%
  ungroup()

valid_origins <- st_drop_geometry(valid_origins)
# Get the origin row index for each valid origin
origins <- pts %>%
  semi_join(valid_origins,
            by = c("Event_ID", "ORIG_FID", "ORIG_SEQ")) %>%
  dplyr::select(Event_ID, ORIG_FID, origin_row = row) %>%
  st_drop_geometry()

# Keep origin + next 9 points (10 total)
transects <- pts %>%
  inner_join(origins, by = c("Event_ID", "ORIG_FID")) %>%
  filter(row >= origin_row & row <= origin_row + 9) %>%
  group_by(Event_ID, ORIG_FID, origin_row) %>%
  filter(
    n() == 10,                                # must have 10 points
    (max(ORIG_SEQ) - min(ORIG_SEQ)) == 9,     # span must be 9
    n_distinct(ORIG_SEQ) == 10                # no missing sequence numbers
  ) %>%
  ungroup()
```

## Make transects into line objects and enforce spacing

```{r}
transects <- transects %>%
  group_by(Event_ID, Incid_Name, ORIG_FID, origin_row) %>%
  mutate(TransectID = cur_group_id()) %>%
  ungroup()

transect_lines <- transects %>%
  arrange(Event_ID, Incid_Name, ORIG_FID, TransectID, ORIG_SEQ) %>%  
  group_by(TransectID, Event_ID, Incid_Name, ORIG_FID) %>%
  summarize(geometry = st_cast(st_combine(geometry), "LINESTRING"),
            .groups = "drop")

#Remove transect lines within 50m of fire perimeter to avoid effects of fire suppression
  # logical matrix: rows = lines, cols = polygons
within_mat <- st_covered_by(transect_lines, fire_interiors, sparse = FALSE)

  # keep rows that are within at least one polygon
transects_lines <- transect_lines[rowSums(within_mat) > 0, ]

# Enforce 500m spacing between transect lines
keep_indices <- c()
remaining_indices <- 1:nrow(transect_lines)

while (length(remaining_indices) > 0) {
  # Pick the first (highest priority) line remaining
  current_idx <- remaining_indices[1]
  keep_indices <- c(keep_indices, current_idx)
  
  # Find all lines within 500m of this line
  too_close <- st_is_within_distance(transect_lines[current_idx, ], 
                                    transect_lines[remaining_indices, ], 
                                    dist = 500)[[1]]
  
  # Remove the current line and all "too close" lines from the pool
  remaining_indices <- remaining_indices[-too_close]
}

# Filter the original dataset
thinned_transects <- transect_lines[keep_indices, ]

st_write(thinned_transects, "data/processed/TreatmentTransects/FinalTreatmentTransects.shp", delete_dsn = TRUE)

SumStatsAfter <- thinned_transects %>%
  arrange(Incid_Name) %>%
  group_by(Incid_Name) %>%
  summarize(n = n()) %>%
  st_drop_geometry()
SumStatsAfter
```

## Add points back to transects now that they are thinned

```{r}
points_thinned <- transects %>%
  semi_join(thinned_transects %>% st_drop_geometry(), 
            by = c("Event_ID", "ORIG_FID", "TransectID"))

st_write(points_thinned, 
         "data/processed/TreatmentTransects/FinalTreatmentTransectPoints.shp",
         delete_dsn = TRUE)
```

## Select control transects 

Following the same method as for treatment transects. 

```{r}
#identify Control transects (sequence of 10 not intersecting)
control_origins <- points %>%
  group_by(Incid_Name, Bearing) %>% 
  arrange(ORIG_LEN, .by_group = TRUE) %>%
  mutate(
    # Count how many of the next 10 points are treated
    sum_next_10 = roll_sum(as.numeric(is_treated), n = 10, align = "left", fill = NA)
  ) %>%
  filter(
    (sum_next_10 == 0)
  ) %>%
  mutate(type = "control") %>%
  ungroup()
```

## Add points back to identified origins to create transects

```{r}
# Order points and index within each radial
pts <- points %>%
  arrange(Event_ID, ORIG_FID, ORIG_SEQ) %>%
  group_by(Event_ID, ORIG_FID) %>%
  mutate(row = row_number()) %>%
  ungroup()

control_origins <- st_drop_geometry(control_origins)
# Get the origin row index for each valid origin
control_origins <- pts %>%
  semi_join(control_origins,
            by = c("Event_ID", "ORIG_FID", "ORIG_SEQ")) %>%
  dplyr::select(Event_ID, ORIG_FID, origin_row = row) %>%
  st_drop_geometry()

# Keep origin + next 9 points (10 total)
control_transects <- pts %>%
  inner_join(control_origins, by = c("Event_ID", "ORIG_FID")) %>%
  filter(row >= origin_row & row <= origin_row + 9) %>%
  group_by(Event_ID, ORIG_FID, origin_row) %>%
  filter(
    n() == 10,                                # must have 10 points
    (max(ORIG_SEQ) - min(ORIG_SEQ)) == 9,     # span must be 9
    n_distinct(ORIG_SEQ) == 10                # no missing sequence numbers
  ) %>%
  ungroup()
```

## Make transects into line objects and enforce spacing

```{r}
control_transects <- control_transects %>%
  group_by(Event_ID, Incid_Name, ORIG_FID, origin_row) %>%
  mutate(TransectID = cur_group_id()) %>%
  ungroup()

control_transect_lines <- control_transects %>%
  arrange(Event_ID, Incid_Name, ORIG_FID, TransectID, ORIG_SEQ) %>%  
  group_by(TransectID, Event_ID, Incid_Name, ORIG_FID) %>%
  summarize(geometry = st_cast(st_combine(geometry), "LINESTRING"),
            .groups = "drop")

#Remove transect lines within 50m of fire perimeter to avoid effects of fire suppression
  # logical matrix: rows = lines, cols = polygons
within_mat2 <- st_covered_by(control_transect_lines, fire_interiors, sparse = FALSE)

  # keep rows that are within at least one polygon
control_transect_lines <- control_transect_lines[rowSums(within_mat2) > 0, ]

# Enforce 500m spacing between transect lines
keep_indices <- c()
remaining_indices <- 1:nrow(control_transect_lines)

while (length(remaining_indices) > 0) {
  # Pick the first (highest priority) line remaining
  current_idx <- remaining_indices[1]
  keep_indices <- c(keep_indices, current_idx)
  
  # Find all lines within 500m of this line
  too_close <- st_is_within_distance(control_transect_lines[current_idx, ], 
                                    control_transect_lines[remaining_indices, ], 
                                    dist = 500)[[1]]
  
  # Remove the current line and all "too close" lines from the pool
  remaining_indices <- remaining_indices[-too_close]
}

# Filter the original dataset
control_thinned_transects <- control_transect_lines[keep_indices, ]

st_write(control_thinned_transects, "data/processed/ControlTransects/FinalControlTransects.shp", delete_dsn = TRUE)

SumStatsAfter <- control_thinned_transects %>%
  arrange(Incid_Name) %>%
  group_by(Incid_Name) %>%
  summarize(n = n()) %>%
  st_drop_geometry()
SumStatsAfter
```

## Add points back to transects now that they are thinned

```{r}
control_points_thinned <- control_transects %>%
  semi_join(control_thinned_transects %>% st_drop_geometry(), 
            by = c("Event_ID", "ORIG_FID", "TransectID"))

st_write(control_points_thinned, 
         "data/processed/ControlTransects/FinalTreatmentTransectPoints.shp",
         delete_dsn = TRUE)
```

