---
title: "3a--Site-Selection"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: word_document
---

This document identifies sites within selected fires (fires from 2010-2024, >=10% extent treated area, >1500 min elevation, in Colorado) for multi-scale landscape analysis (plots) and fine-scale progression analysis (transects) following methods from Jones et al 2025. Sites will be used to analyze the effect of treatment extent and distance from treatment edge, among other treatment characteristics and covariates.

Plot Selection:

  1. Generate points with random x-y coordinates in each fire, minimum 500m from the fire perimeter and 1000m away from each other. 
  2. Generate buffered circles surrounding points with varying radii (100, 250, and 500m). Adjusted distances from Jones 2025 due to fewer sampling opportunities in my fires.

Transect Selection:

  1.  Create radial lines from point of ignition to fire perimeter. Generate points 50m apart along these lines.
  2. Identify sequences of 10 points which have 5 points in a treatment and 5 outside.
  3. Thin treatment transects to 500m apart on all sides.
  4. Select control transects (sequence of 10 points outside of a treatment by at least 500m).
  

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Load packages
```{r}
library(tidyverse)
library(sf)
library(terra)
library(lwgeom) #spherical functions for transects
library(RcppRoll) #rolling operations for transects
library(slider) #2nd try sliding operations for transects
library(exactextractr) #get tcc for plots and transects
```

```{r}
target_crs <- "EPSG:26913"
```

Read in MTBS Fire Perimeters (Already filtered)
```{r}
fires <- st_read("data/processed/fuels_analysis.gpkg", layer = "fires") %>%
  mutate(
    Area_SqM = as.numeric(st_area(.))   # Get area in square meters
  ) %>%
  st_transform(target_crs)
```

```{r}
#Read in 1999 TCC Cover to only select sites with >10% tree cover prior to any treatment or fire

tcc <- rast("D:/Spatial_Data/NLCD/nlcd_tcc_CONUS_1999_v2023-5_wgs84/nlcd_tcc_conus_wgs84_v2023-5_19990101_19991231.tif")
```

# Multi-Scale Plot Selection
## Point Selection with Perimeter Buffering and Minimum Distance
```{r plot-selection}
# 1. Shrink the fires by 550m to satisfy the edge requirement
# st_buffer with a negative value creates an inner buffer
fire_interiors <- st_buffer(fires, dist = -550)

# 2. Functions
sample_with_min_dist <- function(polygon, min_dist = 1000) {
  pts <- st_sample(polygon, size = 500)
  if (length(pts) == 0) return(NULL)
  kept_pts <- pts[1]
  if (length(pts) > 1) {
    for (i in 2:length(pts)) {
      distances <- st_distance(pts[i], kept_pts)
      if (all(as.numeric(distances) >= min_dist)) {
        kept_pts <- c(kept_pts, pts[i])
      }
    }
  }
  return(kept_pts)
}

radii <- c(100, 250, 500)

# 3. Per-fire iterative selection loop
final_points_list  <- vector("list", nrow(fire_interiors))
final_buffers_list <- vector("list", nrow(fire_interiors))

for (i in 1:nrow(fire_interiors)) {
  
  fire        <- fire_interiors[i, ]
  event_id    <- fire$Event_ID
  incid_name  <- fire$Incid_Name
  fire_year   <- fire$year
  
  # --- Initial candidate sample ---
  pts_raw <- sample_with_min_dist(fire, min_dist = 1000)
  if (is.null(pts_raw)) next
  
  candidates <- st_sf(geometry = pts_raw) %>%
    mutate(
      Incid_Name = incid_name,
      Event_ID   = event_id,
      year       = fire_year,
      point_id   = paste0(substr(incid_name, 1, 4), "_", row_number())
    )
  
  # --- Iterative selection: keep adding points that fit + pass TCC ---
  kept      <- candidates[0, ]
  remaining <- candidates
  
  repeat {
    newly_added <- remaining[0, ]
    
    for (j in 1:nrow(remaining)) {
      candidate <- remaining[j, ]
      
      # Check minimum distance against already-kept + newly_added this round
      if (nrow(kept) == 0 && nrow(newly_added) == 0) {
        distance_ok <- TRUE
      } else {
        existing <- rbind(kept, newly_added)
        dists    <- st_distance(candidate, existing)
        distance_ok <- all(as.numeric(dists) >= 1000)
      }
      
      if (distance_ok) {
        # Build buffers and check TCC for all 3 scales
        bufs <- lapply(radii, function(r) {
          st_buffer(candidate, dist = r) %>% mutate(buffer_radius = r)
        }) %>% do.call(rbind, .)
        
        bufs_proj        <- st_transform(bufs, crs(tcc))
        bufs_proj$mean_tcc <- exact_extract(tcc, bufs_proj, "mean")
        
        if (all(bufs_proj$mean_tcc >= 10, na.rm = TRUE)) {
          newly_added <- rbind(newly_added, candidate)
        }
      }
    }
    
    if (nrow(newly_added) == 0) break
    
    kept      <- rbind(kept, newly_added)
    remaining <- remaining[!remaining$point_id %in% kept$point_id, ]
    if (nrow(remaining) == 0) break
  }
  
  if (nrow(kept) == 0) next
  
  # --- Build final buffers for kept points ---
  kept_buffers <- lapply(radii, function(r) {
    st_buffer(kept, dist = r) %>% mutate(buffer_radius = r)
  }) %>%
    do.call(rbind, .) %>%
    st_transform(crs(tcc)) %>%
    mutate(mean_tcc = exact_extract(tcc, ., "mean")) %>%
    st_transform(target_crs)
  
  final_points_list[[i]]  <- kept
  final_buffers_list[[i]] <- kept_buffers
  
  message("Fire ", i, " (", incid_name, "): ", nrow(kept), " plots selected")
}

# --- Combine ---
final_points <- do.call(rbind, final_points_list)
all_buffers  <- do.call(rbind, final_buffers_list) %>%
  mutate(unique_id = row_number())

# --- Summary ---
final_points %>%
  st_drop_geometry() %>%
  group_by(Incid_Name) %>%
  summarise(n = n()) %>%
  arrange(Incid_Name)
```
 Note: the above code dropped Iron because no points met the minimum TCC. We will also need to drop Trail Springs because it only has one plot.
 
## Save objects

```{r}
plots_100m <- all_buffers %>% filter(buffer_radius == 100)
plots_250m <- all_buffers %>% filter(buffer_radius == 250)
plots_500m <- all_buffers %>% filter(buffer_radius == 500)


st_write(
  obj = all_buffers,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "all_buffers_selected",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_100m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_100m_selected",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_250m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_250m_selected",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_500m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_500m_selected",
  driver = "GPKG",
  delete_layer= TRUE
)
```

## Summary Statistics

```{r treatment-extent}
# #Collapse treatment dataset
# treatments <- st_read("data/processed/fuels_analysis.gpkg", layer = "treatments") %>%
#   st_make_valid() %>%
#   st_transform(target_crs) %>%
#   filter(ACTIVITY != "Plant Trees")
# flat <- st_union(treatments)
# 
# # Calculate the total area of each circle plot
# all_buffers <- all_buffers %>%
#   mutate(total_plot_area = st_area(.))
# 
# # Intersect plots with treatments
# intersections <- st_intersection(all_buffers, flat)
# 
# # Sum treated area per plot
# treated_proportions <- intersections %>%
#   mutate(treated_area = st_area(.)) %>%
#   st_drop_geometry() %>%
#   # Adding total_plot_area to the group_by preserves it
#   group_by(unique_id, total_plot_area) %>% 
#   summarise(total_treated_area = sum(treated_area), .groups = "drop") %>%
#   # Now total_plot_area exists for this calculation
#   mutate(prop_treated = as.numeric(total_treated_area / total_plot_area))
# 
# # Join back to the original buffers to include 0s
# # (Plots with no treatment overlap are dropped in st_intersection)
# final_analysis <- all_buffers %>%
#   st_drop_geometry() %>%
#   left_join(treated_proportions) %>%
#   mutate(prop_treated = replace_na(prop_treated, 0))

# ggplot(final_analysis, aes(x = prop_treated)) +
#   geom_histogram(bins = 20) +
#   # Create a separate panel for each radius
#   facet_wrap(~buffer_radius, labeller = label_both) + 
#   theme_bw() +
#   labs(
#     title = "Distribution of Treatment Proportion by Buffer Size",
#     x = "Proportion of Area Treated",
#     y = "Number of Plots"
#   ) +
#   scale_x_continuous(labels = scales::percent)
```


# Fine-Scale Progression Transect Selection

This section selects transects entering and exiting treatment boundaries as well as control transects.

## Step 1: Prepare data by finding fire origins and creating a csv with 360 rows and an empty column for each fire. In ArcGIS, create radial lines from the fire origin. Generate points every 50m along each radial line. Create columns with a line ID and distance along line.

### Fire Origin and prep

Before generating lines in ArcGIS, we will prepare our data by filtering our fire occurrence data from InFORM to only include our fires of interest.

InForm (NIFC) was downloaded here: https://data-nifc.opendata.arcgis.com/datasets/nifc::inform-fire-occurrence-data-records/about 

Two fires (Junkins and Lake Christine) have ignition points outside of the fire perimeters. The Junkins point of origin was thus gathered from this report: https://wildfiretoday.com/tag/junkins-fire/. The Lake Christine point of origin was approximated as the southern most point (while the southern most point may be slightly west of the actual origin, due to a change in wind direction it should be more accurate for approximating fire spread).

```{r occurrence-prep}
# occurrence <- st_read("D:/Spatial_Data/InFORM_FireOccurrence_Public_1766480872051252493.gpkg") %>%
#   st_transform(target_crs) %>%
#   mutate(FireDiscoveryDateTime = as.POSIXct(FireDiscoveryDateTime, tz = "UTC")) %>%
#   filter(year(FireDiscoveryDateTime) >= 2010,
#          year(FireDiscoveryDateTime) <= 2024) %>%
#   mutate(occurrence_year = year(FireDiscoveryDateTime)) %>%
#   mutate(IncidentName = toupper(IncidentName))%>%
#   dplyr::select(occurrence_year, IncidentName)
# 
# ##Adjust points which are outside of perimeter########################
# #Add in Junkins point of origin by selecting western most point
# junkcoords <- fires %>%
#   filter(Incid_Name == "JUNKINS") %>%
#   st_coordinates()
# junkx <- min(junkcoords[ ,1]) #Minimum X coordinate = western most, which we can see is the origin from the report
# junkorig <- junkcoords[junkcoords[, 1] == junkx, ]
# junkorig <- st_sfc(
#   st_point(c(junkorig[1], junkorig[2])),
#   crs = st_crs(fires)
# )
# junkorigfin <- st_sf(
#   occurrence_year = 2016,
#   IncidentName   = "JUNKINS",
#   SHAPE           = junkorig
# )
# #Add Lake Christine by selecting southern most point
# lchristcoords <- fires %>%
#   filter(Incid_Name == "LAKE CHRISTINE") %>%
#   st_coordinates()
# lchristy <- min(lchristcoords[ ,2]) #Minimum Y coordinate = southernmost
# lchristorig <- lchristcoords[lchristcoords[, 2] == lchristy, ]
# lchristorig <- st_sfc( #make spatial object from list
#   st_point(c(lchristorig[1], lchristorig[2])),
#   crs = st_crs(fires)
# )
# lchristorigfin <- st_sf(
#   occurrence_year = 2018,
#   IncidentName   = "LAKE CHRISTINE",
#   SHAPE           = lchristorig
# )
# #Add our new origin points back to our occurrence dataset
# occurrence <- rbind(occurrence, junkorigfin, lchristorigfin)
# ####################################################################
# 
# 
# occurrence_in_fires <- st_join(
#   occurrence,
#   fires,
#   join = st_intersects,
#   left = FALSE
# ) %>%
#   filter(occurrence_year == year,
#          IncidentName == Incid_Name) %>%
#   dplyr::select(c("Event_ID", "Incid_Name", "Ig_Date", "year", "Area_SqM"))
# 
# st_write(
#   obj = occurrence_in_fires,
#   dsn = "data/processed/fuels_analysis.gpkg",
#   layer = "ignition points",
#   driver = "GPKG",
#   delete_layer= TRUE
# )
# 
# #We need a csv as well for the ArcGIS methods
# # Extract the Projected Coordinates (X and Y)
# # This creates new columns for X and Y based on your SHAPE field
# coords <- st_coordinates(occurrence_in_fires)
# occurrence_in_fires$X_Coord <- coords[,1]
# occurrence_in_fires$Y_Coord <- coords[,2]
# 
# # Drop the spatial geometry temporarily to work with it as a table
# occurrence_in_fires_df <- st_drop_geometry(occurrence_in_fires)
# 
# # Calculate a distance for each fire so that the radial lines cover the fire
# occurrence_in_fires_df <- occurrence_in_fires_df %>%
#   mutate(LineDist = sqrt(Area_SqM) * 3)
# 
# # Create a data frame of angles
# angles <- data.frame(Bearing = 1:360)
# 
# # The "Cross Join": This multiplies your fires by the 360 angles
# # Every fire will now have 360 rows, one for each degree.
# radial_table <- occurrence_in_fires_df %>%
#   crossing(angles)
# 
# # Save as a CSV to bring back into ArcGIS Pro
# write.csv(radial_table, "data/processed/Radial_Lines_Input.csv", row.names = FALSE)
```

In ArcGIS I created radial lines from the ignition points to the fire perimeter using the "Bearing Distance to Line" tools. This generated 360 geodetic lines to the fire perimeter. The input for generating the lines was a csv made above. Preserve the attributes so that the Event_ID is retained.I then generated points along the lines, and did a spatial join with the fires layer using an attribute filter with Event_ID. This gives me points within the fire perimeter spaced every 50m (planar) from the ignition point. Export the results of the spatial join as AllPoints. You will have to take out the TargetFID and Date fields for export.



## Step 2: Identify which transects (set of 10 sequential points) intersect treatments with 5 points in/out, and which transects are controls

### Identify origins

```{r}
#load points
points <- st_read("data/processed/AllPoints/AllPoints.shp") %>%
  st_transform(crs(tcc))

tcc <- as.numeric(tcc)

#Remove points with <10% 1999 TCC
points$tcc <- terra::extract(tcc, vect(points))[, 2]

# Filter
points <- points %>%
  filter(tcc >= 10)

points <- st_transform(points, target_crs)

treatments <- st_read("data/processed/fuels_analysis.gpkg", layer = "treatments") %>%
  st_make_valid() %>%
  st_transform(target_crs)
flat <- st_union(treatments)

#label treated points
spiltmilk1 <- st_intersects(points, flat, sparse = FALSE)
points$is_treated <- rowSums(spiltmilk1) > 0

#identify the treatment progression transects
valid_origins <- points %>%
  group_by(Incid_Name, Bearing) %>% 
  arrange(ORIG_LEN, .by_group = TRUE) %>% #arrange sequentially, by length from origin within each fire along each line
  mutate(
    # Count how many of the NEXT 5 points are treated
    sum_next_5 = roll_sum(as.numeric(is_treated), n = 5, align = "left", fill = NA),
    # Get the sum of the 5 points AFTER those (points 6-10)
    sum_after_5 = lead(sum_next_5, 5),
    # Identify transects where all 10 points are not treated
    sum_next_10 = roll_sum(as.numeric(is_treated, n = 10, align = "left", fill = NA))
  ) %>%
  filter(
    (sum_next_5 == 5 & sum_after_5 == 0) | # 5 IN then 5 OUT
    (sum_next_5 == 0 & sum_after_5 == 5) |  # 5 OUT then 5 IN
    (sum_next_10 == 0) #Control
  ) %>%
  mutate(type = case_when(
    sum_next_5 == 5 & sum_after_5 == 0 ~ "exiting",
    sum_next_5 == 0 & sum_after_5 == 5 ~ "entering",
    sum_next_10 == 0 ~ "control"
  )) %>%
  ungroup()
```

### Add points back to identified origins to create transects

```{r}
# Order points and index within each radial
pts <- points %>%
  arrange(Event_ID, ORIG_FID, ORIG_SEQ) %>%
  group_by(Event_ID, ORIG_FID) %>%
  mutate(row = row_number()) %>%
  ungroup()

valid_origins <- st_drop_geometry(valid_origins)
# Get the origin row index for each valid origin
origins <- pts %>%
  inner_join(valid_origins,
            dplyr::select(Event_ID, ORIG_FID, ORIG_SEQ, type),
            by = c("Event_ID", "ORIG_FID", "ORIG_SEQ")) %>%
  dplyr::select(Event_ID, ORIG_FID, origin_row = row, type) %>%
  st_drop_geometry()

# Keep origin + next 9 points (10 total)
transect_pts <- pts %>%
  inner_join(origins, by = c("Event_ID", "ORIG_FID")) %>%
  filter(row >= origin_row & row <= origin_row + 9) %>% #must be in order
  group_by(Event_ID, ORIG_FID, origin_row, type) %>%
  filter(
    n() == 10,                                # must have 10 points
    (max(ORIG_SEQ) - min(ORIG_SEQ)) == 9,     # span must be 9
    n_distinct(ORIG_SEQ) == 10                # no missing sequence numbers
  ) %>%
  mutate(Point_ID = row_number()) %>%
  ungroup()
```

## Step 3: Make transects into line objects and enforce spacing

```{r}
transect_pts <- transect_pts %>%
  group_by(Event_ID, Incid_Name, ORIG_FID, origin_row, type) %>%
  mutate(Transect_ID = cur_group_id()) %>%
  ungroup()

transect_lines <- transect_pts %>%
  arrange(Event_ID, Incid_Name, ORIG_FID, Transect_ID, ORIG_SEQ) %>%  
  group_by(Transect_ID, Event_ID, Incid_Name, ORIG_FID) %>%
  summarize(geometry = st_cast(st_combine(geometry), "LINESTRING"),
            type = first(type),
            .groups = "drop") 

#Treatment transects should be prioritized in selection, which can be done by ordering them first. Use type column to create treated column then arrange.

transect_lines <- transect_lines %>%
  mutate(treatment = ifelse(type == "control", FALSE, TRUE)) %>%
  arrange(desc(treatment))


#Remove transect lines within 50m of fire perimeter to avoid effects of fire suppression
fire_interiors2 <- st_buffer(fires, dist = -50)
transect_lines <- transect_lines %>%
  st_filter(fire_interiors2, .predicate = st_within)

# Enforce 500m spacing between transect lines
keep_indices <- c()
remaining_indices <- 1:nrow(transect_lines)

while (length(remaining_indices) > 0) {
  # Pick the first (highest priority) line remaining
  current_idx <- remaining_indices[1]
  keep_indices <- c(keep_indices, current_idx)
  
  # Find all lines within 500m of this line
  too_close <- st_is_within_distance(transect_lines[current_idx, ], 
                                    transect_lines[remaining_indices, ], 
                                    dist = 500)[[1]]
  
  # Remove the current line and all "too close" lines from the pool
  remaining_indices <- remaining_indices[-too_close]
}

# Filter the original dataset
thinned_transects <- transect_lines[keep_indices, ]
```

### Filter control transects to equal the number of treatment transects
```{r}
#First buffer treatments by 500m and remove the control transects that are too close
treatments_buffered <- st_buffer(flat, dist = 500)

  #Separate the treatment
treated_yes <- thinned_transects %>% 
  filter(treatment == TRUE)

  #Filter the FALSE rows (only keep if NOT within treatments_buffered)
treated_no_filtered <- thinned_transects %>% 
  filter(treatment == FALSE) %>%
  # st_filter keeps what intersects; we use [!apply(...)] or st_disjoint
  st_filter(treatments_buffered, .predicate = st_disjoint)

  #Combine them back into one object
thinned_transects <- bind_rows(treated_yes, treated_no_filtered)

#Randomly select control transects to equal the number of treatment transects

balanced_transects <- thinned_transects %>%
  # 1. Group by the fire ID and the treatment status
  group_by(Event_ID) %>%
  group_modify(~ {
    # Count how many treatment transects exist in this specific fire
    n_treated <- sum(.x$treatment == TRUE)
    
    # Separate the treated and control groups for this fire
    treated_rows <- .x %>% filter(treatment == TRUE)
    control_rows <- .x %>% filter(treatment == FALSE)
    
    # 2. Randomly sample from controls to match n_treated
    # Use pmin to avoid errors if a fire has fewer controls than treated lines
    sampled_controls <- control_rows %>%
      slice_sample(n = pmin(nrow(control_rows), n_treated))
    
    # 3. Combine them back for this fire
    bind_rows(treated_rows, sampled_controls)
  }) %>%
  ungroup()

SumStatsTransects <- balanced_transects %>%
  arrange(Incid_Name) %>%
  group_by(Incid_Name, type) %>%
  summarize(n = n()) %>%
  st_drop_geometry()
SumStatsTransects
```

### Add points back to transects now that they are thinned

```{r}
transects_final <- transect_pts %>%
  semi_join(balanced_transects %>% st_drop_geometry(), 
            by = c("Event_ID", "ORIG_FID", "Transect_ID")) %>%
  dplyr::select(Event_ID, Incid_Name, year, Transect_ID, Point_ID, type)

st_write(transects_final, 
         "data/processed/Transects/FinalTransectsALL.shp",
         delete_dsn = TRUE)
```

# Save object in GPKG

```{r}
st_write(
  obj = transects_final,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "transects_selected",
  driver = "GPKG",
  delete_layer= TRUE
)
```

