---
title: "3a--Site-Selection"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: word_document
---

This document will perform analysis for random site selection of plots with varying radii in selected fires (2010-2024, >=10% extent treated area, >1500 min elevation, in Colorado). It will use MTBS perimeters to generate points. 

  1. Generate points with random x-y coordinates in each fire, minimum 1000m from the fire perimeter and 2000m away from each other. 
  2. Generate buffered circles surrounding points with varying radii (100m, 500m, 1000m). Following distance from Jones 2025 but adding one smaller radii and removing greater radius in anticipation of fewer sampling opportunities in my fires. Updated to 100, 250, and 500m
  
After these plots are selected, I will summarize response and predictor variables at each plot and scale. For example, proportion treated as a predictor variable and proportion burned at high severity as a response. 
  



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Load packages
```{r}
library(tidyverse)
library(sf)
library(terra)
library(lwgeom) #spherical functions for transects
library(RcppRoll) #rolling operations for transects
library(slider) #2nd try sliding operations for transects
```


# MTBS Fire Perimeters
```{r}
fires <- st_read("data/processed/fuels_analysis.gpkg", layer = "fires") %>%
  mutate(
    Area_SqM = as.numeric(st_area(.))   # Get area in square meters
  )
```

# Point Selection with Perimeter Buffering and Minimum Distance
```{r}
# 1. Shrink the fires by 1000m to satisfy the edge requirement
# st_buffer with a negative value creates an inner buffer
fire_interiors <- st_buffer(fires, dist = -500)

# Filter out any fires that became too small/disappeared after buffering
fire_interiors <- fire_interiors[!st_is_empty(fire_interiors), ]

# 2. Function to sample points with a minimum distance constraint
sample_with_min_dist <- function(polygon, min_dist = 1000) {
  # Start with an initial random sample (oversample slightly)
  # We use "regular" or "random" then filter
  pts <- st_sample(polygon, size = 50) # Adjust size based on fire area
  
  if (length(pts) == 0) return(NULL)
  
  # Thin the points to ensure they are 2000m apart
  # This uses a simple iterative 'greedy' approach
  kept_pts <- pts[1]
  if (length(pts) > 1) {
    for (i in 2:length(pts)) {
      distances <- st_distance(pts[i], kept_pts)
      if (all(as.numeric(distances) >= min_dist)) {
        kept_pts <- c(kept_pts, pts[i])
      }
    }
  }
  return(kept_pts)
}

# 3. Apply to each fire
# Using purrr::map or a loop to keep the points associated with fire IDs
sampled_points_list <- lapply(1:nrow(fire_interiors), function(i) {
  pts <- sample_with_min_dist(fire_interiors[i, ], min_dist = 1000)
  if (is.null(pts)) return(NULL)
  
  # Convert to sf object and keep the original fire's attributes
  st_sf(geometry = pts) %>%
    mutate(fire_id = fire_interiors$id_field[i]) # Replace with your ID column name
})

# Combine all points into one object
final_points <- do.call(rbind, sampled_points_list)

st_write(
  obj = final_points,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots",
  driver = "GPKG",
  delete_layer= TRUE
)
```

# Creating circle plots with varying radii

```{r}
# 1. Define the radii you want
radii <- c(100, 250, 500)

# 2. Create a list of sf objects, one for each radius
# We use 'lapply' to iterate through the radii
buffer_plots <- lapply(radii, function(r) {
  st_buffer(final_points, dist = r) %>%
    mutate(buffer_radius = r)
})

# 3. If you want them all in one single object:
all_buffers <- do.call(rbind, buffer_plots) %>%
  mutate(unique_id = row_number())

# 4. If you prefer separate objects in your environment:
plots_100m <- buffer_plots[[1]]
plots_250m <- buffer_plots[[2]]
plots_500m <- buffer_plots[[3]]


st_write(
  obj = plots_100m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_100m",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_250m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_250m",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = plots_500m,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "plots_500m",
  driver = "GPKG",
  delete_layer= TRUE
)
```

# Summary Statistics

```{r}
# Read in treatment layer
treatments <- st_read("data/processed/fuels_analysis.gpkg", layer = "treatments")

# Calculate the total area of each circle plot
all_buffers <- all_buffers %>%
  mutate(total_plot_area = st_area(.))

# Flatten the treatments into a single footprint
treatments_union <- st_union(st_make_valid(treatments))

# Intersect plots with treatments
intersections <- st_intersection(all_buffers, treatments_union)

# Sum treated area per plot
treated_proportions <- intersections %>%
  mutate(treated_area = st_area(.)) %>%
  st_drop_geometry() %>%
  # Adding total_plot_area to the group_by preserves it
  group_by(unique_id, total_plot_area) %>% 
  summarise(total_treated_area = sum(treated_area), .groups = "drop") %>%
  # Now total_plot_area exists for this calculation
  mutate(prop_treated = as.numeric(total_treated_area / total_plot_area))

# 5. Join back to the original buffers to include 0s
# (Plots with no treatment overlap are dropped in st_intersection)
final_analysis <- all_buffers %>%
  st_drop_geometry() %>%
  left_join(treated_proportions) %>%
  mutate(prop_treated = replace_na(prop_treated, 0))
```

```{r}
ggplot(final_analysis, aes(x = prop_treated)) +
  geom_histogram(fill = "steelblue", color = "white", bins = 20) +
  # Create a separate panel for each radius
  facet_wrap(~buffer_radius, labeller = label_both) + 
  theme_minimal() +
  labs(
    title = "Distribution of Treatment Proportion by Buffer Size",
    x = "Proportion of Area Treated (0.0 - 1.0)",
    y = "Number of Plots"
  ) +
  scale_x_continuous(labels = scales::percent)
```


# Transect Selection

This section selects transects entering and exiting treatment boundaries and control transects following methods outlined in Jones et al 2025. 

## Fire Origin and prep

The analysis begins by creating radial lines from the fire origin to the fire perimeter. We will start by filtering our fire occurrence data from InFORM to only include our fires of interest.

InForm was downloaded here: https://data-nifc.opendata.arcgis.com/datasets/nifc::inform-fire-occurrence-data-records/about 

```{r}
occurrence <- st_read("data/spatial/InFORM_FireOccurrence_Public_1766480872051252493.gpkg")

occurrence <- st_transform(occurrence, crs(fires))

occurrence <- occurrence %>%
  mutate(FireDiscoveryDateTime = as.POSIXct(FireDiscoveryDateTime, tz = "UTC")) %>%
  filter(year(FireDiscoveryDateTime) >= 2010,
         year(FireDiscoveryDateTime) <= 2024) %>%
  mutate(occurrence_year = year(FireDiscoveryDateTime)) %>%
  mutate(IncidentName = toupper(IncidentName))

occurrence_in_fires <- st_join(
  occurrence,
  fires,
  join = st_within,
  left = FALSE
) %>%
  filter(occurrence_year == year,
         IncidentName == Incid_Name) %>%
  select(c("Event_ID", "Incid_Name", "InitialLatitude", "InitialLongitude", "Ig_Date", "year", "Area_SqM", "SHAPE" ))

st_write(
  obj = occurrence_in_fires,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "ignition points",
  driver = "GPKG",
  delete_layer= TRUE
)

#We need a csv as well for the ArcGIS methods
# Extract the Projected Coordinates (X and Y)
# This creates new columns for X and Y based on your SHAPE field
coords <- st_coordinates(occurrence_in_fires)
occurrence_in_fires$X_Coord <- coords[,1]
occurrence_in_fires$Y_Coord <- coords[,2]

# Drop the spatial geometry temporarily to work with it as a table
occurrence_in_fires_df <- st_drop_geometry(occurrence_in_fires)

# Calculate a distance for each fire so that the radial lines cover the fire
occurrence_in_fires_df <- occurrence_in_fires_df %>%
  mutate(LineDist = sqrt(Area_SqM) * 3)

# 4. Create a data frame of angles
angles <- data.frame(Bearing = 1:360)

# 5. The "Cross Join": This multiplies your fires by the 360 angles
# Every fire will now have 360 rows, one for each degree.
radial_table <- occurrence_in_fires_df %>%
  crossing(angles)

# 6. Save as a CSV to bring back into ArcGIS Pro
write.csv(radial_table, "data/processed/Radial_Lines_Input.csv", row.names = FALSE)
```

# Transect Generation

Now in R I will generate transects of 10 points along the lines, with each point separated by 50m. All transects need to be 500m distance apart, meaning there will be fewer transects near the fire origin where the radial lines are closely spaced. Importantly, I want half control transects which do not intersect with treatments (a separate layer), and half treatment transects which include transects entering or exiting a treatment. For treatment transects, half (5) points should be in the transect and 5 outside.

Steps: 
1) Generate points every 50m along each radial line. Create columns with a line ID and distance along line.

## Step 1: Generate points every 50m along each radial line. Create columns with a line ID and distance along line.

In ArcGIS I created radial lines from the ignition points to the fire perimeter using the "Bearing Distance to Line" tools. This generated 360 geodetic lines to the fire perimeter.

I then generated points along the lines, added the Event_ID from the radial lines, and did a spatial join with the fires layer using an attribute filter with Event_ID. This gives me points within the fire perimeter spaced every 50m (planar) from the ignition point.


#Generate Points along lines 


## Identify which transects (origins) intersect treatments (with 5 points in/out)

```{r}
#Collapse treatment dataset
treatments <- st_make_valid(treatments)
flat <- st_union(treatments)

#load points
points <- st_read("data/processed/AllPoints/AllPoints.shp")

#label treated points
spiltmilk1 <- st_intersects(points, flat, sparse = FALSE)
points$is_treated <- rowSums(spiltmilk1) > 0

#identify the treatment progression transects
valid_origins <- points %>%
  group_by(Incid_Name, Bearing) %>% 
  arrange(ORIG_LEN, .by_group = TRUE) %>%
  mutate(
    # Count how many of the NEXT 5 points are treated
    sum_next_5 = roll_sum(as.numeric(is_treated), n = 5, align = "left", fill = NA),
    # Get the sum of the 5 points AFTER those (points 6-10)
    sum_after_5 = lead(sum_next_5, 5)
  ) %>%
  filter(
    (sum_next_5 == 5 & sum_after_5 == 0) | # 5 IN then 5 OUT
    (sum_next_5 == 0 & sum_after_5 == 5)   # 5 OUT then 5 IN
  ) %>%
  mutate(type = ifelse(sum_next_5 == 5, "exiting", "entering")) %>%
  ungroup()

SumStatsB4 <- valid_origins %>%
  group_by(Incid_Name) %>%
  summarize(n = n(),
            entering = sum(type == "entering"),
            exiting = sum(type == "exiting"))

colnames(points)
colnames(valid_origins)
```

## Add points back to identified origins to create transects

```{r}
# 1) Order points and index within each radial
pts <- points %>%
  arrange(Event_ID, ORIG_FID, ORIG_SEQ) %>%
  group_by(Event_ID, ORIG_FID) %>%
  mutate(row = row_number()) %>%
  ungroup()

valid_origins <- st_drop_geometry(valid_origins)
# 2) Get the origin row index for each valid origin
origins <- pts %>%
  semi_join(valid_origins,
            by = c("Event_ID", "ORIG_FID", "ORIG_SEQ")) %>%
  select(Event_ID, ORIG_FID, origin_row = row) %>%
  st_drop_geometry()

# 3) Keep origin + next 9 points (10 total)
transects <- pts %>%
  inner_join(origins, by = c("Event_ID", "ORIG_FID")) %>%
  filter(row >= origin_row & row <= origin_row + 9) %>%
  group_by(Event_ID, ORIG_FID, origin_row) %>%
  filter(
    n() == 10,                                # must have 10 points
    (max(ORIG_SEQ) - min(ORIG_SEQ)) == 9,     # span must be 9
    n_distinct(ORIG_SEQ) == 10                # no missing sequence numbers
  ) %>%
  ungroup()

st_write(transects, "data/processed/TreatmentTransects/TreatmentTransects.shp", delete_dsn = TRUE)
```

## Make transects into line objects and enforce spacing

```{r}
transects <- transects %>%
  group_by(Event_ID, ORIG_FID, origin_row) %>%
  mutate(TransectID = cur_group_id()) %>%
  ungroup()

transect_lines <- transects %>%
  arrange(Event_ID, ORIG_FID, TransectID, ORIG_SEQ) %>%  # ensure order!
  group_by(TransectID, Event_ID, ORIG_FID) %>%
  summarize(geometry = st_cast(st_combine(geometry), "LINESTRING"),
            .groups = "drop")

keep_indices <- c()
remaining_indices <- 1:nrow(transect_lines)

while (length(remaining_indices) > 0) {
  # Pick the first (highest priority) line remaining
  current_idx <- remaining_indices[1]
  keep_indices <- c(keep_indices, current_idx)
  
  # Find all lines within 500m of this line
  # st_is_within_distance returns a list; we check which lines intersect the buffer
  too_close <- st_is_within_distance(transect_lines[current_idx, ], 
                                    transect_lines[remaining_indices, ], 
                                    dist = 500)[[1]]
  
  # Remove the current line and all "too close" lines from the pool
  remaining_indices <- remaining_indices[-too_close]
}

# 4. Filter the original dataset
thinned_transects <- transect_lines[keep_indices, ]

st_write(thinned_transects, "data/processed/TreatmentTransects/FinalTreatmentTransects.shp", delete_dsn = TRUE)

SumStatsAfter <- thinned_transects %>%
  group_by(Event_ID) %>%
  summarize(n = n())
```

```{r}

st_write(valid_origins, "data/processed/TransectOrigins/TransectOrigins.shp", delete_dsn = TRUE)
st_write(filter(points_flagged, is_start_of_group == TRUE), "data/processed/TransectOrigins/TreatmentOrigins.shp", delete_dsn = TRUE)
# 2. Convert those points into 10-point LineStrings (Transects)
final_transect_lines <- final_points_expanded %>%
  group_by(FID_Radial, ORIG_SEQ.start) %>%
  arrange(ORIG_SEQ) %>%
  summarize(do_union = FALSE, .groups = "drop") %>% 
  st_cast("LINESTRING")

# 3. Export to ArcGIS
st_write(final_transect_lines, "data/processed/Fire_Transects_Lines.shp", delete_dsn = TRUE)
```


