---
title: "3b--Joining-Covariates"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: pdf_document
---

This document will join covariates from various data sources with the selected multi-scale plots and fine-scale progression transects. Covariates are in several groups:

Pre-Processing - Day of Burning

Topographic - LANDFIRE - Aspect, Elevation, and Slope

Weather & Climate - gridMET - maximum temperature, precipitation, wind velocity, vpd, pdsi, erc

Pre-Treatment Fuels - LANDFIRE ESP, NLCD Canopy Cover

Treatment Characteristics - Forest Tracker

Response Variable - CBI+


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(terra)
```

# Setup
```{r}
target_crs <- "EPSG:26913"

CO <- st_read("data/raw/States/cb_2018_us_state_20m.shp") %>%
  filter(STUSPS == "CO")

plots100 <- st_read("data/processed/fuels_analysis.gpkg",
  layer = "plots_100m")
plots250 <- st_read("data/processed/fuels_analysis.gpkg",
  layer = "plots_250m")
plots500 <- st_read("data/processed/fuels_analysis.gpkg",
  layer = "plots_500m")
allplots <- rbind(plots100, plots250, plots500)

transects <- st_read("data/processed/Transects/FinalTransectsALL.shp") %>%
  filter(Incid_Name != "ALKALI") #removed, too few 4 kriging

# Change this path to a folder in hard drive
Sys.setenv(TMPDIR = "D:/temp")
raster::rasterOptions(tmpdir = "D:/temp") # If using 'raster'
terra::terraOptions(tempdir = "D:/temp")  # If using 'terra'
```

# Day of Burning

We will need to know Day of Burning when we get to our weather variables. 

```{r}
doblist <- list.files(path = "C:/Users/Elizabeth Buhr/OneDrive/Documents/fuels-analysis/data/processed/DOB", pattern = "\\.tif$", full.names = TRUE)
dob_rast <- lapply(doblist, terra::rast)

# Reproject transects to match raster CRS
transects_proj <- st_transform(transects, crs(dob_rast[[1]]))

# Extract fire name/ID from filenames to use as a matching key
fire_names <- tools::file_path_sans_ext(basename(doblist))
names(dob_rast) <- fire_names

# Extract DOB for each transect point from its matching raster
# This loops over each unique fire, extracts from the right raster, then recombines

transects <- transects_proj %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(pts, fire) {
    
    # Find the matching raster (partial match in case names aren't exact)
    rast_match <- dob_rast[[grep(fire, names(dob_rast), value = TRUE)[1]]]
    
    if (is.null(rast_match)) {
      warning(paste("No matching raster for fire:", fire))
      return(pts %>% mutate(dob = NA))
    }
    
    # Extract values
    extracted <- terra::extract(rast_match, vect(pts))
    pts$dob <- extracted[, 2]  # column 1 is ID, column 2 is the raster value
    return(pts)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)

#Reproject transects back to target crs
transects <- st_transform(transects, target_crs)


transects <- transects %>%
  mutate(date = as.Date(paste(year, dob, sep = "-"), format = "%Y-%j"))
```

# Topography

Aspect, elevation, and slope were downloaded from LANDFIRE. Aspect was converted into "northness" and "eastness".

## Load data

```{r}
CO <- CO %>% st_transform("EPSG:5070")

aspect <- rast("data/raw/LANDFIRE/LF2020_Asp_CONUS/Tif/LF2020_Asp_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

#in landfire, flat areas are coded as -1. change to na
aspect[aspect == -1] <- NA

northness <- cos(aspect * pi / 180)
eastness <- sin(aspect * pi / 180)

elevation <- rast("data/raw/LANDFIRE/LF2020_Elev_220_CONUS/Tif/LC20_Elev_220.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

slope <- rast("data/raw/LANDFIRE/LF2020_SlpD_CONUS/Tif/LF2020_SlpD_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
```

## Transects Join

```{r}
north_trans <- extract(northness, transects)
east_trans <- extract(eastness, transects)
elev_trans <- extract(elevation, transects)
slope_trans <- extract(slope, transects)

transects <- transects %>%
  cbind(north_trans, east_trans, elev_trans, slope_trans) %>%
  select(-starts_with("ID")) %>%
  rename("northness" = "LF2020_Asp_CONUS",
         "eastness" = "LF2020_Asp_CONUS.1",
         "elevation" = "LC20_Elev_220",
         "slope" = "LF2020_SlpD_CONUS")
```

## Plots Join

```{r}
north_plots <- extract(northness, allplots, fun = mean)
east_plots <- extract(eastness, allplots, fun = mean)
elev_plots <- extract(elevation, allplots, fun = mean)
slope_plots <- extract(slope, allplots, fun = mean)

allplots <- allplots %>%
  cbind(north_plots, east_plots, elev_plots, slope_plots) %>%
  select(-starts_with("ID")) %>%
  rename("northness" = "LF2020_Asp_CONUS",
         "eastness" = "LF2020_Asp_CONUS.1",
         "elevation" = "LC20_Elev_220",
         "slope" = "LF2020_SlpD_CONUS")
```

# Weather & Climate

Weather variables were obtained from DayMET.

## Load data

```{r}
install.packages("daymetr")
library("daymetr")
```

## Transects Join

```{r}
final_data <- NULL

# Loop by Year
for(yr in unique(transects$year)){
  
  # 1. Subset points for this year
  transects_yr <- transects[transects$year == yr, ]
  
  # 2. Loop through your variables
  for(v in c("TMax", "PR", "VS", "VPD", "ERC")){
    
    # Map folder name to the prefix
    prefix <- switch(v, "TMax"="tmmx", "PR"="pr", "VS"="vs", "VPD"="vpd", "ERC"="erc")
    file_path <- file.path(base_path, v, paste0(prefix, "_", yr, ".nc"))
    
    if(!file.exists(file_path)) {
       warning(paste("File not found:", file_path))
       transects_yr[[v]] <- NA_real_
       next
    }
    
    # Load the raster cube
    r <- rast(file_path)
    
    # 3. Extract all layers for the points in this year
    ext <- terra::extract(r, vect(transects_yr))
    
    # 4. "Pluck" the specific day value
    values_vector <- sapply(1:nrow(transects_yr), function(i) {
      day_val <- transects_yr$dob[i]
      
      # Handle cases where dob is NA
      if(is.na(day_val)) return(NA_real_)
      
      # Determine the column (Day 1 is col 2, Day 2 is col 3, etc.)
      day_col <- day_val + 1
      
      # Safety check for bounds and missing extraction rows
      if(day_col > ncol(ext) || day_col < 2) return(NA_real_)
      
      return(ext[i, day_col])
    })
    
    # Add this variable to our year-subset
    transects_yr[[v]] <- values_vector
  }
  
  # 5. Bind results
  final_data <- bind_rows(final_data, st_as_sf(transects_yr))
}

transects <- final_data
```


```{r}
transects <- transects %>%
  mutate(
    TMax = TMax - 273.15,      # K to C
  )
```

## Plots Join

```{r}

```


# Pre-Fire Fuels

Data is obtained from Landfire. Pre-fire vegetation is represented by the Environmental Site Potential. Pre-fire fuels are represented by Canopy Cover, which I would say is associated with the risk of high-severity crown fire.

## Load data

```{r}
CO <- st_read("data/raw/States/cb_2018_us_state_20m.shp") %>%
  filter(NAME == "Colorado") %>%
  st_transform("EPSG:5070") #match to crs of landfire for cropping

esp <- rast("data/raw/LANDFIRE/US_140_ESP/Tif/us_140esp.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
activeCat(esp) <- "ESP_NAME"

```

## Transects Join

```{r}
# regime_trans <- extract(regime, transects)
esp_trans <- extract(esp, transects)
# evt_trans <- extract(evt, transects)

transects <- transects %>%
  cbind(esp_trans) %>%
  select(-starts_with("ID")) %>%
  rename("esp" = "ESP_NAME")
```

## Plots Join

```{r}

```


# Treatment Characteristics

- Treatment type (Thinning (No Removal), Thinning (With Removal), Broadcast Burning, Previous Wildfire (with prev wildfire severity))
- Years Since Treatment
- Number of Treatments
- Treatment Size
- Distance to treatment edge

For Plots
- Treatment Extent
- Shape Complexity

## Load data

```{r}
cfri <- st_read("data/raw/CO_Forest_Tracker_2024/CO_Forest_Tracker_2024.gpkg", layer = "forest_tracker_v1_2024")


```

## Transects Join

```{r}

```

## Plots Join

```{r}

```


# Response Variable

## Load Data

```{r}

```

## Transects Join

```{r}
cbilist <- list.files(path = "D:/fuels-analysis/data/processed/GEE/FireSeverity/CBI", pattern = "\\.tif$", full.names = TRUE)
cbi_rast <- lapply(cbilist, terra::rast)

# Reproject transects to match raster CRS
transects_proj <- st_transform(transects, crs(cbi_rast[[1]]))

# Extract fire name/ID from filenames to use as a matching key
fire_names <- tools::file_path_sans_ext(basename(cbilist))
names(cbi_rast) <- fire_names

# Extract cbi for each transect point from its matching raster
# This loops over each unique fire, extracts from the right raster, then recombines

transects <- transects_proj %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(pts, fire) {
    
    # Find the matching raster (partial match in case names aren't exact)
    rast_match <- cbi_rast[[grep(fire, names(cbi_rast), value = TRUE)[1]]]
    
    if (is.null(rast_match)) {
      warning(paste("No matching raster for fire:", fire))
      return(pts %>% mutate(cbi = NA))
    }
    
    # Extract values
    extracted <- terra::extract(rast_match, vect(pts))
    pts$cbi <- extracted[, 2]  # column 1 is ID, column 2 is the raster value
    return(pts)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)

#Reproject transects back to target crs
transects <- st_transform(transects, target_crs)
```

## Plots Join

```{r}

```

