---
title: "3b--Joining-Covariates"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: pdf_document
---

This document will join covariates from various data sources with the selected multi-scale plots and fine-scale progression transects. Covariates are in several groups:

Pre-Processing - Day of Burning
Topographic - LANDFIRE - Aspect, Elevation, and Slope
Weather & Climate - gridMET - maximum temperature, precipitation, wind velocity, vpd, pdsi, erc
Pre-Fire Fuels - LANDFIRE
Treatment Characteristics - Forest Tracker
Response Variable - CBI+

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(terra)
```

```{r}
target_crs <- "EPSG:26913"

transects <- st_read("data/processed/Transects/FinalTransectsALL.shp") 

CO <- st_read("data/raw/States/cb_2018_us_state_20m.shp") %>%
  filter(NAME == "Colorado") %>%
  st_transform("EPSG:5070") #match to crs of landfire for cropping
```

# Pre-processing

We will need to know Day of Burning when we get to our weather variables. Adding that here.

```{r}
doblist <- list.files(path = "C:/Users/Elizabeth Buhr/OneDrive/Documents/fuels-analysis/data/processed/DOB", pattern = "\\.tif$", full.names = TRUE)
dob_rast <- lapply(doblist, terra::rast)

crs(dob_rast[[1]])

```

# Topography

Aspect, elevation, and slope were downloaded from LANDFIRE. Aspect was converted into "northness" and "eastness".

## Load data

```{r}
aspect <- rast("data/raw/LANDFIRE/LF2020_Asp_CONUS/Tif/LF2020_Asp_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

#in landfire, flat areas are coded as -1. change to na
aspect[aspect == -1] <- NA

northness <- cos(aspect * pi / 180)
eastness <- sin(aspect * pi / 180)

elevation <- rast("data/raw/LANDFIRE/LF2020_Elev_220_CONUS/Tif/LC20_Elev_220.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

slope <- rast("data/raw/LANDFIRE/LF2020_SlpD_CONUS/Tif/LF2020_SlpD_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
```

## Transects Join

```{r}
north_trans <- extract(northness, transects)
east_trans <- extract(eastness, transects)
elev_trans <- extract(elevation, transects)
slope_trans <- extract(slope, transects)

transects <- transects %>%
  cbind(north_trans, east_trans, elev_trans, slope_trans) %>%
  select(-starts_with("ID")) %>%
  rename("northness" = "LF2020_Asp_CONUS",
         "eastness" = "LF2020_Asp_CONUS.1",
         "elevation" = "LC20_Elev_220",
         "slope" = "LF2020_SlpD_CONUS")
```

## Plots Join

```{r}

```

# Weather & Climate

Most weather variables were obtained from gridMET. FWI (and maybe PDSI?) obtained from X. 

## Load data

```{r}

```

## Transects Join

```{r}

```

## Plots Join

```{r}

```


# Vegetation & Pre-Fire Fuels

Data is obtained from Landfire. Pre-fire vegetation is represented by the Environmental Site Potential. Pre-fire fuels are represented by Canopy Cover, which I would say is associated with the risk of high-severity crown fire.

## Load data

```{r}
regime <- rast("data/raw/LANDFIRE/LF2016_FRG_200_CONUS/LF2016_FRG_200_CONUS_20250814/LF2016_FRG_200_CONUS/Tif/LC16_FRG_200.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

esp <- rast("data/raw/LANDFIRE/US_140_ESP/Tif/us_140esp.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
activeCat(esp) <- "ESP_NAME"

evt <- rast("data/raw/LANDFIRE/LF2024_EVT_CONUS/Tif/LF2024_EVT_CONUS.tif")%>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
```

## Transects Join

```{r}
regime_trans <- extract(regime, transects)
esp_trans <- extract(esp, transects)
evt_trans <- extract(evt, transects)

transects <- transects %>%
  cbind(regime_trans, esp_trans, evt_trans) %>%
  select(-starts_with("ID")) %>%
  rename("regime" = "BPS_NAME",
         "esp" = "ESP_NAME",
         "evt" = "EVT_NAME")

summary(transects)
```

## Plots Join

```{r}

```
