---
title: "3b--Joining-Covariates"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: pdf_document
---

This document will join covariates from various data sources with the selected multi-scale plots and fine-scale progression transects. Covariates are in several groups:

Pre-Processing - Day of Burning

Topographic - LANDFIRE - Aspect, Elevation, and Slope

Weather & Climate - gridMET - maximum temperature, precipitation, wind velocity, vpd, pdsi, erc

Pre-Treatment Fuels - LANDFIRE ESP, NLCD Canopy Cover

Treatment Characteristics - Forest Tracker

Response Variable - CBI+


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(terra)
library(exactextractr)
library(appeears) #Nasa Earthdata download
```

# Setup
```{r}
target_crs <- "EPSG:26913"

CO <- st_read("D:/Spatial_Data/States/cb_2018_us_state_20m.shp") %>%
  filter(STUSPS == "CO")

allplots <- st_read("data/processed/fuels_analysis.gpkg",
  layer = "all_buffers_selected") %>%
  filter(Incid_Name != c("ALKALI", "IRON", "TRAIL SPRINGS"))

transects <- st_read("data/processed/fuels_analysis.gpkg",
  layer = "transects_selected") %>%
  filter(Incid_Name != c("ALKALI", "IRON", "TRAIL SPRINGS")) #removed, too few 4 kriging, not forested, and too few plots


# Change this path to a folder in hard drive
Sys.setenv(TMPDIR = "D:/temp")
raster::rasterOptions(tmpdir = "D:/temp") # If using 'raster'
terra::terraOptions(tempdir = "D:/temp")  # If using 'terra'
```

# Day of Burning

We will need to know Day of Burning when we get to our weather variables. 

```{r}
doblist <- list.files(path = "C:/Users/Elizabeth Buhr/OneDrive/Documents/fuels-analysis/data/processed/DOB", pattern = "\\.tif$", full.names = TRUE)
dob_rast <- lapply(doblist, terra::rast)

# Reproject transects to match raster CRS
transects_proj <- st_transform(transects, crs(dob_rast[[1]]))

# Extract fire name/ID from filenames to use as a matching key
fire_names <- tools::file_path_sans_ext(basename(doblist))
names(dob_rast) <- fire_names

# Extract DOB for each transect point from its matching raster
# This loops over each unique fire, extracts from the right raster, then recombines

transects <- transects_proj %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(pts, fire) {
    
    # Find the matching raster 
    rast_match <- dob_rast[[grep(fire, names(dob_rast), value = TRUE)[1]]]
    
    # Extract values
    extracted <- terra::extract(rast_match, vect(pts))
    pts$dob <- extracted[, 2]  # column 1 is ID, column 2 is the raster value
    return(pts)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)

#Reproject transects back to target crs
transects <- st_transform(transects, target_crs)

transects <- transects %>%
  mutate(date = as.Date(paste(year, dob, sep = "-"), format = "%Y-%j"))

#Plots join
# Reproject plots to match the raster CRS
allplots_proj <- st_transform(allplots, crs(dob_rast[[1]]))

# Define a custom function to find the most common value (majority)
get_majority <- function(x, na.rm = TRUE) {
  if (na.rm) x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Extract DOB for each plot polygon from its matching raster
allplots_processed <- allplots_proj %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(polys, fire) {
    
    # Find the matching raster
    match_idx <- grep(fire, names(dob_rast))
    
    rast_match <- dob_rast[[match_idx[1]]]
    
    # Extract values using our custom majority function
    extracted <- terra::extract(
      x = rast_match, 
      y = vect(polys), 
      fun = get_majority,
      na.rm = TRUE, 
      ID = FALSE
    )
    
    # Assign extracted values to a new dob column
    polys$dob <- extracted[[1]] 
    return(polys)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)

# Reproject plots back to your target CRS
allplots <- st_transform(allplots_processed, target_crs)

# Calculate calendar date from Julian day
allplots <- allplots %>%
  mutate(date = as.Date(paste(year, dob, sep = "-"), format = "%Y-%j"))
```

# Topography

Aspect, elevation, and slope were downloaded from LANDFIRE. Aspect was converted into "northness" and "eastness".

## Load data

```{r}
CO <- CO %>% st_transform("EPSG:5070")

aspect <- rast("D:/Spatial_Data/LANDFIRE/LF2020_Asp_CONUS/Tif/LF2020_Asp_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

#in landfire, flat areas are coded as -1. change to na
aspect[aspect == -1] <- NA

northness <- cos(aspect * pi / 180)
eastness <- sin(aspect * pi / 180)

elevation <- rast("D:/Spatial_Data/LANDFIRE/LF2020_Elev_CONUS/Tif/LF2020_Elev_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)

slope <- rast("D:/Spatial_Data/LANDFIRE/LF2020_SlpD_CONUS/Tif/LF2020_SlpD_CONUS.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
```

## Transects Join

```{r}
north_trans <- extract(northness, transects)
east_trans <- extract(eastness, transects)
elev_trans <- extract(elevation, transects)
slope_trans <- extract(slope, transects)

transects <- transects %>%
  cbind(north_trans, east_trans, elev_trans, slope_trans) %>%
  select(-starts_with("ID")) %>%
  rename("northness" = "LF2020_Asp_CONUS",
         "eastness" = "LF2020_Asp_CONUS.1",
         "elevation" = "LF2020_Elev_CONUS",
         "slope" = "LF2020_SlpD_CONUS")
```

## Plots Join

```{r}
north_plots <- extract(northness, allplots, fun = mean)
east_plots <- extract(eastness, allplots, fun = mean)
elev_plots <- extract(elevation, allplots, fun = mean)
slope_plots <- extract(slope, allplots, fun = mean)

allplots <- allplots %>%
  cbind(north_plots, east_plots, elev_plots, slope_plots) %>%
  select(-starts_with("ID")) %>%
  rename("northness" = "LF2020_Asp_CONUS",
         "eastness" = "LF2020_Asp_CONUS.1",
         "elevation" = "LF2020_Elev_CONUS",
         "slope" = "LF2020_SlpD_CONUS")
```

# Weather & Climate

Weather variables were obtained from GridMet.

## Load data

```{r}
base_path <- "D:/Spatial_Data/GridMet"
```

## Transects Join

```{r}
final_data <- NULL

# Loop by Year
for(yr in unique(transects$year)){
  
  # Subset points for this year
  transects_yr <- transects[transects$year == yr, ]
  
  # Loop through variables
  for(v in c("TMax", "PR", "VS", "VPD", "ERC")){
    
    # Map folder name to the prefix
    prefix <- switch(v, "TMax"="tmmx", "PR"="pr", "VS"="vs", "VPD"="vpd", "ERC"="erc")
    file_path <- file.path(base_path, v, paste0(prefix, "_", yr, ".nc"))
    
    if(!file.exists(file_path)) {
       warning(paste("File not found:", file_path))
       transects_yr[[v]] <- NA_real_
       next
    }
    
    # Load the raster cube
    r <- rast(file_path)
    
    # Extract all layers for the points in this year
    ext <- terra::extract(r, vect(transects_yr))
    
    # Take the specific day value
    values_vector <- sapply(1:nrow(transects_yr), function(i) {
      day_val <- transects_yr$dob[i]
      
      # Handle cases where dob is NA
      if(is.na(day_val)) return(NA_real_)
      
      # Determine the column (Day 1 is col 2, Day 2 is col 3, etc.)
      day_col <- day_val + 1
      
      # Safety check for bounds and missing extraction rows
      if(day_col > ncol(ext) || day_col < 2) return(NA_real_)
      
      return(ext[i, day_col])
    })
    
    # Add this variable to our year-subset
    transects_yr[[v]] <- values_vector
  }
  
  # Bind results
  final_data <- bind_rows(final_data, st_as_sf(transects_yr))
}

transects <- final_data

transects <- transects %>%
  mutate(
    TMax = TMax - 273.15,      # K to C
  )

#Reproject transects back to target crs
transects <- st_transform(transects, target_crs)
```

## Plots Join

```{r}
final_data <- NULL

# Loop by Year
for(yr in unique(allplots$year)){
  
  # Subset points for this year
  allplots_yr <- allplots[allplots$year == yr, ]
  
  # Loop through your variables
  for(v in c("TMax", "PR", "VS", "VPD", "ERC")){
    
    # Map folder name to the prefix
    prefix <- switch(v, "TMax"="tmmx", "PR"="pr", "VS"="vs", "VPD"="vpd", "ERC"="erc")
    file_path <- file.path(base_path, v, paste0(prefix, "_", yr, ".nc"))
    
    if(!file.exists(file_path)) {
       warning(paste("File not found:", file_path))
       allplots_yr[[v]] <- NA_real_
       next
    }
    
    # Load the raster cube
    r <- rast(file_path)
    
    # Extract all layers for the points in this year
    ext <- terra::extract(r, vect(allplots_yr), fun = mean)
    
    # Take the specific day value
    values_vector <- sapply(1:nrow(allplots_yr), function(i) {
      day_val <- allplots_yr$dob[i]
      
      # Handle cases where dob is NA
      if(is.na(day_val)) return(NA_real_)
      
      # Determine the column (Day 1 is col 2, Day 2 is col 3, etc.)
      day_col <- day_val + 1
      
      # Safety check for bounds and missing extraction rows
      if(day_col > ncol(ext) || day_col < 2) return(NA_real_)
      
      return(ext[i, day_col])
    })
    
    # Add this variable to our year-subset
    allplots_yr[[v]] <- values_vector
  }
  
  # 5. Bind results
  final_data <- bind_rows(final_data, st_as_sf(allplots_yr))
}

allplots <- final_data

allplots <- allplots %>%
  mutate(
    TMax = TMax - 273.15,      # K to C
  )

#Reproject transects back to target crs
allplots <- st_transform(allplots, target_crs)
```

## Download Daymet from Nasa Earthdata

```{r}
#login for nasa earth data - already saved username and password
options(keyring_backend = "file")
keyring::keyring_unlock(keyring = "file")

products <- rs_products()
daymet_info <- products[grepl("daymet", products$Product, ignore.case = TRUE),
                         c("Product", "ProductAndVersion", "Description")]
print(daymet_info)                        # confirm which version (V4) you want

daymet_product <- daymet_info$ProductAndVersion[1]   # set explicitly once confirmed

daymet_layers <- rs_layers(daymet_product)
print(daymet_layers$Layer)                # confirm exact tmax/prcp layer strings

layer_tmax <- grep("tmax", daymet_layers$Layer, ignore.case = TRUE, value = TRUE)
layer_tmin <- grep("tmin", daymet_layers$Layer, ignore.case = TRUE, value = TRUE)
layer_prcp <- grep("prcp", daymet_layers$Layer, ignore.case = TRUE, value = TRUE)
```

## Daymet Transects Join

```{r}
transects_wgs <- st_transform(transects, 4326)
transects_wgs <- transects_wgs %>%
  filter(!is.na(dob)) %>%                                   # drop points with no burn date - can't query without one
  mutate(
    burn_date = as.Date(paste0(year, "-01-01")) + (dob - 1),
    unique_id = paste(Event_ID, Transect_ID, Point_ID, sep = "_")
  )

coords <- st_coordinates(transects_wgs)

locs <- data.frame(
  subtask   = transects_wgs$unique_id,
  latitude  = coords[, 2],
  longitude = coords[, 1],
  start     = as.character(transects_wgs$burn_date),
  end       = as.character(transects_wgs$burn_date),
  stringsAsFactors = FALSE
)

task_df <- bind_rows(
  locs %>% mutate(task = "transects_daymet", product = daymet_product, layer = layer_tmax),
  locs %>% mutate(task = "transects_daymet", product = daymet_product, layer = layer_tmin),
  locs %>% mutate(task = "transects_daymet", product = daymet_product, layer = layer_prcp)
)

task_df <- data.frame(
  task      = task_df$task,
  subtask   = task_df$subtask,
  latitude  = task_df$latitude,
  longitude = task_df$longitude,
  start     = task_df$start,
  end       = task_df$end,
  product   = task_df$product,
  layer     = task_df$layer,
  stringsAsFactors = FALSE
)

str(task_df)   # confirm exactly these 8 columns, nothing extra, before proceeding

unique_ids <- unique(task_df$subtask)
length(unique_ids)   # should be ~4633 if 13899 / 3 layers

chunk_size <- 200     # ~1200 rows per chunk at 3 layers/point - adjust down if this still overflows
id_chunks <- split(unique_ids, ceiling(seq_along(unique_ids) / chunk_size))

task_list <- lapply(id_chunks, function(ids) {
  d <- task_df[task_df$subtask %in% ids, ]
  d$task <- paste0(d$task, "_chunk", match(list(ids), id_chunks))  # unique task name per chunk
  rs_build_task(df = d)
})

# rs_request_batch(
#   request_list = task_list,
#   user    = "buhrea0926",
#   path    = "D:/Spatial_Data/Daymet/transects",
#   workers = 2,
#   verbose = TRUE
# )

## Submit tasks on appeaars and save
submitted <- lapply(task_list, function(t) {
  rs_request(
    request  = t,
    user     = "buhrea0926",
    transfer = FALSE,     # stage only - returns almost immediately, doesn't download
    path     = "D:/Spatial_Data/Daymet/transects",
    verbose  = TRUE
  )
})

task_ids <- sapply(submitted, function(r) r$get_task_id())
saveRDS(task_ids, "D:/Spatial_Data/Daymet/transects/task_ids.rds")  # so you can resume later, even in a new session

## Download data 
task_ids <- readRDS("D:/Spatial_Data/Daymet/transects/task_ids.rds")

done_names <- basename(list.dirs("D:/Spatial_Data/Daymet/transects", recursive = FALSE))

for (i in seq_along(task_ids)) {
  chunk_name <- paste0("transects_daymet_chunk", i)
  if (chunk_name %in% done_names) {
    message("Already have chunk ", i, " - skipping")
    next
  }

  tryCatch({
    rs_transfer(
      task_id = task_ids[i],
      user    = "buhrea0926",
      path    = "D:/Spatial_Data/Daymet/transects",
      verbose = TRUE
    )
  }, error = function(e) {
    message("Chunk ", i, " not ready yet or failed: ", conditionMessage(e))
  })
}
## Done with download

status_df <- rs_list_task(user = "buhrea0926")
names(status_df)   # confirm actual column names - likely something like task_id / status

status_df[status_df$task_id %in% task_ids, c("task_id", "status")]




result_files <- list.files(
  "D:/Spatial_Data/Daymet/transects",
  pattern = "-results\\.csv$", recursive = TRUE, full.names = TRUE
)
daymet_pts <- do.call(rbind, lapply(result_files, read.csv))
nrow(daymet_pts) 
names(daymet_pts)     # <-- check actual column names before the join below

daymet_pts <- daymet_pts %>%
  rename(Point_ID = ID) %>%              # adjust to the real ID column
  mutate(date = as.Date(Date),           # adjust to the real date column
         yday = as.integer(format(date, "%j")))

transects <- transects %>%
  left_join(
    daymet_pts %>%
      select(Point_ID, yday,
             TMax_daymet = matches(layer_tmax),
             PR_daymet   = matches(layer_prcp)),
    by = c("Point_ID" = "Point_ID", "dob" = "yday")
  ) %>%
  st_transform(target_crs)
```

## Daymet Plots Join

```{r}
library(terra)
library(exactextractr)

plot_daymet_path <- "D:/Spatial_Data/Daymet_AppEEARS/area"
dir.create(plot_daymet_path, showWarnings = FALSE, recursive = TRUE)

final_data <- NULL

for (yr in unique(allplots$year)) {

  allplots_yr <- allplots[allplots$year == yr, ]

  roi_yr <- allplots_yr %>%
    st_transform(4326) %>%
    st_union() %>%
    st_convex_hull() %>%
    st_buffer(0.01) %>%          # small buffer so edge plots aren't clipped
    st_as_sf()

  for (v in list(list(name = "TMax", layer = layer_tmax),
                 list(name = "PR",   layer = layer_prcp))) {

    area_df <- data.frame(
      task    = paste0("plots_", yr, "_", v$name),
      subtask = paste0(v$name, "_", yr),
      start   = paste0(yr, "-01-01"),
      end     = paste0(yr, "-12-31"),
      product = daymet_product,
      layer   = v$layer
    )

    area_task <- rs_build_task(df = area_df, roi = roi_yr, format = "geotiff")

    rs_request(
      request  = area_task,
      user     = "your_earthdata_username",
      transfer = TRUE,
      path     = plot_daymet_path,
      verbose  = TRUE
    )

    tif_files <- list.files(
      file.path(plot_daymet_path, area_df$task[1]),
      pattern = "\\.tif$", full.names = TRUE
    )
    # confirm these come back sorted by date/day-of-year before trusting band indexing
    r <- rast(tif_files)

    allplots_yr_proj <- st_transform(allplots_yr, crs(r))
    ext <- exact_extract(r, allplots_yr_proj, fun = "mean")

    values_vector <- sapply(1:nrow(allplots_yr), function(i) {
      day_val <- allplots_yr$dob[i]
      if (is.na(day_val) || day_val > ncol(ext) || day_val < 1) return(NA_real_)
      return(ext[i, day_val])
    })

    allplots_yr[[paste0(v$name, "_daymet")]] <- values_vector
  }

  final_data <- bind_rows(final_data, st_as_sf(allplots_yr))
}

allplots <- final_data
allplots <- st_transform(allplots, target_crs)
```


# Treatment Characteristics

- Treatment type (Thinning (No Removal), Thinning (With Removal), Broadcast Burning, Previous Wildfire (with prev wildfire severity))
- Years Since Treatment
- Number of Treatments
- Treatment Size

For Plots
- Treatment Extent

## Load data

```{r}
#save the layer
treats <- st_read("data/processed/fuels_analysis.gpkg", layer = "treatments")

treats <- treats %>%
  mutate(TRT_TYPE = case_when(
    # Label Wildfires
    MGT_TYPE == "Wildfire" ~ "Wildfire",
    
    # Catch NAs in ACTIVITY where MGT_TYPE is Wildfire 
    is.na(ACTIVITY) & MGT_TYPE == "Wildfire" ~ "Wildfire",
    
    # Catch specific removal activities
    ACTIVITY %in% c("Removal", "Pile Burn") ~ "Removal",
    
    ACTIVITY %in% "Broadcast Burn" ~ "Broadcast Burn",
    
    # Everything else becomes thinning
    TRUE ~ "Thinning"
  ))
```

## Transects Join

```{r}
# Ensure the crs match
if (st_crs(transects) != st_crs(treats)) {
  treats <- st_transform(treats, st_crs(transects))
}

# Add a unique identifier to your transects to track individual points
transects <- transects %>% 
  mutate(point_id = row_number())

# Perform the spatial intersection (points x polygons)
# left = TRUE ensures we keep all transect points, even if they have NO treatments
joined_raw <- st_join(transects, treats, join = st_intersects, left = TRUE)

# Clean up the geometry and process the overlaps
# We drop geometry temporarily to make the group/summarize operations fast on RAM
treatment_summary <- joined_raw %>%
  st_drop_geometry() %>%
  
  # Group by our unique point ID
  group_by(point_id) %>%
  
  # Calculate the metrics per point
  mutate(
    # Count how many treatments hit this point (ignores NA if no treatment matched)
    COUNT = sum(!is.na(YEAR_COMP)),
    
    # Replace 0 with 1 for structural consistency if a point hit exactly one treatment, 
    # or keep 0 if it never intersected any treatment polygons
    COUNT = ifelse(COUNT == 0, 0, COUNT),
    
    # Calculate the earliest treatment year
    # If COUNT is 0, min() returns Inf, so we correct it back to NA
    EARLY_YEAR = ifelse(COUNT > 0, min(YEAR_COMP, na.rm = TRUE), NA_real_),
    
    AGE = ifelse(COUNT > 0, (year - YEAR_COMP), NA_real_)
      ) %>%
  
  # Arrange by YEAR_COMP (descending) so the latest year is always on top (Row 1)
  # using NA_real_ sort behavior to keep untreated points accessible
  arrange(desc(YEAR_COMP), .by_group = TRUE) %>%
  
  # Keep only the top row per point (the latest treatment)
  slice(1) %>%
  ungroup() %>%
  
  # Keep only our point key, the count, and the specific variables you requested
  select(point_id, COUNT, YEAR_COMP, EARLY_YEAR, AGE, TRT_TYPE)

# Join the processed attributes back to your original spatial transect object
transects <- transects %>%
  left_join(treatment_summary, by = "point_id") %>%
  select(-point_id) # Drop the temporary ID tracking column
```

## Plots Join

```{r}
# Match alignment CRS
if (st_crs(allplots) != st_crs(treats)) {
  treats <- st_transform(treats, st_crs(allplots))
}

# Add structural tracking columns to the plots base
allplots_base <- allplots %>%
  mutate(
    plot_id = row_number(),
    total_plot_area = as.numeric(st_area(.))
  )

# Add a row identifier to treatments to act as a proxy tracker for individual polygons
treats_prep <- treats %>%
  mutate(treatment_id = row_number())

# Compute Total Treated Extent AND Count of Overlaps
# We use st_intersection with the raw treats layer to find out how many shapes hit each plot area
plots_joined_raw <- st_intersection(allplots_base %>% select(plot_id, total_plot_area), treats_prep) %>%
  mutate(slice_area = as.numeric(st_area(.))) %>%
  st_drop_geometry()

# Calculate total distinct footprints without counting overlaps
treated_footprint <- st_intersection(allplots_base %>% select(plot_id), st_union(treats)) %>%
  mutate(treated_area_no_overlap = as.numeric(st_area(.))) %>%
  st_drop_geometry() %>%
  select(plot_id, treated_area_no_overlap)

# Calculate the actual raw count of treatments that touched each plot boundary
overlap_counts <- plots_joined_raw %>%
  group_by(plot_id) %>%
  summarize(
    # How many unique treatment shapes physically intersected this plot footprint?
    TOTAL_COUNT = n_distinct(treatment_id),
    # What is the minimum treatment year intersecting this plot footprint?
    # Since plots_joined_raw only has rows for actual intersections, na.rm isn't strictly needed,
    # but it keeps things robust.
    EARLY_YEAR = min(YEAR_COMP, na.rm = TRUE),
    .groups = "drop"
  )

# Compute Categorical Proportions and Weighted Means
# Process continuous variables (Area-Weighted Means for YEAR_COMP)
continuous_summary <- plots_joined_raw %>%
  group_by(plot_id) %>%
  summarize(
    # Weight the completion year by the size of the slice it covers inside the plot
    mean_YEAR_COMP = sum(YEAR_COMP * slice_area) / sum(slice_area),
    .groups = "drop"
  )


# Process categorical variables (Proportions of Plot Area)
# Dissolve treatment polygons by type first, so overlapping polygons
# of the same TRT_TYPE don't get double-counted
treats_by_type <- treats_prep %>%
  group_by(TRT_TYPE) %>%
  summarize(.groups = "drop")

# Intersect plots with the dissolved-by-type layer
plots_by_type <- st_intersection(
  allplots_base %>% select(plot_id, total_plot_area),
  treats_by_type
) %>%
  mutate(slice_area = as.numeric(st_area(.))) %>%
  st_drop_geometry()

# Now proportions per type are safe to sum directly (no overlap within a type)
categorical_summary <- plots_by_type %>%
  group_by(plot_id, TRT_TYPE) %>%
  summarize(slice_sum = sum(slice_area), total_plot_area = first(total_plot_area), .groups = "drop") %>%
  mutate(prop = slice_sum / total_plot_area) %>%
  pivot_wider(
    id_cols = plot_id,
    names_from = TRT_TYPE,
    values_from = prop,
    values_fill = 0,
    names_prefix = "prop_"
  )

# Combine everything back onto the original plot spatial data frame
allplots_final <- allplots_base %>%
  left_join(treated_footprint, by = "plot_id") %>%
  left_join(overlap_counts, by = "plot_id") %>%
  left_join(continuous_summary, by = "plot_id") %>%
  left_join(categorical_summary, by = "plot_id") %>%
  mutate(
    # Convert metrics to structural zeros/defaults if the plot never touched a treatment
    prop_extent_treated = coalesce(treated_area_no_overlap / total_plot_area, 0),
    TOTAL_COUNT = coalesce(TOTAL_COUNT, 0),
    # Safety Check: If TOTAL_COUNT is 0, make sure EARLY_YEAR is NA, not Inf
    EARLY_YEAR = ifelse(TOTAL_COUNT > 0, EARLY_YEAR, NA_real_)
  ) %>%
  # Clean up structural calculation columns and replace missing join values with 0
  select(-plot_id, -total_plot_area, -treated_area_no_overlap) %>%
  mutate(across(starts_with("prop_"), ~coalesce(., 0)))

allplots <- allplots_final
```

# Pre-Fire Fuels

Data is obtained from Landfire. Pre-fire vegetation is represented by the Existing Vegetation Type. Pre-fire fuels are represented by Canopy Cover.

## EVT

### Load data

```{r}
CO <- st_read("D:/Spatial_Data/States/cb_2018_us_state_20m.shp") %>%
  filter(NAME == "Colorado") %>%
  st_transform("EPSG:5070") #match to crs of landfire for cropping

evt <- rast("D:/Spatial_Data/LANDFIRE/US_105_EVT_20250916_2001/US_105_EVT_20250916/US_105_EVT/Tif/us_105evt.tif") %>%
  crop(CO) %>%
  mask(CO) %>%
  project(target_crs)
activeCat(evt) <- "SYSTMGRPNA"
```

### Transects Join

```{r}
evt_trans <- extract(evt, transects)

transects <- transects %>%
  cbind(evt_trans) %>%
  select(-starts_with("ID")) %>%
  rename("EVT" = "SYSTMGRPNA")

unique(transects$EVT)

evt_counts <- transects %>%
  st_drop_geometry() %>%       # Drops the spatial points matrix to save RAM
  count(EVT, sort = TRUE)      # Groups by EVT, counts rows, and sorts descending

# View the resulting summary table
print(evt_counts)

# Reclassification lookup: raw SystmGrpName -> simplified EVT_Group
evt_lookup <- tibble::tribble(
  ~EVT,                                                              ~EVT_Group,
  "Ponderosa Pine Forest, Woodland and Savanna",                     "ponderosa",
  "Lodgepole Pine Forest and Woodland",                              "lodgepole",
  "Spruce-Fir Forest and Woodland",                                  "spruce-fir",
  "Douglas-fir-Ponderosa Pine-Lodgepole Pine Forest and Woodland",   "mixed conifer",
  "Aspen-Mixed Conifer Forest and Woodland",                         "mixed conifer",
  "Aspen Forest, Woodland, and Parkland",                            "aspen",
  "Douglas-fir-Grand Fir-White Fir Forest and Woodland",             "mixed conifer",
  "Pinyon-Juniper Woodland",                                         "pinyon-juniper",
  "Western Riparian Woodland and Shrubland",                         "other",
  "Deciduous Shrubland",                                             "other",
  "Big Sagebrush Shrubland and Steppe",                              "other",
  "Limber Pine Woodland",                                            "other",
  "Developed-Upland Evergreen Forest",                               "other",
  "NLCD-Herbaceous Semi-dry",                                        "other",
  "Developed-Upland Deciduous Forest",                               "other",
  "Introduced Perennial Grassland and Forbland",                     "other",
  "Open Water",                                                      "other",
  "Barren",                                                          "other",
  "Agriculture-Pasture and Hay",                                     "other",
  "Alpine Dwarf-Shrubland, Fell-field and Meadow",                   "other",
  "Developed-Upland Mixed Forest",                                   "other",
  "Developed-Upland Shrubland",                                      "other",
  "Developed-Roads",                                                 "other",
  "Mountain Mahogany Woodland and Shrubland",                        "other",
  "Sand Shrubland",                                                  "other",
  "Grassland",                                                       "other",
  "Dry Tundra",                                                      "other",
  "Introduced Riparian Vegetation",                                  "other",
  "Chaparral",
          "other",
      "Desert Scrub",                               "other",
      "Developed-Medium Intensity",                 "other",
      "Developed-Upland Herbaceous",                "other",
      "Introduced Annual Grassland",                "other",
      "Juniper Woodland and Savanna",                "other",
      "Low Sagebrush Shrubland and Steppe",          "other",
      "NLCD-Herbaceous Semi-wet",                    "other",
      "NLCD-Herbaceous Wetlands",                    "other",
      "NLCD-Recently Disturbed Forest",              "other",
      "Salt Desert Scrub",                           "other",
      "Snow-Ice",                                    "other",
      "Sparse Vegetation",                           "other"
)

transects <- transects %>%
  left_join(evt_lookup, by = "EVT")

# Check for any unmatched categories (should be none / NA)
transects %>%
  st_drop_geometry() %>%
  filter(is.na(EVT_Group)) %>%
  distinct(EVT)

# Recount using the simplified grouping
evt_group_counts <- transects %>%
  st_drop_geometry() %>%
  count(EVT_Group, sort = TRUE)

print(evt_group_counts)
```

### Plots Join

```{r}
# Pull the raster's category table and map raw categories to simplified groups
cat_tbl <- cats(evt)[[1]]
names(cat_tbl)

cat_tbl <- cat_tbl %>%
  rename(EVT = SYSTMGRPNA) %>%            
  left_join(evt_lookup, by = "EVT") %>%     # evt_lookup from before
  mutate(EVT_Group = ifelse(is.na(EVT_Group), EVT, EVT_Group))

# Reclassify the raster so cell values = simplified group, not raw SystmGrpNA
evt_group_levels <- sort(unique(cat_tbl$EVT_Group))

rcl_matrix <- cat_tbl %>%
  transmute(from = VALUE, to = match(EVT_Group, evt_group_levels)) %>%
  as.matrix()

evt_group_rast <- classify(evt, rcl_matrix)
levels(evt_group_rast) <- data.frame(id = seq_along(evt_group_levels),
                                      EVT_Group = evt_group_levels)

# Exact-area fraction of each group within each plot polygon
evt_frac <- exact_extract(
  evt_group_rast,
  allplots,
  fun = "frac",
  append_cols = "unique_id",   
  progress = FALSE
)

# Rename frac_<n> -> prop_<group name>
frac_id_map <- tibble(
  col   = paste0("frac_", seq_along(evt_group_levels)),
  group = evt_group_levels
)

names(evt_frac) <- ifelse(
  names(evt_frac) %in% frac_id_map$col,
  paste0("prop_", frac_id_map$group[match(names(evt_frac), frac_id_map$col)]),
  names(evt_frac)
)

# Join proportions back onto your plots
allplots <- allplots %>%
  left_join(evt_frac, by = "unique_id")
```

## Canopy Cover

Add column to plots with the year we want TCC (year prior to earliest treatment, or year prior to fire for controls)

### Load data

```{r}
tcc_base_dir <- "D:/Spatial_Data/NLCD"

# Search recursively to find all .tif files within the subfolders
all_tcc_files <- list.files(
  path = tcc_base_dir, 
  pattern = "\\.tif$", 
  recursive = TRUE, 
  full.names = TRUE
)

# Extract the 4-digit year from the filename using a regular expression
tcc_years <- str_extract(basename(all_tcc_files), "(19|20)\\d{2}(?=0101)")

# Build a clean named vector where the names are the years and values are paths
tcc_file_map <- all_tcc_files
names(tcc_file_map) <- tcc_years

# Quick check: print years found
print(names(tcc_file_map))
```

### Transects Join

```{r}
# 1. Calculate the target year per transect row
transects_prep <- transects %>%
  mutate(
    # Handle both your point datasets (use YEAR_COMP if it exists, otherwise default to fire year)
    base_year = if_all(any_of("EARLY_YEAR"), ~ !is.na(.)) %>% 
                ifelse(coalesce(EARLY_YEAR, year), year),
                
    # If treated, use YEAR_COMP - 1 (or your explicit EARLY_YEAR - 1 variable). Otherwise fire year - 1.
    target_tcc_year = as.character(base_year - 1),
    row_id = row_number()
  )

# Match CRS to the first available TCC raster
first_tcc_raster <- rast(tcc_file_map[1])
transects_proj <- st_transform(transects_prep, crs(first_tcc_raster))

# 2. Extract values by looping over each unique target year group
transects_with_tcc <- transects_proj %>%
  split(.$target_tcc_year) %>%
  imap_dfr(function(pts, yr) {
    
    # Check if we actually have a raster file for that year
    if (!yr %in% names(tcc_file_map)) {
      warning(paste("No NLCD TCC raster available for year:", yr))
      pts$TCC <- NA_real_
      return(pts)
    }
    
    # Load annual raster pointer
    r_tcc <- rast(tcc_file_map[[yr]])
    
    # Extract points values
    extracted <- terra::extract(r_tcc, vect(pts), ID = FALSE)
    pts$TCC <- extracted[[1]]
    return(pts)
  }) %>%
  arrange(row_id) %>%
  select(-row_id, -base_year, -target_tcc_year)

# Return to your working target CRS
transects <- st_transform(transects_with_tcc, target_crs)
```

### Plots Join

```{r}
# 1. Calculate target year for plots
allplots_prep <- allplots %>%
  mutate(
    # If TOTAL_COUNT > 0, we assume treated. Use mean_YEAR_COMP or your target year calculation.
    base_year = if_all(any_of("EARLY_YEAR"), ~ !is.na(.)) %>% 
                ifelse(coalesce(as.integer(EARLY_YEAR), year), year),
    target_tcc_year = as.character(base_year - 1),
    row_id = row_number()
  )

# Match CRS
allplots_proj <- st_transform(allplots_prep, crs(first_tcc_raster))

# 2. Extract polygon spatial means grouped by year
allplots_with_tcc <- allplots_proj %>%
  split(.$target_tcc_year) %>%
  imap_dfr(function(polys, yr) {
    
    if (!yr %in% names(tcc_file_map)) {
      warning(paste("No NLCD TCC raster available for year:", yr))
      polys$TCC <- NA_real_
      return(polys)
    }
    
    r_tcc <- rast(tcc_file_map[[yr]])
    
    # Extract spatial average across each plot polygon boundary
    extracted <- terra::extract(r_tcc, vect(polys), fun = mean, na.rm = TRUE, ID = FALSE)
    polys$TCC <- extracted[[1]]
    return(polys)
  }) %>%
  arrange(row_id) %>%
  select(-row_id, -base_year, -target_tcc_year)

# Return to target CRS
allplots <- st_transform(allplots_with_tcc, target_crs)
```


# Response Variable

## Load Data

```{r}
cbilist <- list.files(path = "data/processed/GEE/FireSeverity/CBI", pattern = "\\.tif$", full.names = TRUE)
cbi_rast <- lapply(cbilist, terra::rast)

rdnbrlist <- list.files(path = "data/processed/GEE/FireSeverity/RdNBR", pattern = "\\.tif$", full.names = TRUE)
rdnbr_rast <- lapply(rdnbrlist, terra::rast)
```

## Transects Join

```{r}
# Reproject transects to match raster CRS
transects_proj <- st_transform(transects, crs(cbi_rast[[1]]))

# Extract fire name/ID from filenames to use as a matching key
fire_names <- tools::file_path_sans_ext(basename(cbilist))
names(cbi_rast) <- fire_names

# Extract cbi for each transect point from its matching raster
# This loops over each unique fire, extracts from the right raster, then recombines

transects <- transects_proj %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(pts, fire) {
    
    # Find the matching raster (partial match in case names aren't exact)
    rast_match <- cbi_rast[[grep(fire, names(cbi_rast), value = TRUE)[1]]]
    
    if (is.null(rast_match)) {
      warning(paste("No matching raster for fire:", fire))
      return(pts %>% mutate(cbi = NA))
    }
    
    # Extract values
    extracted <- terra::extract(rast_match, vect(pts))
    pts$cbi <- extracted[, 2]  # column 1 is ID, column 2 is the raster value
    return(pts)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)


## Join RdNBR 

# Extract fire name/ID from filenames to use as a matching key
fire_names <- tools::file_path_sans_ext(basename(rdnbrlist))
names(rdnbr_rast) <- fire_names

# Extract RdNBR for each transect point from its matching raster
# This loops over each unique fire, extracts from the right raster, then recombines

transects <- transects %>%
  mutate(row_id = row_number()) %>%  # preserve original order
  split(.$Event_ID) %>%              
  imap_dfr(function(pts, fire) {
    
    # Find the matching raster (partial match in case names aren't exact)
    rast_match <- rdnbr_rast[[grep(fire, names(rdnbr_rast), value = TRUE)[1]]]
    
    if (is.null(rast_match)) {
      warning(paste("No matching raster for fire:", fire))
      return(pts %>% mutate(rdnbr = NA))
    }
    
    # Extract values
    extracted <- terra::extract(rast_match, vect(pts))
    pts$rdnbr <- extracted[, 2]  # column 1 is ID, column 2 is the raster value
    return(pts)
    
  }) %>%
  arrange(row_id) %>%   # restore original order
  select(-row_id)

#Reproject transects back to target crs
transects <- st_transform(transects, target_crs)
```

## Plots Join

```{r}
# 1. List all CBI files and name the list using the Fire IDs from the filenames
cbi_dir <- "data/processed/GEE/FireSeverity/CBI"
cbi_list <- list.files(path = cbi_dir, pattern = "\\.tif$", full.names = TRUE)
cbi_names <- tools::file_path_sans_ext(basename(cbi_list))
names(cbi_list) <- cbi_names

# 2. Match coordinate reference systems using the first available raster
if (length(cbi_list) > 0) {
  first_raster <- rast(cbi_list[[1]])
  allplots_proj <- st_transform(allplots, crs(first_raster))
} else {
  stop("No .tif files found in the specified CBI directory.")
}

# 3. Loop through each unique fire, extract pixel values, and calculate thresholds
allplots_with_cbi <- allplots_proj %>%
  mutate(row_id = row_number()) %>%  # Track original row order
  split(.$Event_ID) %>%              # Split into a list of datasets by Fire ID
  imap_dfr(function(polys, fire) {
    
    # Locate the matching raster file for this fire
    match_idx <- grep(fire, names(cbi_list))
    
    if (length(match_idx) == 0 || is.na(match_idx[1])) {
      warning(paste("No matching CBI raster found for fire ID:", fire))
      polys$mean_cbi <- NA_real_
      polys$prop_high_sev <- NA_real_
      polys$prop_mod_high_sev <- NA_real_
      return(polys)
    }
    
    # Load the specific pointer for this fire's raster
    raster_match <- rast(cbi_list[[match_idx[1]]])
    
    # Convert polygons to SpatVector for terra
    v_polys <- vect(polys)
    
    # Extract ALL underlying pixels per polygon circle. 
    # Setting ID = TRUE allows us to keep track of which pixels belong to which plot.
    extracted_pixels <- terra::extract(raster_match, v_polys, ID = TRUE)
    
    # Identify the name of the data column (it matches the raster file name)
    cbi_col_name <- names(extracted_pixels)[2]
    
    # Calculate the summaries for each plot in this fire subset
    cbi_summary <- extracted_pixels %>%
      filter(!is.na(.data[[cbi_col_name]])) %>%
      group_by(ID) %>%
      summarize(
        # 1. Continuous Mean
        mean_cbi = mean(.data[[cbi_col_name]]),
        
        # 2. Proportion of High Severity (CBI > 2.25)
        prop_high_sev = sum(.data[[cbi_col_name]] > 2.25) / n(),
        
        # 3. Proportion of Moderate to High Severity (CBI >= 1.26 & CBI <= 3.0)
        prop_mod_high_sev = sum(.data[[cbi_col_name]] >= 1.26 & .data[[cbi_col_name]] <= 3.0) / n(),
        .groups = "drop"
      )
    
    # Bind the summarized metrics back onto the spatial plot rows
    # The 'ID' column generated by terra perfectly matches the row index of v_polys
    polys_summarized <- polys %>%
      mutate(ID = row_number()) %>%
      left_join(cbi_summary, by = "ID") %>%
      select(-ID)
    
    return(polys_summarized)
    
  }) %>%
  arrange(row_id) %>%   # Reorder rows back to the exact input state
  select(-row_id)

## Join rdnbr
# 1. List all rdnbr files and name the list using the Fire IDs from the filenames
rdnbr_dir <- "data/processed/GEE/FireSeverity/rdnbr"
rdnbr_list <- list.files(path = rdnbr_dir, pattern = "\\.tif$", full.names = TRUE)
rdnbr_names <- tools::file_path_sans_ext(basename(rdnbr_list))
names(rdnbr_list) <- rdnbr_names

# 2. Match coordinate reference systems using the first available raster
if (length(rdnbr_list) > 0) {
  first_raster <- rast(rdnbr_list[[1]])
  allplots_proj <- st_transform(allplots, crs(first_raster))
} else {
  stop("No .tif files found in the specified rdnbr directory.")
}

# 3. Loop through each unique fire, extract pixel values, and calculate thresholds
allplots_with_rdnbr <- allplots_with_cbi %>%
  mutate(row_id = row_number()) %>%  # Track original row order
  split(.$Event_ID) %>%              # Split into a list of datasets by Fire ID
  imap_dfr(function(polys, fire) {
    
    # Locate the matching raster file for this fire
    match_idx <- grep(fire, names(rdnbr_list))
    
    if (length(match_idx) == 0 || is.na(match_idx[1])) {
      warning(paste("No matching rdnbr raster found for fire ID:", fire))
      polys$mean_rdnbr <- NA_real_
      return(polys)
    }
    
    # Load the specific pointer for this fire's raster
    raster_match <- rast(rdnbr_list[[match_idx[1]]])
    
    # Convert polygons to SpatVector for terra
    v_polys <- vect(polys)
    
    # Extract ALL underlying pixels per polygon circle. 
    # Setting ID = TRUE allows us to keep track of which pixels belong to which plot.
    extracted_pixels <- terra::extract(raster_match, v_polys, ID = TRUE)
    
    # Identify the name of the data column (it matches the raster file name)
    rdnbr_col_name <- names(extracted_pixels)[2]
    
    # Calculate the summaries for each plot in this fire subset
    rdnbr_summary <- extracted_pixels %>%
      filter(!is.na(.data[[rdnbr_col_name]])) %>%
      group_by(ID) %>%
      summarize(
        # 1. Continuous Mean
        mean_rdnbr = mean(.data[[rdnbr_col_name]]),
        .groups = "drop"
      )
    
    # Bind the summarized metrics back onto the spatial plot rows
    # The 'ID' column generated by terra perfectly matches the row index of v_polys
    polys_summarized <- polys %>%
      mutate(ID = row_number()) %>%
      left_join(rdnbr_summary, by = "ID") %>%
      select(-ID)
    
    return(polys_summarized)
    
  }) %>%
  arrange(row_id) %>%   # Reorder rows back to the exact input state
  select(-row_id)


# 4. Project your plots back to your target project CRS
allplots <- st_transform(allplots_with_rdnbr, target_crs)
```

# Save Datasets

```{r}
st_write(
  obj = transects,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "transectsDS",
  driver = "GPKG",
  delete_layer= TRUE
)

st_write(
  obj = allplots,
  dsn = "data/processed/fuels_analysis.gpkg",
  layer = "allplotsDS",
  driver = "GPKG",
  delete_layer= TRUE
)
```

