---
title: "6--Regression-Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, tidy = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(car)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)
library(here)

transects <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "transectsDS") 

transects <- filter(transects,!Incid_Name %in% c("TRAIL SPRINGS", "JUNKINS"))
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Rearrangement", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)
```

## Response Variable Distribution

```{r}
hist(transects$cbi, breaks = 30)
hist(transects$rdnbr, breaks = 30)
plot(transects$cbi, transects$rdnbr)

#Going to remove negatives and 0s from rdnbr to play around with it
transects$rdnbr_pos <- ifelse(transects$rdnbr > 0, transects$rdnbr, NA)
hist(transects$rdnbr_pos)
hist(log(transects$rdnbr_pos))

ggplot(transects, aes(x = rdnbr, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()

ggplot(transects, aes(x = cbi, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()
```

## Look at shape of relationship between each predictor and RdNBR

```{r}
ggplot(transects, aes(x = TRT_TYPE, y = rdnbr)) +
  geom_boxplot()

ggplot(transects, aes(x = COUNT, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = AGE, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = elevation, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = elevation, y = cbi)) +
  geom_point()+
  geom_smooth(method="loess")

ggplot(transects, aes(x = TCC, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = slope, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = northness, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")


ggplot(transects, aes(x = Point_ID, y = rdnbr, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(x = "Point ID", y = "rdnbr", color = "Type")

library(tidyverse)

# Check categorical drivers (e.g., Treatment Status or Cover Type)
ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(alpha = 0.5) +
  theme_bw()
```

## Create linear regression model including nested random effects of transect ID and fire ID. Fixed effects include treatment characteristics and an interaction between point distance and transect type

- currently not including age or prev burn sev

```{r}
m3 <- lmer(rdnbr ~ TRT_TYPE + COUNT + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m3)
vif(m3)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m3), main = "Q-Q Plot of Residuals")
qqline(resid(m3), col = "red")

# Residuals vs Fitted
plot(fitted(m3), resid(m3),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

## RdNBR + Dispersion Formula

```{r}
transects <- transects %>%
  arrange(Transect_ID, Point_ID) %>%
  mutate(Point_Factor = factor(Point_ID))

m3 <- glmmTMB(rdnbr ~ TRT_TYPE + EVT_Group + Point_ID*type + (1|Incid_Name/Transect_ID), data = transects, dispformula = ~Incid_Name, REML = TRUE)

summary(m3)

library(DHARMa)
sim <- simulateResiduals(m3) 
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
plotResiduals(sim, form = transects$Incid_Name)
testQuantiles(sim)
```

## Try a model with only positive rdnbr

```{r}
m4 <- lmer(rdnbr_pos ~ TRT_TYPE + COUNT + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m4)
vif(m4)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m4), main = "Q-Q Plot of Residuals")
qqline(resid(m4), col = "red")

# Residuals vs Fitted
plot(fitted(m4), resid(m4),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Try a Gamma family

Used to model a response distribution that is positive, right skewed, and continuous

```{r}
library(glmmTMB)

m4_gamma <- glmmTMB(
  rdnbr_pos ~ TRT_TYPE + COUNT + Point_ID * type + (1 | Event_ID/Transect_ID),
  data = transects,
  family = Gamma(link = "log")
)

simulateResiduals(m4_gamma) |> plot()
```

# Create a model with only Cameron Peak

```{r}
df_cam <- filter(transects, Incid_Name == "CAMERON PEAK")

m5 <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = df_cam, REML = TRUE)

summary(m5)
vif(m5)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m5), main = "Q-Q Plot of Residuals")
qqline(resid(m5), col = "red")

# Residuals vs Fitted
plot(fitted(m5), resid(m5),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Cameron Peak + Dispersion argument

```{r}
m6 <- glmmTMB(rdnbr ~ TRT_TYPE + EVT_Group + Point_ID*type + (1|Transect_ID), data = df_cam, dispformula = ~TRT_TYPE, REML = TRUE)

summary(m6)

library(DHARMa)
sim <- simulateResiduals(m6)
plot(sim)
plotResiduals(sim, form = df_cam$TRT_TYPE)
testQuantiles(sim)
```

## Zero-Inflated Beta Regression

Transform CBI to be with 0-1 exclusive boundaries using Smithson & Verkuilen (2006) transformation. Using a zero inflated formula because 0s account for about 6% of my data while 3s are negligible (0.23%)
```{r}
transects <- transects %>%
  mutate(cbi_01 = cbi / 3,
         cbi_squeezed = case_when(
           cbi_01 == 1 ~ 1 - (0.5 / n()),   # nudge exact 3s just below 1
           TRUE ~ cbi_01                      # leave 0s and interior values as-is
         ))

m7 <- glmmTMB(cbi_squeezed ~ TRT_TYPE + Point_ID*type + (1|Incid_Name/Transect_ID),
                  data = transects,
                  family = beta_family(),
                  ziformula = ~ Incid_Name,
                  REML = TRUE)

sim <- simulateResiduals(m7)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)
```

# Gemini - Zero-Inflated Beta and Ordered Beta via BRMS

```{r}
# Option A: Zero-Inflated Beta (requires squeezing ONLY if values equal exactly 1)
# Smithson & Verkuilen squeeze on non-zero values, OR handle 0s natively in ziformula:

transects <- transects %>%
  mutate(
    # Standard 0-1 scaling for CBI (range 0 to 3)
    cbi_prop = cbi / 3,
    # Squeeze ONLY values == 1 down to 0.9999 so beta family can evaluate the interior
    cbi_zi = ifelse(cbi_prop == 1, 0.9999, cbi_prop)
  )

m7_zi <- glmmTMB(
  cbi_zi ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  family = beta_family(link = "logit"), # Zero-inflated beta
  ziformula = ~ 1 # Models probability of exact 0 (unburned)
)

sim <- simulateResiduals(m7_zi)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)

# Option B: Ordered Beta via brms (Handles exact 0s AND 3s natively without squeezing!)
library(ordbetareg)
library(brms)
m7_ordbeta <- ordbetareg(
  cbi_prop ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  chains = 4, cores = 4
)
```

