---
title: "6--Regression-Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(glmmTMB)
library(DHARMa)
library(ggeffects)

transects <- st_read("data/processed/fuels_analysis.gpkg", layer = "transectsDS") 
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Thinning", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)
```

## Response Variable Distribution

```{r}
hist(transects$cbi)
```

## Look at shape of relationship between each predictor and cbi

```{r}
transects_test <- filter(transects, Incid_Name != "CAMERON PEAK")
ggplot(transects, aes(x = TRT_TYPE, y = cbi)) +
  geom_boxplot()
ggplot(transects_test, aes(x = TRT_TYPE, y = cbi)) +
  geom_boxplot()

ggplot(transects, aes(x = COUNT, y = cbi)) +
  geom_point()
ggplot(transects_test, aes(x = COUNT, y = cbi)) +
  geom_point()

ggplot(transects, aes(x = AGE, y = cbi)) +
  geom_point()
ggplot(transects_test, aes(x = AGE, y = cbi)) +
  geom_point()


ggplot(transects, aes(x = Point_ID, y = cbi, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(x = "Point ID", y = "CBI", color = "Type")
ggplot(transects_test, aes(x = Point_ID, y = cbi, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(x = "Point ID", y = "CBI", color = "Type")
```

## Create linear regression model including nested random effects of transect ID and fire ID. Fixed effects include treatment characteristics (type, age, number, and prev. burn severity), and an interaction between point distance and transect type

- prev burn sev

```{r}
m1 <- lmer(cbi ~ TRT_TYPE + COUNT + AGE:TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m1)
vif(m1)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m1), main = "Q-Q Plot of Residuals")
qqline(resid(m1), col = "red")

# Residuals vs Fitted
plot(fitted(m1), resid(m1),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Beta Regression

Trying a beta regression because the diamond shape in residuals is likely because we are treating bounded data as continuous (cite).

```{r}
transects$cbi_scaled <- transects$cbi/3

#Adjust to get rid of exact 0s and 1s
n <- nrow(transects)
transects$cbi_squeezed <- (transects$cbi_scaled * (n - 1) + 0.5) / n

m2 <- glmmTMB(cbi_squeezed ~ TRT_TYPE + COUNT + AGE + Point_ID*type + (1|Event_ID/Transect_ID), family = beta_family(), data = transects)

sim_beta <- simulateResiduals(m2)
plot(sim_beta)
```


```{r}
# Check residual patterns against key predictors directly
plotResiduals(sim_beta, form = transects$TRT_TYPE)
plotResiduals(sim_beta, form = transects$type)
plotResiduals(sim_beta, form = transects$COUNT)
plotResiduals(sim_beta, form = transects$AGE)
```

# Add Age interaction

```{r}
m3 <- glmmTMB(cbi_squeezed ~ TRT_TYPE + COUNT + AGE:treated + Point_ID*type + (1|Event_ID/Transect_ID), family = beta_family(), data = transects)

summary(m3)

sim3 <- simulateResiduals(m3)
plot(sim3)

# Re-check the AGE-specific residual pattern that flagged the problem originally
plotResiduals(sim3, form = transects$AGE)
```

# Add a dispersion formula 

Addresses uneven variance between treatment type groups

```{r}
m4 <- glmmTMB(
  cbi_squeezed ~ TRT_TYPE + COUNT + AGE:treated + Point_ID*type + 
    (1 | Event_ID/Transect_ID),
  family = beta_family(),
  dispformula = ~ TRT_TYPE,
  data = transects
)

summary(m4)
sim4 <- simulateResiduals(m4)
plot(sim4)
```

