---
title: "6--Regression-Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, tidy = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(car)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)
library(here)

transects <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "transectsDS") 

transects <- filter(transects,!Incid_Name %in% c("TRAIL SPRINGS", "JUNKINS"))
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Rearrangement", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)

#Sort start of transects into L/M/H based on average severity of first 5 points
transects <- transects %>%
  group_by(Transect_ID) %>% 
mutate(
    start_value = mean(rdnbr[Point_ID %in% 1:5], na.rm = TRUE),
    START_SEV = case_when(
      start_value < 300              ~ "Low",
      start_value >= 300 & start_value <= 500 ~ "Moderate",
      start_value > 500              ~ "High",
    )
  ) %>%
  ungroup() %>%
  select(-start_value)
```

## Response Variable Distribution

```{r}
hist(transects$cbi, breaks = 30)
hist(transects$rdnbr, breaks = 30)
plot(transects$cbi, transects$rdnbr)


#Going to remove negatives and 0s from rdnbr to play around with it
transects$rdnbr_pos <- ifelse(transects$rdnbr > 0, transects$rdnbr, NA)
hist(transects$rdnbr_pos)
hist(log(transects$rdnbr_pos))

ggplot(transects, aes(x = rdnbr, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()

ggplot(transects, aes(x = cbi, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()

# View distribution of response faceted by starting severity
ggplot(transects, aes(x = rdnbr, fill = START_SEV)) +
  geom_density(alpha = 0.5) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.5) + #add after_stat to scale each group based on sample size (area under curve = 1 across all groups, not for each individual group)
  theme_bw()

#Facetted by EVT for easier viewing
ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  facet_wrap(~EVT_Group, ncol = 2) +
  theme_bw() +
  theme(legend.position = "none")  
```

## Look at shape of relationship between each predictor and RdNBR

```{r}
ggplot(transects, aes(x = TRT_TYPE, y = rdnbr)) +
  geom_boxplot()

ggplot(transects, aes(x = COUNT, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = AGE, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = elevation, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = elevation, y = cbi)) +
  geom_point()+
  geom_smooth(method="loess")

ggplot(transects, aes(x = TCC, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = slope, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = northness, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")
```

## Create linear regression model including nested random effects of transect ID and fire ID. Fixed effects include treatment characteristics and an interaction between point distance and transect type

- currently not including age or prev burn sev

```{r}
m3 <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m3)
vif(m3)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m3), main = "Q-Q Plot of Residuals")
qqline(resid(m3), col = "red")

# Residuals vs Fitted
plot(fitted(m3), resid(m3),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

## RdNBR + Dispersion Formula

```{r}
m3 <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Incid_Name/Transect_ID), data = transects, dispformula = ~Incid_Name, REML = TRUE)

summary(m3)

library(DHARMa)
sim <- simulateResiduals(m3) 
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
plotResiduals(sim, form = transects$Incid_Name)
testQuantiles(sim)
```

# Create a model with only Cameron Peak

```{r}
df_cam <- filter(transects, Incid_Name == "CAMERON PEAK")

m5 <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = df_cam, REML = TRUE)

summary(m5)
vif(m5)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m5), main = "Q-Q Plot of Residuals")
qqline(resid(m5), col = "red")

# Residuals vs Fitted
plot(fitted(m5), resid(m5),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Cameron Peak + Dispersion argument

```{r}
m6 <- glmmTMB(rdnbr ~ TRT_TYPE + EVT_Group + Point_ID*type + (1|Transect_ID), data = df_cam, dispformula = ~TRT_TYPE, REML = TRUE)

summary(m6)

library(DHARMa)
sim <- simulateResiduals(m6)
plot(sim)
plotResiduals(sim, form = df_cam$TRT_TYPE)
testQuantiles(sim)
```

## Zero-Inflated Beta Regression

Transform CBI to be with 0-1 exclusive boundaries using Smithson & Verkuilen (2006) transformation. Using a zero inflated formula because 0s account for about 6% of my data while 3s are negligible (0.23%)
```{r}
transects <- transects %>%
  mutate(cbi_01 = cbi / 3,
         cbi_squeezed = case_when(
           cbi_01 == 1 ~ 1 - (0.5 / n()),   # nudge exact 3s just below 1
           TRUE ~ cbi_01                      # leave 0s and interior values as-is
         ))

m7 <- glmmTMB(cbi_squeezed ~ TRT_TYPE + Point_ID*type + (1|Incid_Name/Transect_ID),
                  data = transects,
                  family = beta_family(),
                  ziformula = ~ Incid_Name,
                  REML = TRUE)

sim <- simulateResiduals(m7)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)
```

# Zero-Inflated Beta and Ordered Beta via BRMS

```{r}
# Option A: Zero-Inflated Beta (requires squeezing ONLY if values equal exactly 1)
# Smithson & Verkuilen squeeze on non-zero values, OR handle 0s natively in ziformula:

transects <- transects %>%
  mutate(
    # Standard 0-1 scaling for CBI (range 0 to 3)
    cbi_prop = cbi / 3,
    # Squeeze ONLY values == 1 down to 0.9999 so beta family can evaluate the interior
    cbi_zi = ifelse(cbi_prop == 1, 0.9999, cbi_prop)
  )

m7_zi <- glmmTMB(
  cbi_zi ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  family = beta_family(link = "logit"), # Zero-inflated beta
  ziformula = ~ 1 # Models probability of exact 0 (unburned)
)

sim <- simulateResiduals(m7_zi)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)

# Option B: Ordered Beta via brms (Handles exact 0s AND 3s natively without squeezing!)
library(ordbetareg)
library(brms)

m7_ordbeta <- ordbetareg(
  formula = cbi_prop ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  chains = 4, 
  cores = 4,
  iter = 2000
)

library(bayesplot)

# 1A. Density Overlay: Checks overall distribution shape & boundary peaks
pp_check(m7_ordbeta, type = "dens_overlay", ndraws = 50)

# 1B. Histograms: Compare actual data distribution to 8 simulated replicates
pp_check(m7_ordbeta, type = "hist", ndraws = 8)

# 1C. Check performance across groups (e.g., Treatment Types)
pp_check(m7_ordbeta, type = "stat_grouped", group = "TRT_TYPE", stat = "mean")

# Generate simulated response matrix from the posterior draws
preds <- posterior_predict(m7_ordbeta, ndraws = 250)

# Calculate DHARMa residuals
dharma_res <- createDHARMa(
  simulatedResponse = t(preds),
  observedResponse = transects$cbi_prop,
  fittedPredictedResponse = apply(preds, 2, median),
  integerResponse = FALSE
)

# Plot standard DHARMa diagnostic plots
plot(dharma_res)
```

# Ordered beta Regression with Dispersion Formula

```{r}
# 1. Define formula predicting both mean and phi
f_m8 <- bf(
  cbi_prop ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  phi ~ TRT_TYPE
)

# 2. Fit with phi_reg specified so ordbetareg sets the correct phi priors
m8_ordbeta <- ordbetareg(
  formula = f_m8,
  data = transects,
  phi_reg = "both",  # Tells ordbetareg to expect a regression model for phi
  chains = 4, 
  cores = 4,
  iter = 2000,
  control = list(adapt_delta = 0.95)
)

```

# L/M/H Models

```{r}
transects %>% 
  st_drop_geometry() %>% 
  group_by(Incid_Name) %>%
  count(START_SEV) %>% 
  tidyr::pivot_wider(
    names_from = Incid_Name,
    values_from = n,
    values_fill = 0
  ) %>%
  print(n = Inf, width = Inf)

low_df <- transects %>%
  filter(START_SEV == "Low") #940 obs
mod_df <- transects %>%
  filter(START_SEV == "Moderate") #550 obs
high_df <- transects %>%
  filter(START_SEV == "High") #1860 obs

summary(low_df$type)
summary(mod_df$type)
summary(high_df$type)

low_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
mod_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
high_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
```

# Low Starting Severity Model

```{r}
low_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = low_df, REML = TRUE)

summary(low_fit)
vif(low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_fit), resid(low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Moderate Starting Severity Model

```{r}
mod_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mod_df, REML = TRUE)

summary(mod_fit)
vif(mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mod_fit), resid(mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# High Starting Severity Model

```{r}
high_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = high_df, REML = TRUE)

summary(high_fit)
vif(high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(high_fit), resid(high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + L/M/H Models

```{r}
low_enter_df <- low_df %>%
  filter(type == "entering") #940 obs -> 410 obs
mod_enter_df <- mod_df %>%
  filter(type == "entering") #550 obs -> 240 obs
high_enter_df <- high_df %>%
  filter(type == "entering") #1860 obs -> 780 obs
```

# Entering + Low Starting Severity Model

```{r}
low_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = low_enter_df, REML = TRUE)

summary(low_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_ent_fit), resid(low_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + Moderate Starting Severity Model

```{r}
mod_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = mod_enter_df, REML = TRUE)

summary(mod_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mod_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mod_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mod_ent_fit), resid(mod_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering + High Starting Severity Model

```{r}
high_ent_fit <- lmer(rdnbr ~ TRT_TYPE + (1|Event_ID/Transect_ID), data = high_enter_df, REML = TRUE)

summary(high_ent_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(high_ent_fit), main = "Q-Q Plot of Residuals")
qqline(resid(high_ent_fit), col = "red")

# Residuals vs Fitted
plot(fitted(high_ent_fit), resid(high_ent_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Forest Type Models

```{r}
pondo_df <- transects %>%
  filter(EVT_Group == "ponderosa") %>% #857 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

mixed_df <- transects %>%
  filter(EVT_Group == "mixed conifer") %>% #775 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

lodge_df <- transects %>%
  filter(EVT_Group == "lodgepole") %>% #749 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #200 observations
```

# Low Starting Severity Model

```{r}
low_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = low_df, REML = TRUE)

summary(low_fit)
vif(low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_fit), resid(low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

