---
title: "6--Regression-Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, tidy = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(car)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)
library(here)

transects <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "transectsDS") 

transects <- filter(transects,!Incid_Name %in% c("TRAIL SPRINGS", "JUNKINS"))
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Rearrangement", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)

#Sort start of transects into L/M/H based on average severity of first 5 points
transects <- transects %>%
  group_by(Transect_ID) %>% 
mutate(
    start_value = mean(rdnbr[Point_ID %in% 1:5], na.rm = TRUE),
    START_SEV = case_when(
      start_value < 300              ~ "Low",
      start_value >= 300 & start_value <= 500 ~ "Moderate",
      start_value > 500              ~ "High",
    )
  ) %>%
  ungroup() %>%
  select(-start_value)

#Drop geometry
transects <- st_drop_geometry(transects)
```

## Response Variable Distribution

```{r}
hist(transects$cbi, breaks = 30)
hist(transects$rdnbr, breaks = 30)

ggplot(transects, aes(x = rdnbr, groups = Incid_Name)) +
  facet_wrap(~Incid_Name) +
  geom_histogram()

ggplot(transects, aes(x = cbi, groups = Incid_Name)) +
  facet_wrap(~Incid_Name, scales="free_y") +
  geom_histogram()

# View distribution of response faceted by starting severity
ggplot(transects, aes(x = rdnbr, fill = START_SEV)) +
  geom_density(alpha = 0.5) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.5) + #add after_stat to scale each group based on sample size (area under curve = 1 across all groups, not for each individual group)
  theme_bw()

#Faceted by EVT for easier viewing
ggplot(transects, aes(x = rdnbr, fill = EVT_Group)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  facet_wrap(~EVT_Group, ncol = 2) +
  theme_bw() +
  theme(legend.position = "none")

ggplot(transects, aes(x = rdnbr, fill = TRT_TYPE)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = type)) +
  geom_density(aes(y = after_stat(count/sum(count))), alpha = 0.5) +
  theme_bw()

ggplot(transects, aes(x = rdnbr, fill = Incid_Name)) +
  geom_density(aes(y = after_stat(count / sum(count))), alpha = 0.7) +
  facet_wrap(~Incid_Name, ncol = 4) +
  theme_bw() +
  theme(legend.position = "none")

ggplot(transects, aes(x = rdnbr, fill = extreme_area)) +
  geom_density(alpha = 0.7) +
  theme_bw() #not scaled for sample size, getting error
```

## Look at shape of relationship between each predictor and RdNBR

```{r}
ggplot(transects, aes(x = TRT_TYPE, y = rdnbr)) +
  geom_boxplot()

ggplot(transects, aes(x = extreme_area, y = rdnbr)) +
  geom_boxplot()

ggplot(transects, aes(x = EVT_Group, y = rdnbr)) +
  geom_boxplot()

ggplot(transects, aes(x = COUNT, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = AGE, y = rdnbr)) +
  geom_point()

ggplot(transects, aes(x = elevation, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = elevation, y = cbi)) +
  geom_point()+
  geom_smooth(method="loess")

ggplot(transects, aes(x = TCC, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = slope, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")

ggplot(transects, aes(x = northness, y = rdnbr)) +
  geom_point() +
  geom_smooth(method="loess")
```

## Create linear regression model including nested random effects of transect ID and fire ID. Fixed effects include treatment characteristics and an interaction between point distance and transect type

- currently not including age or prev burn sev

```{r}
m3 <- lmer(rdnbr ~ TRT_TYPE + EVT_Group + extreme_area + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m3)
vif(m3)

#Residuals actually look normally distributed
hist(resid(m3), breaks = 40)
# or
plot(density(resid(m3)))

#Check where outliers are clustered
transects$resid_m3 <- resid(m3)
transects[abs(transects$resid_m3) > 1000, c("Incid_Name", "TRT_TYPE", "EVT_Group", "extreme_area", "rdnbr", "resid_m3")]

table(transects$extreme_area, transects$Incid_Name)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m3), main = "Q-Q Plot of Residuals")
qqline(resid(m3), col = "red")

# Residuals vs Fitted
plot(fitted(m3), resid(m3),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

## RdNBR + Dispersion Formula

```{r}
m3 <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Incid_Name/Transect_ID), data = transects, dispformula = ~Incid_Name, REML = TRUE)

summary(m3)

sim <- simulateResiduals(m3) 
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
plotResiduals(sim, form = transects$Incid_Name)
testQuantiles(sim)
testOutliers(sim)
testDispersion(sim)

library(DHARMa)
library(dplyr)
library(ggplot2)

# Extract residuals and attach fire ID (and other grouping vars)
res <- simulateResiduals(m3, plot = FALSE)
resid_df <- transects %>%
  mutate(dharma_resid = res$scaledResiduals)

# 1. Flag the "spike" observations near the upper boundary
resid_df <- resid_df %>%
  mutate(near_one = dharma_resid > 0.9)

# 2. Proportion of near-1 residuals by fire
resid_df %>%
  group_by(Incid_Name) %>%
  summarise(
    n = n(),
    prop_near_one = mean(near_one),
    .groups = "drop"
  ) %>%
  arrange(desc(prop_near_one)) %>%
  print(n = Inf)

# 3. Visual: faceted histograms by fire
ggplot(resid_df, aes(x = dharma_resid)) +
  geom_histogram(bins = 30, boundary = 0) +
  facet_wrap(~ Incid_Name, scales = "free_y") +
  geom_vline(xintercept = 0.9, color = "red", linetype = "dashed") +
  theme_minimal()

# 4. If a few fires dominate the spike, check chi-square style enrichment
tab <- table(resid_df$Incid_Name, resid_df$near_one)
chisq.test(tab)  # will likely be significant with 13 fires + uneven spike
```

# Create a model with only Cameron Peak

```{r}
df_cam <- filter(transects, Incid_Name == "CAMERON PEAK")

m5 <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = df_cam, REML = TRUE)

summary(m5)
vif(m5)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m5), main = "Q-Q Plot of Residuals")
qqline(resid(m5), col = "red")

# Residuals vs Fitted
plot(fitted(m5), resid(m5),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Cameron Peak + Dispersion argument

```{r}
m6 <- glmmTMB(rdnbr ~ TRT_TYPE + EVT_Group + Point_ID*type + (1|Transect_ID), data = df_cam, dispformula = ~TRT_TYPE, REML = TRUE)

summary(m6)

library(DHARMa)
sim <- simulateResiduals(m6)
plot(sim)
plotResiduals(sim, form = df_cam$TRT_TYPE)
testQuantiles(sim)
```

## Zero-Inflated Beta Regression

Transform CBI to be with 0-1 exclusive boundaries using Smithson & Verkuilen (2006) transformation. Using a zero inflated formula because 0s account for about 6% of my data while 3s are negligible (0.23%)
```{r}
transects <- transects %>%
  mutate(cbi_01 = cbi / 3,
         cbi_squeezed = case_when(
           cbi_01 == 1 ~ 1 - (0.5 / n()),   # nudge exact 3s just below 1
           TRUE ~ cbi_01                      # leave 0s and interior values as-is
         ))

m7 <- glmmTMB(cbi_squeezed ~ TRT_TYPE + Point_ID*type + (1|Incid_Name/Transect_ID),
                  data = transects,
                  family = beta_family(),
                  ziformula = ~ Incid_Name,
                  REML = TRUE)

sim <- simulateResiduals(m7)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)
```

# Zero-Inflated Beta and Ordered Beta via BRMS

```{r}
# Option A: Zero-Inflated Beta (requires squeezing ONLY if values equal exactly 1)
# Smithson & Verkuilen squeeze on non-zero values, OR handle 0s natively in ziformula:

transects <- transects %>%
  mutate(
    # Standard 0-1 scaling for CBI (range 0 to 3)
    cbi_prop = cbi / 3,
    # Squeeze ONLY values == 1 down to 0.9999 so beta family can evaluate the interior
    cbi_zi = ifelse(cbi_prop == 1, 0.9999, cbi_prop)
  )

m7_zi <- glmmTMB(
  cbi_zi ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  family = beta_family(link = "logit"), # Zero-inflated beta
  ziformula = ~ 1 # Models probability of exact 0 (unburned)
)

sim <- simulateResiduals(m7_zi)
plot(sim)
plotResiduals(sim, form = transects$TRT_TYPE)
testQuantiles(sim)

# Option B: Ordered Beta via brms (Handles exact 0s AND 3s natively without squeezing!)
library(ordbetareg)
library(brms)

m7_ordbeta <- ordbetareg(
  formula = cbi_prop ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  data = transects,
  chains = 4, 
  cores = 4,
  iter = 2000
)

library(bayesplot)

# 1A. Density Overlay: Checks overall distribution shape & boundary peaks
pp_check(m7_ordbeta, type = "dens_overlay", ndraws = 50)

# 1B. Histograms: Compare actual data distribution to 8 simulated replicates
pp_check(m7_ordbeta, type = "hist", ndraws = 8)

# 1C. Check performance across groups (e.g., Treatment Types)
pp_check(m7_ordbeta, type = "stat_grouped", group = "TRT_TYPE", stat = "mean")

# Generate simulated response matrix from the posterior draws
preds <- posterior_predict(m7_ordbeta, ndraws = 250)

# Calculate DHARMa residuals
dharma_res <- createDHARMa(
  simulatedResponse = t(preds),
  observedResponse = transects$cbi_prop,
  fittedPredictedResponse = apply(preds, 2, median),
  integerResponse = FALSE
)

# Plot standard DHARMa diagnostic plots
plot(dharma_res)
```

# Ordered beta Regression with Dispersion Formula

```{r}
# 1. Define formula predicting both mean and phi
f_m8 <- bf(
  cbi_prop ~ TRT_TYPE + Point_ID * type + (1 | Incid_Name / Transect_ID),
  phi ~ TRT_TYPE
)

# 2. Fit with phi_reg specified so ordbetareg sets the correct phi priors
m8_ordbeta <- ordbetareg(
  formula = f_m8,
  data = transects,
  phi_reg = "both",  # Tells ordbetareg to expect a regression model for phi
  chains = 4, 
  cores = 4,
  iter = 2000,
  control = list(adapt_delta = 0.95)
)

```

# L/M/H Models

```{r}
low_df <- transects %>%
  filter(START_SEV == "Low") #940 obs
mod_df <- transects %>%
  filter(START_SEV == "Moderate") #550 obs
high_df <- transects %>%
  filter(START_SEV == "High") #1860 obs

summary(low_df$type)
summary(mod_df$type)
summary(high_df$type)

low_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
mod_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )
high_df %>% 
  st_drop_geometry() %>% 
  count(EVT_Group, TRT_TYPE) %>% 
  tidyr::pivot_wider(
    names_from = EVT_Group,
    values_from = n,
    values_fill = 0
  )

low_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
mod_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
high_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
```

### Low Starting Severity Model

```{r}
low_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = low_df, REML = TRUE)

summary(low_fit)
vif(low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(low_fit), resid(low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Moderate Starting Severity Model

```{r}
mod_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mod_df, REML = TRUE)

summary(mod_fit)
vif(mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mod_fit), resid(mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### High Starting Severity Model

```{r}
high_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = high_df, REML = TRUE)

summary(high_fit)
vif(high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(high_fit), resid(high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Forest Type Models

```{r}
pondo_df <- transects %>%
  filter(EVT_Group == "ponderosa") %>% #857 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

mixed_df <- transects %>%
  filter(EVT_Group == "mixed conifer") %>% #775 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #170 observations

lodge_df <- transects %>%
  filter(EVT_Group == "lodgepole") %>% #749 obs
  group_by(Transect_ID) %>%
  filter(n() == 10) %>% #only use complete 10-point transects
  ungroup() #200 observations
```

### Ponderosa Model

```{r}
pondo_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = pondo_df, REML = TRUE)

summary(pondo_fit)
vif(pondo_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(pondo_fit), main = "Q-Q Plot of Residuals")
qqline(resid(pondo_fit), col = "red")

# Residuals vs Fitted
plot(fitted(pondo_fit), resid(pondo_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Mixed Conifer Model

```{r}
mixed_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = mixed_df, REML = TRUE)

summary(mixed_fit)
vif(mixed_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(mixed_fit), main = "Q-Q Plot of Residuals")
qqline(resid(mixed_fit), col = "red")

# Residuals vs Fitted
plot(fitted(mixed_fit), resid(mixed_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Lodgepole Model

```{r}
lodge_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = lodge_df, REML = TRUE) #Remove Event_ID in random effects because the only fire is Cameron Peak

summary(lodge_fit)
vif(lodge_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(lodge_fit), main = "Q-Q Plot of Residuals")
qqline(resid(lodge_fit), col = "red")

# Residuals vs Fitted
plot(fitted(lodge_fit), resid(lodge_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Dispersion L/M/H

Uneven sample size and heterogeneity within fires means that the variance within fires varies widely

```{r}
table(transects$Incid_Name)

transects %>%
  st_drop_geometry %>%
  group_by(Incid_Name) %>%
  summarise(n = n(), var_rdnbr = var(rdnbr), sd_rdnbr = sd(rdnbr)) %>%
  arrange(desc(var_rdnbr))
```

### Disp + Low Starting Severity Model

```{r}
low_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = low_df, dispformula = ~Event_ID, REML = TRUE)

summary(low_disp_fit)

sim_low_disp <- simulateResiduals(low_disp_fit)
plot(sim_low_disp)
```

### Disp + Moderate Starting Severity  Model

Warning message:
In finalizeTMB(TMBStruc, obj, fit, h, data.tmb.old) :
  Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting')
  
  -> fixed warning by simplifying random effect structure (just including Transect_ID)

```{r}
mod_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = mod_df, dispformula = ~ Event_ID, REML = TRUE)

summary(mod_disp_fit)

sim_mod_disp <- simulateResiduals(mod_disp_fit)
plot(sim_mod_disp)
```

### Disp + High Starting Severity Model

Warning message:
In finalizeTMB(TMBStruc, obj, fit, h, data.tmb.old) :
  Model convergence problem; non-positive-definite Hessian matrix. See vignette('troubleshooting') 
  
  -> fixed warning by simplifying random effect structure (just including Transect_ID)

```{r}
high_disp_fit <- glmmTMB(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Transect_ID), data = high_df, dispformula = ~ Event_ID, REML = TRUE)

summary(high_disp_fit)

sim_high_disp <- simulateResiduals(high_disp_fit)
plot(sim_high_disp)
```

# Treatment Type Models

Create new dataframes for each "TRT_TYPE" based on points 6-10 for entering and points 1-5 for exiting.

While "TRT_TYPE" is a point level variable, these dataframes will be created based on full transects (i.e. the untreated points in each entering/exiting transect will be included in the dataframe for the treatment type of the treated points).

If treated points in transects have a mix of treatments (n = 9 transects), they will be flagged and put in both relevant dataframes. "Untreated" will only include control transects.


```{r}
#Create Datasets for each treatment type based on the values for points 6-10 if entering and points 1-5 if exiting
treated_half_summary <- transects %>%
  filter(
    (type == "entering" & Point_ID >= 6) |
    (type == "exiting"  & Point_ID <= 5)
  ) %>%
  group_by(Transect_ID, type) %>%
  summarise(
    n_trt_types   = n_distinct(TRT_TYPE),
    trt_types     = paste(sort(unique(TRT_TYPE)), collapse = ", "),
    n_points      = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(n_trt_types))

# How many transects are "mixed" (more than one TRT_TYPE in the treated half)?
mixed_transects <- treated_half_summary %>%
  filter(n_trt_types > 1)
```

Create each dataframe 

```{r}
# Build a lookup of treated-half TRT_TYPE diversity per transect
treated_half_lookup <- transects %>%
  filter(
    (type == "entering" & Point_ID >= 6) |
    (type == "exiting"  & Point_ID <= 5)
  ) %>%
  group_by(Transect_ID) %>%
  summarise(
    n_trt_types    = n_distinct(TRT_TYPE),
    mixed_trt      = n_trt_types > 1,
    # dominant type = the TRT_TYPE with the most points in the treated half
    # ties broken alphabetically for reproducibility -- adjust if you want
    # a different tie-break rule
    dominant_trt_type = TRT_TYPE[which.max(table(TRT_TYPE)[TRT_TYPE])],
    .groups = "drop"
  ) %>%
  select(Transect_ID, mixed_trt, dominant_trt_type)

# Untreated: whole control transects, flagged as not mixed (no treated half)
untreated_df <- transects %>%
  filter(type == "control") %>%
  mutate(mixed_trt = FALSE, dominant_trt_type = NA_character_)

# Helper: find transects whose TREATED half contains `trt` anywhere,
# return the full transect with mixed_trt / dominant_trt_type attached
make_trt_df <- function(df, trt, lookup) {
  ids <- df %>%
    filter(
      (type == "entering" & Point_ID >= 6 & TRT_TYPE == trt) |
      (type == "exiting"  & Point_ID <= 5 & TRT_TYPE == trt)
    ) %>%
    pull(Transect_ID) %>%
    unique()

  df %>%
    filter(Transect_ID %in% ids) %>%
    left_join(lookup, by = "Transect_ID")
}

rearrangement_df <- make_trt_df(transects, "Rearrangement", treated_half_lookup)
removal_df <- make_trt_df(transects, "Removal", treated_half_lookup)
broadcast_df <- make_trt_df(transects, "Broadcast Burn", treated_half_lookup)
wildfire_df <- make_trt_df(transects, "Wildfire", treated_half_lookup)
```

### Summary of Transects per Fire

```{r}
untreated_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))

rearrangement_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))

removal_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))

broadcast_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))

wildfire_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
```

### Untreated Model

Does not include TRT_TYPE since they are all untreated

Does not include Point_ID*type interaction since these are all controls

```{r}
untreated_fit <- lmer(rdnbr ~ (1|Event_ID/Transect_ID), data = untreated_df, REML = TRUE)

summary(untreated_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(untreated_fit), main = "Q-Q Plot of Residuals")
qqline(resid(untreated_fit), col = "red")

# Residuals vs Fitted
plot(fitted(untreated_fit), resid(untreated_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Rearrangement Model

Back to including TRT_TYPE and Point_ID*type

```{r}
rearrangement_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = rearrangement_df, REML = TRUE)

summary(rearrangement_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(rearrangement_fit), main = "Q-Q Plot of Residuals")
qqline(resid(rearrangement_fit), col = "red")

# Residuals vs Fitted
plot(fitted(rearrangement_fit), resid(rearrangement_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))

# Filtering out the mixed transects
rearrangement_pure_df <- rearrangement_df %>%
  filter(mixed_trt == FALSE)

rearrangement_pure_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = rearrangement_pure_df, REML = TRUE)

summary(rearrangement_pure_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(rearrangement_pure_fit), main = "Q-Q Plot of Residuals")
qqline(resid(rearrangement_pure_fit), col = "red")

# Residuals vs Fitted
plot(fitted(rearrangement_pure_fit), resid(rearrangement_pure_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Removal Model

```{r}
removal_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = removal_df, REML = TRUE)

summary(removal_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(removal_fit), main = "Q-Q Plot of Residuals")
qqline(resid(removal_fit), col = "red")

# Residuals vs Fitted
plot(fitted(removal_fit), resid(removal_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))

# Filtering out the mixed transects
removal_pure_df <- removal_df %>%
  filter(mixed_trt == FALSE)

removal_pure_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = removal_pure_df, REML = TRUE)

summary(removal_pure_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(removal_pure_fit), main = "Q-Q Plot of Residuals")
qqline(resid(removal_pure_fit), col = "red")

# Residuals vs Fitted
plot(fitted(removal_pure_fit), resid(removal_pure_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Broadcast Model

```{r}
broadcast_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = broadcast_df, REML = TRUE)

summary(broadcast_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(broadcast_fit), main = "Q-Q Plot of Residuals")
qqline(resid(broadcast_fit), col = "red")

# Residuals vs Fitted
plot(fitted(broadcast_fit), resid(broadcast_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))

# Filtering out the mixed transects
broadcast_pure_df <- broadcast_df %>%
  filter(mixed_trt == FALSE)

broadcast_pure_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = broadcast_pure_df, REML = TRUE)

summary(broadcast_pure_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(broadcast_pure_fit), main = "Q-Q Plot of Residuals")
qqline(resid(broadcast_pure_fit), col = "red")

# Residuals vs Fitted
plot(fitted(broadcast_pure_fit), resid(broadcast_pure_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### Wildfire Model

```{r}
wildfire_fit <- lmer(rdnbr ~ TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = wildfire_df, REML = TRUE)

summary(wildfire_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(wildfire_fit), main = "Q-Q Plot of Residuals")
qqline(resid(wildfire_fit), col = "red")

# Residuals vs Fitted
plot(fitted(wildfire_fit), resid(wildfire_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Entering - Gavin Model

Trying to recreate Gavin's model, but with a random effect for fire as well and not using a Poisson log-link distribution.

rdnbr ~ type + treated + (1|Event_ID/Transect_ID)

Where type = control or entering & treated = untreated or broadcast/wildfire

And run separate models for transects starting in L/M/H

```{r}
# Filter by TRT_TYPE to only include broadcast or wildfire
black_df <- transects %>%
  filter(!TRT_TYPE %in% c("Rearrangement", "Removal")) %>%
  filter(type != "exiting")

black_df$TRT_TYPE <- factor(black_df$TRT_TYPE, levels = c("Untreated", "Broadcast Burn", "Wildfire"))
black_df$type <- factor(black_df$type, levels = c("control", "entering"))

black_low_df <- black_df %>%
  filter(START_SEV == "Low") #593 obs
black_mod_df <- black_df %>%
  filter(START_SEV == "Moderate") #352 obs
black_high_df <- black_df %>%
  filter(START_SEV == "High") #1385 obs

black_low_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
black_mod_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
black_high_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
```

### Low Jones Model

```{r}
black_low_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_low_df, REML = TRUE)

summary(black_low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_low_fit), resid(black_low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Moderate Jones Model

Fit is singular, fixed when removing Event_ID random effect

```{r}
black_mod_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_mod_df, REML = TRUE)

summary(black_mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_mod_fit), resid(black_mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### High Jones Model

```{r}
black_high_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_high_df, REML = TRUE)

summary(black_high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black_high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black_high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black_high_fit), resid(black_high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Plot effect

```{r}
# Extract population-level predictions (averaged over random effects)
pred_low  <- ggpredict(black_low_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Low") %>%
  filter(!is.na(predicted))

pred_mod  <- ggpredict(black_mod_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Moderate") %>%
  filter(!is.na(predicted))

pred_high <- ggpredict(black_high_fit, terms = c("type", "treated")) %>%
  mutate(severity = "High") %>%
  filter(!is.na(predicted))

# Combine, drop the non-existent control+treated combination,
# and relabel x-axis categories
pred_all <- bind_rows(pred_low, pred_mod, pred_high) %>%
  filter(!(x == "control" & group == "Yes"))

pred_all$severity <- factor(pred_all$severity, levels = c("High", "Moderate", "Low"))

# Plot
ggplot(pred_all, aes(x = x, y = predicted, group = group)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.15, position = position_dodge(width = 0.3)) +
  facet_wrap(~severity, nrow = 1) +
  labs(
    x = NULL,
    y = "Predicted RdNBR"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

# Exiting - Gavin Model

Trying to recreate Gavin's model, but with a random effect for fire as well and not using a Poisson log-link distribution.

rdnbr ~ type + treated + (1|Event_ID/Transect_ID)

Where type = control or exiting & treated = untreated or broadcast/wildfire

And run separate models for transects starting in L/M/H

```{r}
# Filter by TRT_TYPE to only include broadcast or wildfire
black2_df <- transects %>%
  filter(!TRT_TYPE %in% c("Rearrangement", "Removal")) %>%
  filter(type != "entering")

black2_df$TRT_TYPE <- factor(black2_df$TRT_TYPE, levels = c("Untreated", "Broadcast Burn", "Wildfire"))
black2_df$type <- factor(black2_df$type, levels = c("control", "exiting"))

black2_low_df <- black2_df %>%
  filter(START_SEV == "Low") #593 obs
black2_mod_df <- black2_df %>%
  filter(START_SEV == "Moderate") #352 obs
black2_high_df <- black2_df %>%
  filter(START_SEV == "High") #1385 obs

black2_low_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
black2_mod_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
black2_high_df %>%
    distinct(Transect_ID, Incid_Name) %>%
    count(Incid_Name, name = "n_transects") %>%
    arrange(desc(n_transects))
```

### Low Jones Model

```{r}
black2_low_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_low_df, REML = TRUE)

summary(black2_low_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_low_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_low_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_low_fit), resid(black2_low_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Moderate Jones Model

Fit is singular, fixed when removing Event_ID random effect

```{r}
black2_mod_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_mod_df, REML = TRUE)

summary(black2_mod_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_mod_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_mod_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_mod_fit), resid(black2_mod_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### High Jones Model

```{r}
black2_high_fit <- lmer(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_high_df, REML = TRUE)

summary(black2_high_fit)

# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(black2_high_fit), main = "Q-Q Plot of Residuals")
qqline(resid(black2_high_fit), col = "red")

# Residuals vs Fitted
plot(fitted(black2_high_fit), resid(black2_high_fit),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1)) 
```

### Plot effect

```{r}
# Extract population-level predictions (averaged over random effects)
pred_low2  <- ggpredict(black2_low_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Low") %>%
  filter(!is.na(predicted))

pred_mod2  <- ggpredict(black2_mod_fit,  terms = c("type", "treated")) %>%
  mutate(severity = "Moderate") %>%
  filter(!is.na(predicted))

pred_high2 <- ggpredict(black2_high_fit, terms = c("type", "treated")) %>%
  mutate(severity = "High") %>%
  filter(!is.na(predicted))

# Combine, drop the non-existent control+treated combination,
# and relabel x-axis categories
pred_all2 <- bind_rows(pred_low2, pred_mod2, pred_high2) %>%
  filter(!(x == "control" & group == "Yes"))

pred_all2$severity <- factor(pred_all2$severity, levels = c("High", "Moderate", "Low"))

pred_all2$group <- factor(pred_all2$group, levels = c("Yes", "No"))

# Plot
ggplot(pred_all2, aes(x = x, y = predicted, group = group)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.15, position = position_dodge(width = 0.3)) +
  facet_wrap(~severity, nrow = 1) +
  labs(
    x = NULL,
    y = "Predicted RdNBR"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

# Dispersion Gavin Models

```{r}
black_low_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_low_df, dispformula = ~ Event_ID, REML = TRUE)

summary(black_low_disp_fit)

sim_black_low_disp <- simulateResiduals(black_low_disp_fit)
plot(sim_black_low_disp)
```

```{r}
black_mod_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_mod_df, dispformula = ~ Event_ID, REML = TRUE)

summary(black_mod_disp_fit)

sim_black_mod_disp <- simulateResiduals(black_mod_disp_fit)
plot(sim_black_mod_disp)
```

```{r}
black_high_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black_high_df, dispformula = ~ Event_ID, REML = TRUE) # Convergence problem

summary(black_high_disp_fit)

sim_black_high_disp <- simulateResiduals(black_high_disp_fit)
plot(sim_black_high_disp)
```

```{r}
black2_low_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_low_df, dispformula = ~ Event_ID, REML = TRUE)

summary(black2_low_disp_fit)

sim_black2_low_disp <- simulateResiduals(black2_low_disp_fit)
plot(sim_black2_low_disp)
```

```{r}
black2_mod_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_mod_df, dispformula = ~ Event_ID, REML = TRUE)

summary(black2_mod_disp_fit)

sim_black2_mod_disp <- simulateResiduals(black2_mod_disp_fit)
plot(sim_black2_mod_disp)
```

```{r}
black2_high_disp_fit <- glmmTMB(rdnbr ~ type + treated + (1|Event_ID/Transect_ID), data = black2_high_df, dispformula = ~ Event_ID, REML = TRUE) # Convergence problem

summary(black2_high_disp_fit)

sim_black2_high_disp <- simulateResiduals(black2_high_disp_fit)
plot(sim_black2_high_disp)
```
