---
title: "6--Regression-Models"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(sf)
library(lme4)
library(glmmTMB)
library(DHARMa)
library(ggeffects)
library(brms)

transects <- st_read("data/processed/fuels_analysis.gpkg", layer = "transectsDS") 
transects$TCC <- as.numeric(transects$TCC)
transects$TRT_TYPE[is.na(transects$TRT_TYPE)] <- "Untreated"
transects$TRT_TYPE <- factor(transects$TRT_TYPE, levels = c("Untreated", "Thinning", "Removal", "Broadcast Burn", "Wildfire"))
transects$type <- factor(transects$type, levels = c("control", "entering", "exiting"))

#Add 0 Age for untreated points and create treated y/n column to use in interaction with Age
transects$treated <- ifelse(transects$TRT_TYPE == "Untreated", "No", "Yes")
transects$treated <- factor(transects$treated, levels = c("No", "Yes"))
transects$AGE <- ifelse(transects$TRT_TYPE == "Untreated", 0, transects$AGE)
```

## Response Variable Distribution

```{r}
hist(transects$cbi)
hist(transects$rdnbr)

plot(transects$cbi, transects$rdnbr)
```

## Look at shape of relationship between each predictor and cbi

```{r}
transects_test <- filter(transects, Incid_Name != "CAMERON PEAK")
ggplot(transects, aes(x = TRT_TYPE, y = cbi)) +
  geom_boxplot()
ggplot(transects_test, aes(x = TRT_TYPE, y = cbi)) +
  geom_boxplot()

ggplot(transects, aes(x = COUNT, y = cbi)) +
  geom_point()
ggplot(transects_test, aes(x = COUNT, y = cbi)) +
  geom_point()

ggplot(transects, aes(x = AGE, y = cbi)) +
  geom_point()
ggplot(transects_test, aes(x = AGE, y = cbi)) +
  geom_point()


ggplot(transects, aes(x = Point_ID, y = cbi, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(x = "Point ID", y = "CBI", color = "Type")
ggplot(transects_test, aes(x = Point_ID, y = cbi, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = TRUE) +
  theme_minimal() +
  labs(x = "Point ID", y = "CBI", color = "Type")

ggplot(transects, aes(x = Point_ID, y = cbi, color = type, group = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~TRT_TYPE) +
  theme_minimal() +
  labs(x = "Point ID", y = "CBI", color = "Type")
```

## Create linear regression model including nested random effects of transect ID and fire ID. Fixed effects include treatment characteristics (type, age, number, and prev. burn severity), and an interaction between point distance and transect type

- prev burn sev

```{r}
m1 <- lmer(cbi ~ TRT_TYPE + COUNT + AGE:TRT_TYPE + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m1)
vif(m1)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m1), main = "Q-Q Plot of Residuals")
qqline(resid(m1), col = "red")

# Residuals vs Fitted
plot(fitted(m1), resid(m1),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Ordered Beta Regression

https://www.cambridge.org/core/journals/political-analysis/article/ordered-beta-regression-a-parsimonious-wellfitting-model-for-continuous-data-with-lower-and-upper-bounds/89F4141DA16D4FC217809B5EB45EEE83 

Ordered beta allows for exact 0s and 1s

```{r}
transects$cbi_scaled <- transects$cbi/3

m2 <- glmmTMB(cbi_scaled ~ TRT_TYPE + COUNT + Point_ID*type + (1|Event_ID/Transect_ID), family = ordbeta, data = transects)

sim_beta <- simulateResiduals(m2)
plot(sim_beta)
```


```{r}
# Check residual patterns against key predictors directly
plotResiduals(sim_beta, form = transects$TRT_TYPE)
plotResiduals(sim_beta, form = transects$type)
plotResiduals(sim_beta, form = transects$COUNT)
```

## Try a model with rdnbr

```{r}
m3 <- lmer(rdnbr ~ TRT_TYPE + COUNT + AGE + Point_ID*type + (1|Event_ID/Transect_ID), data = transects, REML = TRUE)

summary(m3)
vif(m3)
```

## Evaluating model assumptions

Fitted vs residuals plot
```{r}
# Residual diagnostic plots for m1
par(mfrow = c(1, 2))

# Q-Q plot
qqnorm(resid(m3), main = "Q-Q Plot of Residuals")
qqline(resid(m3), col = "red")

# Residuals vs Fitted
plot(fitted(m3), resid(m3),
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```
