---
title: "7--Random-Forest"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE)
```

# Load Packages

```{r}
library(tidyverse)
library(sf)
library(here)
library(ranger)
library(tidymodels)
library(DALEXtra) #Explainer (variable importance, partial dependence, etc)
```

# Load Data

```{r, results = FALSE}
plots <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "allplotsDS") %>%
  st_drop_geometry()
plots$northness[is.na(plots$northness)] <- 0
plots$eastness[is.na(plots$eastness)] <- 0

plots <- plots %>%
  mutate(across(c("northness", "eastness", "elevation", "slope", "TMAX", "PRCP", "TMIN", "TOTAL_COUNT", "prop_Rearrangement", "prop_Wildfire", "prop_Broadcast.Burn", "prop_Removal", "prop_extent_treated", "TCC", "prop_aspen", "prop_lodgepole", "prop_mixed.conifer", "prop_other", "prop_pinyon.juniper", "prop_ponderosa", "prop_spruce.fir"), as.numeric))


plots_100m <- filter(plots, buffer_radius == 100)
plots_250m <- filter(plots, buffer_radius == 250)
plots_500m <- filter(plots, buffer_radius == 500)

preds <- c("northness", "slope", "PRCP", "TMAX", "prop_extent_treated",  "TCC") #For use later in VIPs

plots_100m %>%
  group_by(Incid_Name) %>%
  summarize(mean(prop_mod_high_sev))
```

# Proportion of High-Severity

## 100m

```{r}
set.seed(123)
data_split_hi100 <- initial_split(plots_100m, strata = "prop_high_sev", prop = 0.80)
hi100_train <- training(data_split_hi100)
hi100_test  <- testing(data_split_hi100)
```

Create a recipe
```{r}
hi100_rec <- recipe(prop_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = hi100_train)

hi100_prep <- prep(hi100_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi100_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
hi100_folds <- group_vfold_cv(hi100_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

# hi100_folds %>% 
#   mutate(held_out_fire = map(splits, ~ assessment(.x) %>% distinct(Incid_Name))) %>% 
#   unnest(held_out_fire) %>% 
#   select(id, Incid_Name)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi100_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = hi100_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_hi100)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
hi100_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(hi100_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 100 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
hi100_explainer <- explain_tidymodels(
  final_fitted,
  data = hi100_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi100_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi100 <- variable_importance(hi100_explainer,
                                variables = preds)
plot(vip_hi100) 
```

PDP
```{r}
pdp_hi100 <- model_profile(
  hi100_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi100)
```

Shaply Plot - just for fun
```{r}
shap_plots <- predict_parts(hi100_explainer, hi100_test[96, ], type = 'shap')
plot(shap_plots)
```

## 250m

```{r}
set.seed(123)
data_split_hi250 <- initial_split(plots_250m, strata = "prop_high_sev", prop = 0.8)
hi250_train <- training(data_split_hi250)
hi250_test  <- testing(data_split_hi250)
```

Create a recipe
```{r}
hi250_rec <- recipe(prop_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = hi250_train)

hi250_prep <- prep(hi250_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi250_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
hi250_folds <- group_vfold_cv(hi250_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi250_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = hi250_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_hi250)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
hi250_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(hi250_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 250 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
hi250_explainer <- explain_tidymodels(
  final_fitted,
  data = hi250_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi250_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi250 <- variable_importance(hi250_explainer,
                                variables = preds)
plot(vip_hi250) 
```

PDP
```{r}
pdp_hi250 <- model_profile(
  hi250_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi250)
```

## 500m

```{r}
set.seed(123)
data_split_hi500 <- initial_split(plots_500m, strata = "prop_high_sev", prop = 0.8)
hi500_train <- training(data_split_hi500)
hi500_test  <- testing(data_split_hi500)
```

Create a recipe
```{r}
hi500_rec <- recipe(prop_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = hi500_train)

hi500_prep <- prep(hi500_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi500_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
hi500_folds <- group_vfold_cv(hi500_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi500_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = hi500_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_hi500)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
hi500_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(hi500_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 500 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
hi500_explainer <- explain_tidymodels(
  final_fitted,
  data = hi500_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi500_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi500 <- variable_importance(hi500_explainer,
                                variables = preds)
plot(vip_hi500) 
```

PDP
```{r}
pdp_hi500 <- model_profile(
  hi500_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi500)
```

# Proportion of Moderate to High-severity

## 100m
```{r}
set.seed(123)
data_split_mohi100 <- initial_split(plots_100m, strata = "prop_mod_high_sev", prop = 0.8)
mohi100_train <- training(data_split_mohi100)
mohi100_test  <- testing(data_split_mohi100)
```

Create a recipe
```{r}
mohi100_rec <- recipe(prop_mod_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = mohi100_train)

mohi100_prep <- prep(mohi100_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi100_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi100_folds <- group_vfold_cv(mohi100_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi100_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = mohi100_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_mohi100)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
mohi100_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(mohi100_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 100 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
mohi100_explainer <- explain_tidymodels(
  final_fitted,
  data = mohi100_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi100_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi100 <- variable_importance(mohi100_explainer,
                                variables = preds)
plot(vip_mohi100) 
```

PDP
```{r}
pdp_mohi100 <- model_profile(
  mohi100_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi100)
```

## 250m

```{r}
set.seed(123)
data_split_mohi250 <- initial_split(plots_250m, strata = "prop_mod_high_sev", prop = 0.8)
mohi250_train <- training(data_split_mohi250)
mohi250_test  <- testing(data_split_mohi250)
```

Create a recipe
```{r}
mohi250_rec <- recipe(prop_mod_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = mohi250_train)

mohi250_prep <- prep(mohi250_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi250_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi250_folds <- group_vfold_cv(mohi250_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi250_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = mohi250_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_mohi250)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
mohi250_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(mohi250_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 250 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
mohi250_explainer <- explain_tidymodels(
  final_fitted,
  data = mohi250_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi250_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi250 <- variable_importance(mohi250_explainer,
                                variables = preds)
plot(vip_mohi250) 
```

PDP
```{r}
pdp_mohi250 <- model_profile(
  mohi250_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi250)
```


## 500m

```{r}
set.seed(123)
data_split_mohi500 <- initial_split(plots_500m, strata = "prop_mod_high_sev", prop = 0.8)
mohi500_train <- training(data_split_mohi500)
mohi500_test  <- testing(data_split_mohi500)
```

Create a recipe
```{r}
mohi500_rec <- recipe(prop_mod_high_sev ~ northness + slope + PRCP + TMAX + prop_extent_treated + TCC, data = mohi500_train)

mohi500_prep <- prep(mohi500_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi500_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi500_folds <- group_vfold_cv(mohi500_train, group= "Incid_Name")
#can also use spatial_block_cv for true spatial rather than leaving fires out (needs geometry)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 6)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi500_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow %>% 
    fit_resamples(
      resamples = mohi500_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 
```

Fit Final Model
```{r}
final_fit <- final_workflow %>%
  last_fit(data_split_mohi500)

collect_metrics(final_fit)      # test-set RMSE, R², etc.
mohi500_predictions <- collect_predictions(final_fit)  # observed vs. predicted on the held-out test set
ggplot(mohi500_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 500 m"
  )

final_fitted <- final_fit$.workflow[[1]]
final_fitted
```

Explainer
```{r}
mohi500_explainer <- explain_tidymodels(
  final_fitted,
  data = mohi500_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi500_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi500 <- variable_importance(mohi500_explainer,
                                variables = preds)
plot(vip_mohi500) 
```

PDP
```{r}
pdp_mohi500 <- model_profile(
  mohi500_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi500)
```


# Figures

## Variable Importance, Proportion of High Severity

```{r}
plot(vip_500, vip_1000, title = "Variable Importance by buffer scale")

plot(vip_plot, max_vars = 15, show_boxplots = FALSE) +
  ggtitle("Variable Importance, 500 m")
```

## Partial Dependence Plot, Proportion of High Severity

```{r}

```

## Variable Importance, Proportion of Moderate-High Severity

```{r}

```

## Partial Dependence Plot, Proportion of Moderate-High Severity

```{r}

```

#testing out interaction pdp
```{r}
# bin max temp into terciles (adjust name/breaks to your data)
dat <- mohi500_explainer$data |>
  mutate(tmax_class = cut(TMAX,
                          breaks = quantile(TMAX, c(0, 1/3, 2/3, 1), na.rm = TRUE),
                          labels = c("Low", "Mid", "High"),
                          include.lowest = TRUE))

mohi500_explainer$data <- dat

pdp_extent_tmax <- model_profile(
  mohi500_explainer,
  variables = "prop_extent_treated",
  groups = "tmax_class",
  N = NULL
)

plot(pdp_extent_tmax,
     title = "Effect of treated extent by max temperature",
     subtitle = "")
```

```{r}
mohi500_explainer$data <- mohi500_explainer$data |>
  mutate(tmax_class = factor(
    if_else(TMAX <= median(TMAX, na.rm = TRUE), "Low", "High"),
    levels = c("Low", "High")
  ))

pdp_extent_tmax <- model_profile(
  mohi500_explainer,
  variables = "prop_extent_treated",
  groups = "tmax_class",
  N = NULL
)

plot(pdp_extent_tmax,
     title = "Effect of treated extent by max temperature",
     subtitle = "")
```

```{r}
mohi500_explainer$data <- mohi500_explainer$data |>
  mutate(slope_class = factor(
    if_else(slope <= median(slope, na.rm = TRUE), "Mild", "Steep"),
    levels = c("Mild", "Steep")
  ))

pdp_extent_slope <- model_profile(
  mohi500_explainer,
  variables = "prop_extent_treated",
  groups = "slope_class",
  N = NULL
)

plot(pdp_extent_slope,
     title = "Effect of treated extent by slope",
     subtitle = "")
```

