---
title: "7--Random-Forest"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = TRUE)
```

# Load Packages

```{r, include=FALSE}
library(tidyverse)
library(sf)
library(here)
library(ranger)
library(tidymodels)
library(DALEXtra) #Explainer (variable importance, partial dependence, etc)
library(spatialsample) #Use for spatial cross validation
library(car) #VIF function
library(pdp) #for use in heatmaps

conflicted::conflicts_prefer(yardstick::rmse)
conflicted::conflicts_prefer(dplyr::filter)
```

# Load Data

```{r, results = FALSE}
plots <- st_read(here("data/processed/fuels_analysis.gpkg"), layer = "allplotsDS") %>%
  filter(Incid_Name != "EAST FORK") #seeing very little treated area in this fire now
plots$northness[is.na(plots$northness)] <- 0
plots$eastness[is.na(plots$eastness)] <- 0

plots_100m <- filter(plots, buffer_radius == 100)
plots_250m <- filter(plots, buffer_radius == 250)
plots_500m <- filter(plots, buffer_radius == 500)

plots_500m %>%
  st_drop_geometry() %>%
  group_by(Incid_Name) %>%
  summarize(mean(prop_shrubland))
```

# Correlation of Predictor Variables

VIF and linear correlation

```{r}
model <- lm(prop_high_sev ~ extreme_area + northness + eastness + elevation + TPI + slope + Tmax_4k + PR_4k + VS_4k + VPD_4k + ERC_4k + CWD_4k + PDSI_4k + prop_extent_treated + prop_conifer + prop_conifer_hardwood + prop_hardwood + prop_shrubland + prop_grassland + TCC, data = plots)

vif(model)

#removing individual prop_conifer_hardwoord
model2 <- lm(prop_high_sev ~ extreme_area + northness + eastness + elevation + TPI + slope + Tmax_4k + PR_4k + VS_4k + VPD_4k + ERC_4k + CWD_4k + PDSI_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + prop_grassland + TCC, data = plots)
vif(model2)

#Removing TMAX for collinearity with vpd
model3 <- lm(prop_high_sev ~ extreme_area + northness + eastness + elevation + TPI + slope + PR_4k + VS_4k + VPD_4k + ERC_4k + CWD_4k + PDSI_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + prop_grassland + TCC, data = plots)
vif(model3)

#In preds set, also removing monthly CWD and PDSI, and daily ERC, precip, VPD, and temperature because of nonsense pdp plots. Getting rid of prop_grassland because all but one fire has <1% grassland (plateau has 2%). Getting rid of eastness for now for simplicity, also it is a weak relationship but it is in the opposite direction from what I'd expect
preds <- c("DailyAreaBurned", "extreme_area", "northness", "elevation", "TPI", "slope", "VS_4k", "prop_extent_treated", "prop_conifer", "prop_hardwood", "prop_shrubland", "TCC") #For use later in VIPs

model4 <- lm(prop_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = plots)
vif(model4)
```

# Proportion of High-Severity

## 100m

```{r}
set.seed(123)
data_split_hi100 <- initial_split(plots_100m, strata = "prop_high_sev", prop = 0.80)
hi100_train <- training(data_split_hi100)
hi100_test  <- testing(data_split_hi100)
```

Create a recipe
```{r}
hi100_rec <- recipe(prop_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = hi100_train)

hi100_prep <- prep(hi100_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi100_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)

hi100_folds <- spatial_clustering_cv(hi100_train)

# hi100_folds %>% 
#   mutate(held_out_fire = map(splits, ~ assessment(.x) %>% distinct(Incid_Name))) %>% 
#   unnest(held_out_fire) %>% 
#   select(id, Incid_Name)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi100_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_hi100 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_hi100 %>% 
    fit_resamples(
      resamples = hi100_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_hi100 <- final_workflow_hi100 %>%
  last_fit(data_split_hi100)

collect_metrics(final_fit_hi100)      # test-set RMSE, R², etc.
hi100_predictions <- collect_predictions(final_fit_hi100)  # observed vs. predicted on the held-out test set
ggplot(hi100_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 100 m"
  )

final_fitted_hi100 <- final_fit_hi100$.workflow[[1]]
final_fitted_hi100
```

Explainer
```{r}
hi100_explainer <- explain_tidymodels(
  final_fitted_hi100,
  data = hi100_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi100_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi100 <- variable_importance(hi100_explainer,
                                variables = preds)
plot(vip_hi100) 
```

PDP
```{r}
pdp_hi100 <- model_profile(
  hi100_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_hi100)

pdp_hi100 <- model_profile(
  hi100_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi100)
```

Shaply Plot - just for fun
```{r}
# shap_plots <- predict_parts(hi100_explainer, hi100_test[96, ], type = 'shap')
# plot(shap_plots)
```

ALE plot
```{r}
# ale_hi100 <- model_profile(
#   explainer = hi100_explainer,
#   variables = preds,
#   type = 'accumulated'      
# )
# 
# plot(ale_hi100)
```


## 250m

```{r}
set.seed(123)
data_split_hi250 <- initial_split(plots_250m, strata = "prop_high_sev", prop = 0.8)
hi250_train <- training(data_split_hi250)
hi250_test  <- testing(data_split_hi250)
```

Create a recipe
```{r}
hi250_rec <- recipe(prop_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = hi250_train)

hi250_prep <- prep(hi250_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi250_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
hi250_folds <- spatial_clustering_cv(hi250_train)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi250_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_hi250 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_hi250 %>% 
    fit_resamples(
      resamples = hi250_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_hi250 <- final_workflow_hi250 %>%
  last_fit(data_split_hi250)

collect_metrics(final_fit_hi250)      # test-set RMSE, R², etc.
hi250_predictions <- collect_predictions(final_fit_hi250)  # observed vs. predicted on the held-out test set
ggplot(hi250_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 250 m"
  )

final_fitted_hi250 <- final_fit_hi250$.workflow[[1]]
final_fitted_hi250
```

Explainer
```{r}
hi250_explainer <- explain_tidymodels(
  final_fitted_hi250,
  data = hi250_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi250_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi250 <- variable_importance(hi250_explainer,
                                variables = preds)
plot(vip_hi250) 
```

PDP
```{r}
pdp_hi250 <- model_profile(
  hi250_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_hi250)

pdp_hi250 <- model_profile(
  hi250_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi250)
```

## 500m

```{r}
set.seed(123)
data_split_hi500 <- initial_split(plots_500m, strata = "prop_high_sev", prop = 0.8)
hi500_train <- training(data_split_hi500)
hi500_test  <- testing(data_split_hi500)
```

Create a recipe
```{r}
hi500_rec <- recipe(prop_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = hi500_train)

hi500_prep <- prep(hi500_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(hi500_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
hi500_folds <- spatial_clustering_cv(hi500_train)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = hi500_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_hi500 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_hi500 %>% 
    fit_resamples(
      resamples = hi500_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_hi500 <- final_workflow_hi500 %>%
  last_fit(data_split_hi500)

collect_metrics(final_fit_hi500)      # test-set RMSE, R², etc.
hi500_predictions <- collect_predictions(final_fit_hi500)  # observed vs. predicted on the held-out test set
ggplot(hi500_predictions, aes(prop_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of High Severity Fire, 500 m"
  )

final_fitted_hi500 <- final_fit_hi500$.workflow[[1]]
final_fitted_hi500
```

Explainer
```{r}
hi500_explainer <- explain_tidymodels(
  final_fitted_hi500,
  data = hi500_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = hi500_train$prop_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_hi500 <- variable_importance(hi500_explainer,
                                variables = preds)
plot(vip_hi500) 
```

PDP
```{r}
pdp_hi500 <- model_profile(
  hi500_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_hi500)

pdp_hi500 <- model_profile(
  hi500_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_hi500)
```

# Proportion of Moderate to High-severity

## 100m
```{r}
set.seed(123)
data_split_mohi100 <- initial_split(plots_100m, strata = "prop_mod_high_sev", prop = 0.8)
mohi100_train <- training(data_split_mohi100)
mohi100_test  <- testing(data_split_mohi100)
```

Create a recipe
```{r}
mohi100_rec <- recipe(prop_mod_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = mohi100_train)

mohi100_prep <- prep(mohi100_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi100_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi100_folds <- spatial_clustering_cv(mohi100_train)

mohi100_folds %>%
  mutate(held_out_fire = map(splits, ~ assessment(.x) %>% distinct(Incid_Name))) %>%
  unnest(held_out_fire) %>%
  select(id, Incid_Name)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi100_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_mohi100 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_mohi100 %>% 
    fit_resamples(
      resamples = mohi100_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_mohi100 <- final_workflow_mohi100 %>%
  last_fit(data_split_mohi100)

collect_metrics(final_fit_mohi100)      # test-set RMSE, R², etc.
mohi100_predictions <- collect_predictions(final_fit_mohi100)  # observed vs. predicted on the held-out test set
ggplot(mohi100_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 100 m"
  )

final_fitted_mohi100 <- final_fit_mohi100$.workflow[[1]]
final_fitted_mohi100
```

Explainer
```{r}
mohi100_explainer <- explain_tidymodels(
  final_fitted_mohi100,
  data = mohi100_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi100_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi100 <- variable_importance(mohi100_explainer,
                                variables = preds)
plot(vip_mohi100) 
```

PDP
```{r}
pdp_mohi100 <- model_profile(
  mohi100_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_mohi100)

pdp_mohi100 <- model_profile(
  mohi100_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi100)
```

## 250m

```{r}
set.seed(123)
data_split_mohi250 <- initial_split(plots_250m, strata = "prop_mod_high_sev", prop = 0.8)
mohi250_train <- training(data_split_mohi250)
mohi250_test  <- testing(data_split_mohi250)
```

Create a recipe
```{r}
mohi250_rec <- recipe(prop_mod_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = mohi250_train)

mohi250_prep <- prep(mohi250_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi250_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi250_folds <- spatial_clustering_cv(mohi250_train)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi250_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_mohi250 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_mohi250 %>% 
    fit_resamples(
      resamples = mohi250_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_mohi250 <- final_workflow_mohi250 %>%
  last_fit(data_split_mohi250)

collect_metrics(final_fit_mohi250)      # test-set RMSE, R², etc.
mohi250_predictions <- collect_predictions(final_fit_mohi250)  # observed vs. predicted on the held-out test set
ggplot(mohi250_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 250 m"
  )

final_fitted_mohi250 <- final_fit_mohi250$.workflow[[1]]
final_fitted_mohi250
```

Explainer
```{r}
mohi250_explainer <- explain_tidymodels(
  final_fitted_mohi250,
  data = mohi250_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi250_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi250 <- variable_importance(mohi250_explainer,
                                variables = preds)
plot(vip_mohi250) 
```

PDP
```{r}
pdp_mohi250 <- model_profile(
  mohi250_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_mohi250)

pdp_mohi250 <- model_profile(
  mohi250_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi250)
```


## 500m

```{r}
set.seed(123)
data_split_mohi500 <- initial_split(plots_500m, strata = "prop_mod_high_sev", prop = 0.8)
mohi500_train <- training(data_split_mohi500)
mohi500_test  <- testing(data_split_mohi500)
```

Create a recipe
```{r}
mohi500_rec <- recipe(prop_mod_high_sev ~ DailyAreaBurned + extreme_area + northness + elevation + TPI + slope + VS_4k + prop_extent_treated + prop_conifer + prop_hardwood + prop_shrubland + TCC, data = mohi500_train)

mohi500_prep <- prep(mohi500_rec)
```

Set up a model specification to tune
```{r}
tune_spec <- rand_forest(mtry = tune(), 
                         trees = 2000,
                         min_n = tune()) %>%
              set_mode("regression") %>%
              set_engine("ranger", importance = "permutation")
```

Set up a workflow
```{r}
tune_wf <- workflow() %>%
  add_recipe(mohi500_rec) %>%
  add_model(tune_spec)
```

Tune hyperparameters based on RMSE
```{r}
set.seed(234)
mohi500_folds <- spatial_clustering_cv(mohi500_train)

doParallel::registerDoParallel()
#show_notes(.Last.tune.result)

rf_grid <- grid_regular(
  mtry(range = c(1, 12)),
  min_n(range = c(1,40)),
  levels = 5
)

set.seed(456)
regular_res <- tune_grid(
  tune_wf,
  resamples = mohi500_folds,
  grid = rf_grid
)

regular_res %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  mutate(min_n = factor(min_n)) %>%
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(alpha = 0.5, size = 1.5) +
  geom_point()
```

Select Best Fit
```{r}
best_rmse <- select_best(regular_res, metric = "rmse")
final_workflow_mohi500 <- tune_wf %>%
  finalize_workflow(best_rmse)
```

Evaluate metrics across spatial crossvalidation folds
```{r}
rf_res <- 
    final_workflow_mohi500 %>% 
    fit_resamples(
      resamples = mohi500_folds, 
      metrics = metric_set(rmse, rsq),
      control = control_resamples(save_pred = TRUE)
      ) 
  
rf_res %>%  
  collect_metrics(summarize = TRUE) %>%
  mutate(Dataset="training") %>% 
  mutate(Model="RF") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  filter(.metric == "rmse") 

rf_res %>% 
  collect_metrics(summarize = FALSE) %>% 
  dplyr::filter(.metric == "rsq") 
```

Fit Final Model
```{r}
final_fit_mohi500 <- final_workflow_mohi500 %>%
  last_fit(data_split_mohi500)

collect_metrics(final_fit_mohi500)      # test-set RMSE, R², etc.
mohi500_predictions <- collect_predictions(final_fit_mohi500)  # observed vs. predicted on the held-out test set
ggplot(mohi500_predictions, aes(prop_mod_high_sev, .pred)) +
  geom_point(alpha = 0.25)+
  geom_abline(linetype = "dashed", color = "red") +
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Observed",
    y = "Predicted",
    color = NULL,
    title = "Observed vs. Predicted Proportion of Moderate to High Severity Fire, 500 m"
  )

final_fitted_mohi500 <- final_fit_mohi500$.workflow[[1]]
final_fitted_mohi500
```

Explainer
```{r}
mohi500_explainer <- explain_tidymodels(
  final_fitted_mohi500,
  data = mohi500_train %>% st_drop_geometry() %>% dplyr::select(all_of(preds), Incid_Name),
  y = mohi500_train$prop_mod_high_sev,
  verbose = FALSE
)
```

Variable Importance Plot 
```{r}
# Variable Importance Plots
vip_mohi500 <- variable_importance(mohi500_explainer,
                                variables = preds)
plot(vip_mohi500) 
```

PDP
```{r}
pdp_mohi500 <- model_profile(
  mohi500_explainer,
  variables = preds,
  N = NULL
)
plot(pdp_mohi500)

pdp_mohi500 <- model_profile(
  mohi500_explainer,
  variables = "prop_extent_treated",
  N = NULL
)
plot(pdp_mohi500)
```


# Figures

## Variable Importance

```{r}
library(tidyverse)
library(patchwork)

# --- PANEL A: High-Severity Fire VIP ---
vip_hi_list <- list(
  "100m"  = vip_hi100,
  "250m" = vip_hi250,
  "500m" = vip_hi500
)

df_vip_hi <- map_dfr(vip_hi_list, ~ as.data.frame(.x), .id = "Scale") %>%
  filter(!variable %in% c("_full_model_", "_baseline_")) %>%
  group_by(Scale, variable) %>%
  summarise(dropout_loss = mean(dropout_loss, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Scale = factor(Scale, levels = c("100m", "250m", "500m")),
    dropout_loss_pct = dropout_loss * 100
  )

# --- PANEL C: Moderate/High-Severity Fire VIP ---
vip_mohi_list <- list(
  "100m"  = vip_mohi100,
  "250m" = vip_mohi250,
  "500m" = vip_mohi500
)

df_vip_mohi <- map_dfr(vip_mohi_list, ~ as.data.frame(.x), .id = "Scale") %>%
  filter(!variable %in% c("_full_model_", "_baseline_")) %>%
  group_by(Scale, variable) %>%
  summarise(dropout_loss = mean(dropout_loss, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    Scale = factor(Scale, levels = c("100m", "250m", "500m")),
    dropout_loss_pct = dropout_loss * 100
  )
```

## Partial Dependence Plot

```{r}
# --- PANEL B: High-Severity PDP ---
pdp_hi_list <- list(
  "100m"  = pdp_hi100,
  "250m" = pdp_hi250,
  "500m" = pdp_hi500
)

df_pdp_hi <- map_dfr(pdp_hi_list, ~ .x$agr_profiles, .id = "Scale") %>%
  filter(`_vname_` == "prop_extent_treated") %>% # filter to treated area predictor
  mutate(Scale = factor(Scale, levels = c("100m", "250m", "500m")))

# --- PANEL D: Moderate/High-Severity PDP ---
pdp_mohi_list <- list(
  "100m"  = pdp_mohi100,
  "250m" = pdp_mohi250,
  "500m" = pdp_mohi500
)

df_pdp_mohi <- map_dfr(pdp_mohi_list, ~ .x$agr_profiles, .id = "Scale") %>%
  filter(`_vname_` == "prop_extent_treated") %>%
  mutate(Scale = factor(Scale, levels = c("100m", "250m", "500m")))
```

## Build ggplots

```{r}
# Custom color palette matching Figure 2
scale_colors <- c(
  "100m"  = "#0d0887",
  "250m" = "#cc4678",
  "500m" = "#f0f921"
)

# --- Plot A: VIP High Severity ---
p_a <- ggplot(df_vip_hi, aes(x = dropout_loss_pct, y = reorder(variable, dropout_loss_pct), color = Scale)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = scale_colors) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

# --- Plot B: PDP High Severity ---
p_b <- ggplot(df_pdp_hi, aes(x = `_x_`, y = `_yhat_`, color = Scale)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = scale_colors) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  labs(x = NULL, y = "p(high-severity fire)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

# --- Plot C: VIP Moderate/High Severity ---
p_c <- ggplot(df_vip_mohi, aes(x = dropout_loss_pct, y = reorder(variable, dropout_loss_pct), color = Scale)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = scale_colors) +
  labs(x = "Variable importance - increase in MSE (%)", y = NULL) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

# --- Plot D: PDP Moderate/High Severity ---
p_d <- ggplot(df_pdp_mohi, aes(x = `_x_`, y = `_yhat_`, color = Scale)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = scale_colors) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.01)) +
  labs(x = "Proportion of area previously treated", y = "p(moderate/high-severity fire)") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())
```

## Combine plots

```{r}
fig2 <- (p_a + p_b) / (p_c + p_d) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a") &
  theme(
    plot.tag = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

# Display final figure
print(fig2)
```

# Testing out interaction pdp

```{r}
library(pdp)

pred_fun <- function(object, newdata) {
  mean(predict(object, new_data = newdata)$.pred)
}

pd_dab <- pdp::partial(
  final_fitted_hi500,
  pred.var = c("prop_extent_treated", "DailyAreaBurned"),
  train    = hi500_train,    
  pred.fun = pred_fun,
  plot     = TRUE
)
pd_dab

pd_elev <- pdp::partial(
  final_fitted_hi500,
  pred.var = c("prop_extent_treated", "elevation"),
  train    = hi500_train,    
  pred.fun = pred_fun,
  plot     = TRUE
)
pd_elev
```

# Average reduction in percent area burned at high severity between 0 and 1 treated area

```{r}
# trt_effect <- function(explainer) {
#   dat <- explainer$data
# 
#   d0 <- dat; d0$prop_extent_treated <- 0   # untreated
#   d1 <- dat; d1$prop_extent_treated <- 1   # fully treated
# 
#   p0 <- predict(explainer$model, new_data = d0)$.pred
#   p1 <- predict(explainer$model, new_data = d1)$.pred
# 
#   tibble::tibble(
#     mean_pred_untreated = mean(p0),
#     mean_pred_treated   = mean(p1),
#     abs_diff            = mean(p1) - mean(p0),                 # in proportion units
#     pct_diff            = (mean(p1) - mean(p0)) / mean(p0) * 100  # relative % change
#   )
# }
# 
# results <- dplyr::bind_rows(
#   "High sev, 100m"   = trt_effect(hi100_explainer),
#   "High sev, 250m"   = trt_effect(hi250_explainer),
#   "High sev, 500m"   = trt_effect(hi500_explainer),
#   "Mod-high, 100m"   = trt_effect(mohi100_explainer),
#   "Mod-high, 250m"   = trt_effect(mohi250_explainer),
#   "Mod-high, 500m"   = trt_effect(mohi500_explainer),
#   .id = "model"
# )
# results
# mean(results$pct_diff)
```
