---
title: "Forest Tracker Data Exploration"
author: "Elizabeth Buhr"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: 3
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(root.dir = 'C:/Users/Elizabeth Buhr/Documents/fuels-analysis/')
```

```{r}
#load packages
library(sf) 
library(tidyverse)
library(ggplot2)
```

# 1. Data Structure
```{r}
#path to file (accessed 10/2/25)
treats_path <- "data/raw/CFTv2_FINAL_DATA.gdb"

#check layer name
st_layers(treats_path)

#save the layer
treats <- st_read(treats_path, layer = "CFTv2_FinalData_20260220")

#inspect the columns/structure of the layer
str(treats)
```

# 2. Exploratory Graphs

I made graphs to visualize all the columns that were required for an observation to be in the database. These include agency, management type, year completed, and acres managed. I also included forest type, which is recorded for nearly all observations.

## 2.1 AGENCY
```{r agency-bar, fig.width=10, fig.height=6}
#Agency Bar Chart
agency_df <- treats %>%
  st_drop_geometry() %>%
  group_by(AGENCY) %>%
  summarise(Count = n()) %>%
  mutate(
    Percent = Count / sum(Count),
    PercentRounded = round(100 * Percent, 1),
    Label = paste0(PercentRounded, "%")
  ) %>%
  filter(PercentRounded > 0) %>%    # Remove agencies with 0.0%
  arrange(Percent)                  # Smallest to largest

# Plot
ggplot(agency_df, aes(x = reorder(AGENCY, Percent), y = Percent, fill = AGENCY)) +
  geom_bar(stat = "identity", width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = Label), hjust = -0.1, size = 4) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Agencies Overseeing Activity Implementation",
    x = NULL,
    y = NULL
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  # removes percent labels
    axis.ticks.x = element_blank(),  # removes tick marks
    axis.text.y = element_text(size = 12)
  )
```

## 2.2 MGT_TYPE
```{r}
# MGT_TYPE barplot
ggplot(treats, aes(x = MGT_TYPE)) +
  geom_bar() +
  theme_bw() +
  labs(title = "Management Type", x = NULL, y = "Count") 

```

These four management types are further broken into categories in separate columns. The categories for the three management types having to do with fuels reduction are evaluated below.

### 2.2.3 RXFIRE_MGT
```{r}
# RXFIRE_MGT barplot
rxfire_df <- treats %>%
  st_drop_geometry() %>%
  filter(MGT_TYPE == "Prescribed Fire")

ggplot(rxfire_df, aes(x = ACTIVITY)) +
  geom_bar() +
  theme_bw() +
  labs(title = "Rx Fire Type", x = NULL, y = "Count")
```

### 2.2.4 Mechanical and Hand
```{r}
# CANOPY_MGT barplot
canopy_df <- treats %>%
  st_drop_geometry() %>%
  filter(MGT_TYPE == "Mechanical and Hand")

ggplot(canopy_df, aes(x = ACTIVITY)) +
  geom_bar() +
  theme_bw()+
  labs(title = "Canopy Management Type", x = NULL, y = "Count")
```

## 2.3 YEAR_COMP
```{r}
# YEAR_COMP histogram
hist(treats$YEAR_COMP, main = "Treatment Year Completed", xlab = "Year", xlim = c(2000,2025))
```

Years Range:
```{r}
range(treats$YEAR_COMP)
```

## 2.4 FOR_TYPE
```{r}
# FOR_TYPE barplot
treats <- treats %>%
  filter(FOR_TYPE != "None") %>%
  drop_na(FOR_TYPE)

forestplot <- ggplot(treats, aes(x = FOR_TYPE)) +
  geom_bar(fill = "darkblue") +
  theme_bw() +
  labs(title = "Treatments by Forest Type", x = NULL, y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(
  axis.text.x  = element_text(size = 45, angle = 45, hjust = 1),
  axis.text.y  = element_text(size = 45),
  axis.title.y = element_text(size = 45),
  plot.title   = element_text(size = 45, face = "bold")
)

ggsave(
  filename = "forest_type.png",
  plot = forestplot,
  width = 10.5,
  height = 7,      # adjust as needed
  units = "in",
  dpi = 300
)

SumForest <- treats %>%
  st_drop_geometry() %>%
  group_by(FOR_TYPE) %>%
  summarise(n = n(),
            percent = (n()/24666)*100)

```

## 2.5 ACRES_MGT

```{r}
#histogram of ACRES_MGT
hist(treats$ACRES_MGT, main = "Treatment Acres Managed", xlab = "Acres")
```

The distribution of acres is extremely skewed toward 0. Using a log scale here makes the distribution normal, if we end up wanting to do that.

Mean Acres:
```{r}
#but mean is small
mean(treats$ACRES_MGT)
```

Box plot:

```{r}
#we can see there are a couple outliers
boxplot(treats$ACRES_MGT, main = "Boxplot of Treatment Acres Managed", ylab = "Acres")
```

Mean without top 2 outliers:
```{r}
treats_sorted <- treats[order(-treats$ACRES_MGT), ]
treats_trimmed <- treats_sorted[-c(1, 2), ]
  #check out mean without the outliers and compare it to the mean with them
mean(treats_trimmed$ACRES_MGT, na.rm = TRUE)
```

The mean is nearly the same (difference of 1.3 acres) with or without the 2 highest outliers, so they don't influence that too much. The top outlier is in pinyon-juniper so we will end up filtering it out anyway. 

# 3. Questions

## 3.1 Who is burning?

```{r rxagency-bar, fig.width=10, fig.height=6}
#RX Burn Agency
AgRx_df <- rxfire_df %>%
  st_drop_geometry() %>%
  group_by(AGENCY) %>%
  summarise(Count = n()) %>%
  mutate(
    Percent = Count / sum(Count),
    PercentRounded = round(100 * Percent, 1),
    Label = paste0(PercentRounded, "%")
  ) %>%
  filter(PercentRounded > 0) %>%    # Remove agencies with 0.0%
  arrange(Percent)                  # Smallest to largest

# Plot
ggplot(AgRx_df, aes(x = reorder(AGENCY, Percent), y = Percent, fill = AGENCY)) +
  geom_bar(stat = "identity", width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = Label), hjust = -0.1, size = 4) +
  scale_y_continuous(labels = scales::percent_format(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Agencies within Rx Fire Management Type",
    x = NULL,
    y = NULL
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  # removes percent labels
    axis.ticks.x = element_blank(),  # removes tick marks
    axis.text.y = element_text(size = 12)
  )
```

## 3.2 What is the average acreage of each treatment type?

```{r}
treats %>%
  st_drop_geometry() %>%
  filter(MGT_TYPE %in% c("CANOPY", "RX FIRE", "SURFACE")) %>%
  group_by(MGT_TYPE) %>%
  summarise(
    avg_acres = mean(ACRES_MGT, na.rm = TRUE)
  ) %>%
  arrange(MGT_TYPE) %>%
  mutate(avg_acres = round(avg_acres, 2)) %>%  
  kable(col.names = c("Treatment Type", "Mean Acres"), align = "lc")

treats %>%
  filter(MGT_TYPE == "RX FIRE") %>%
  st_drop_geometry() %>%
  group_by(RXFIRE_MGT) %>%
  summarise(n = n())
```

Let's also check size of broadcast burning versus pile burning.

```{r}
rxfire_df %>%
  group_by(RXFIRE_MGT) %>%
  summarise(
    avg_acres = mean(ACRES_MGT, na.rm = TRUE)
  ) %>%
  arrange(RXFIRE_MGT) %>%
  mutate(avg_acres = round(avg_acres, 2)) %>%  
  kable(col.names = c("Rx Fire Type", "Mean Acres"), align = "lc")
```

# Appendix
```{r show-code, ref.label = all_labels(), echo = TRUE, eval = FALSE}
```
